### 2024-08-25 paper.tex
Add plot of RFCs published per year

\documentclass[twocolumn,10pt]{article}
\usepackage[l2tabu,orthodox]{nag}
\usepackage[utf8x]{inputenc}
\usepackage[british]{babel}
\usepackage{ifpdf}
\ifpdf
  \usepackage{microtype}
\fi
\usepackage{amsmath}
\usepackage[all]{onlyamsmath}
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{upquote}
\usepackage{graphicx}
\usepackage{url}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{color}
\usepackage[cm]{fullpage}

\frenchspacing
\uchyph=0

% Define a simple \todo{...} macro:
\newcommand{\todo}[1]{\textbf{\textcolor{red}{To do: #1}}}

%==================================================================================================
\begin{document}

\title{An Empirical Analysis of the Internet Engineering Task Force 
       with Computational Methods}
\author{Colin Perkins\\University of Glasgow}
\maketitle
%==================================================================================================
\begin{abstract}
  % Four sentences:
  %  - State the problem
  %  - Say why it's an interesting problem
  %  - Say what your solution achieves
  %  - Say what follows from your solution

  The abstract goes here.

\end{abstract}
%==================================================================================================
\section{Introduction}

% A good paper introduction is fairly formulaic. If you follow a simple set
% of rules, you can write a very good introduction. The following outline can
% be varied. For example, you can use two paragraphs instead of one, or you
% can place more emphasis on one aspect of the intro than another. But in all
% cases, all of the points below need to be covered in an introduction, and
% in most papers, you don't need to cover anything more in an introduction.
%
% Paragraph 1: Motivation. At a high level, what is the problem area you
% are working in and why is it important? It is important to set the larger
% context here. Why is the problem of interest and importance to the larger
% community?

% Paragraph 2: What is the specific problem considered in this paper? This
% paragraph narrows down the topic area of the paper. In the first
% paragraph you have established general context and importance. Here you
% establish specific context and background.

% Paragraph 3: "In this paper, we show that...". This is the key paragraph
% in the introduction - you summarize, in one paragraph, what are the main
% contributions of your paper, given the context established in paragraphs
% 1 and 2. What's the general approach taken? Why are the specific results
% significant? The story is not what you did, but rather:
%  - what you show, new ideas, new insights
%  - why interesting, important?
% State your contributions: these drive the entire paper.  Contributions
% should be refutable claims, not vague generic statements.

In this paper, we ...

% Paragraph 4: What are the differences between your work, and what others
% have done? Keep this at a high level, as you can refer to future sections
% where specific details and differences will be given, but it is important
% for the reader to know what is new about this work compared to other work
% in the area.

% Paragraph 5: "We structure the remainder of this paper as follows." Give
% the reader a road-map for the rest of the paper. Try to avoid redundant
% phrasing, "In Section 2, In section 3, ..., In Section 4, ... ", etc.

We structure the remainder of this paper as follows.

%==================================================================================================
% Concentrate single-mindedly on a narrative that:
%  - Describes the problem, and why it's interesting
%  - Describes your idea
%  - Defends your idea, showing how it solves the problem, and filling out
%    the details
% On the way, cite relevant work in passing, but defer discussion to the
% end.
%
% Introduce the problem, and your idea, using examples, and only then
% present the general case. Explain the idea as if your were speaking to
% someone using a whiteboard. Conveying the intuition is primary; details
% follow. Write in a top down manner: state broad themes and ideas first,
% then go into details.
%
% The introduction makes claims. The body of the paper provides evidence
% to support each claim. Check each claim in the introduction, identify
% the evidence, and forward-reference it from the claim.
%==================================================================================================
\section{}

\begin{figure}
  \centering
  \includegraphics{figures/rfcs-by-year.pdf}
  \caption{Number of RFCs published per year}
  \label{fig:rfcs-by-year}
\end{figure}

%==================================================================================================
\section{}

%==================================================================================================
\section{}

%==================================================================================================
\section{Related Work}

% This should come near the end, and focussing on discussing how your work
% relates to that of others. Any relevant related work should have been
% cited already, so this is not a list of related work, it's a discussion
% of how that work relates.
%
% Why not put related work after the introduction? 1) because describing
% alternative approaches gets between the reader and your idea; and 2)
% because the reader knows nothing about the problem yet, so your
% (carefully trimmed) description of various technical trade-offs is
% absolutely incomprehensible.
%
% When writing the related work:
%  - Give credit to others where it's due; this doesn't diminish the
%    credit you get from your paper.
%  - Acknowledge weaknesses in your approach.
%  - Ensure related work is accurate and up-to-date

%==================================================================================================
\section{Conclusions}

%==================================================================================================
\section*{Acknowledgements}

% Acknowledge funding sources.

%==================================================================================================
% Set the bibliography style. Choose one of the following, depending on the
% document class being used:
%
%   \bibliographystyle{abbrv}                 When using article class
%   \bibliographystyle{IEEEtran}              When using IEEE style
%   \bibliographystyle{ACM-Reference-Format}  When using ACM style

\bibliographystyle{ACM-Reference-Format}

% Load the bibliography file(s) for this paper:
\bibliography{example}

%==================================================================================================
% The following information gets written into the PDF file information:
\ifpdf
  \pdfinfo{
    /Title        (...)
    /Author       (...)
    /Subject      (...)
    /Keywords     (..., ..., ...)
    /CreationDate (D:20150827110616Z)
    /ModDate      (D:20150827110616Z)
    /Creator      (LaTeX)
    /Producer     (pdfTeX)
  }
  % Suppress unnecessary metadata, to ensure the PDF generated by pdflatex is
  % identical each time it is built. This needs pdfTeX 3.14159265-2.6-1.40.17
  % or later.
  \ifdefined\pdftrailerid
    \pdftrailerid{}
    \pdfsuppressptexinfo=15
  \fi
\fi
%==================================================================================================
\end{document}
% vim: set ts=2 sw=2 tw=75 et ai:

### 2024-08-25 scripts/plot-rfcs-by-year.py

@@ -1,0 +1,49 @@
+#!/usr/bin/env python3
+#
+# Copyright (c) 2024 University of Glasgow
+#
+# Redistribution and use in source and binary forms, with or without
+# modification, are permitted provided that the following conditions are met:
+#
+# 1. Redistributions of source code must retain the above copyright notice, this
+#    list of conditions and the following disclaimer.
+#
+#2. Redistributions in binary form must reproduce the above copyright notice,
+#   this list of conditions and the following disclaimer in the documentation
+#   and/or other materials provided with the distribution.
+#
+# THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS"
+# AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
+# IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE
+# DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE
+# FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL
+# DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR
+# SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER
+# CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY,
+# OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
+# OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+import sys
+import seaborn
+import pandas
+import matplotlib.pyplot as plt
+import matplotlib.ticker as ticker
+ 
+if len(sys.argv) != 3:
+    print(f"Usage: python3 {sys.argv[0]} <datafile.csv> <output_file.pdf>")
+    sys.exit(1)
+
+csv_path = sys.argv[1]
+pdf_path = sys.argv[2]
+
+plt.figure(figsize=(3.5, 2))
+seaborn.set(font_scale=0.6)
+
[…]

### 2024-08-25 scripts/plot-rfcs-by-year.py
Use matplotlib directly, rather than seaborn

@@ -24,8 +24,7 @@
 # OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
 
 import sys
-import seaborn
-import pandas
+import pandas as pd
 import matplotlib.pyplot as plt
 import matplotlib.ticker as ticker
  
@@ -36,14 +35,23 @@ if len(sys.argv) != 3:
 csv_path = sys.argv[1]
 pdf_path = sys.argv[2]
 
-plt.figure(figsize=(3.5, 2))
-seaborn.set(font_scale=0.6)
+csv = pd.read_csv(csv_path)
 
-csv = pandas.read_csv(csv_path)
-dat = csv.query("Year != 1969")
-res = seaborn.barplot(x="Year", y="Total", data=dat)
-res.xaxis.set_major_locator(ticker.MultipleLocator(5))
-fig = res.get_figure()
-fig.savefig(pdf_path)
+plt.figure(figsize=(3.5,2))
+plt.rc('font',**{'family':'serif','serif':['Times'], 'size': 9})
+plt.rc('text', usetex=True)
+plt.rc('axes', axisbelow=True)
+plt.rcParams['pdf.fonttype'] = 42
 
+plt.xlabel("Year")
+plt.ylabel("RFCs Published")
+
+plt.bar(csv["Year"], csv["Total"])
+
+# Annotate with date IETF created
+plt.plot([1986, 1986], [0, 400], ':', color='black')
+plt.text(1978, 410, 'IETF Created', color='black')
+
+plt.plot()
+plt.savefig(pdf_path, format="pdf", bbox_inches="tight")
 

### 2024-08-31 scripts/rfcs-by-year.py
Remove Areas

@@ -40,30 +40,21 @@ ri = RFCIndex(rfc_index = sys.argv[1])
 rfcnum = []
 year   = []
 stream = []
-area   = []
 
 for rfc in ri.rfcs():
     rfcnum.append(rfc.doc_id)
     year.append(rfc.year)
     stream.append(rfc.stream)
-    if rfc.area is None:
-        area.append("Other")
-    else:
-        area.append(rfc.area)
 
 df = pd.DataFrame({
         "rfc_num": rfcnum,
         "stream" : stream,
         "year": year,
-        "area": area
     })
 
 streams = df.pivot_table(values="rfc_num", aggfunc="count", index="year", columns="stream", fill_value=0)
-areas   = df.pivot_table(values="rfc_num", aggfunc="count", index="year", columns="area",   fill_value=0)
 totals  = df.groupby("year").agg("count")["rfc_num"]
 
-tmp = pd.merge(streams, areas, on="year")
-res = pd.merge(tmp, totals, on="year").rename(columns={"rfc_num":"Total"})
-
+res = pd.merge(streams, totals, on="year").rename(columns={"rfc_num":"Total"})
 res.to_csv(sys.argv[2], index_label="Year")
 

### 2024-09-12 paper.tex
Tidy; fix fonts
Rename scripts rfcs-by-year -> rfcs-by-year-stream
Add Abstract

@@
 \usepackage[all]{onlyamsmath}
+\usepackage{newtxtext}
+\usepackage{newtxmath}
 \usepackage[caption=false]{subfig}
@@
 
-  The abstract goes here.
+  The Internet requires interoperability between networks, systems, and
+  applications, as well as cooperation among a growing number of
+  stakeholders. The Internet Engineering Task Force (IETF) is critical in
+  supporting this cooperation and interoperability by bringing together
+  interested parties and standardising the protocols that power the
+  Internet, such as IPv4/6 or HTTP/s. Our work covers more 20 years of IETF
+  data, and analyses the participants of the IETF, the emails they
+  exchange, and the documents they produce. We show how these data can be
+  used to understand who are these Internet actors, and their interests as
+  well as how Internet's growth and maturity has given rise to a longer and
+  complex standardisation process where participants gain and exercise
+  influence and how this is reflected in the language they use and the
+  interactions they have.
 
@@
 %==================================================================================================
-% Concentrate single-mindedly on a narrative that:
-%  - Describes the problem, and why it's interesting
-%  - Describes your idea
-%  - Defends your idea, showing how it solves the problem, and filling out
-%    the details
-% On the way, cite relevant work in passing, but defer discussion to the
-% end.
-%
-% Introduce the problem, and your idea, using examples, and only then
-% present the general case. Explain the idea as if your were speaking to
-% someone using a whiteboard. Conveying the intuition is primary; details
-% follow. Write in a top down manner: state broad themes and ideas first,
-% then go into details.
-%
-% The introduction makes claims. The body of the paper provides evidence
-% to support each claim. Check each claim in the introduction, identify
-% the evidence, and forward-reference it from the claim.
-%==================================================================================================
 \section{}
@@
   \centering
-  \includegraphics{figures/rfcs-by-year.pdf}
+  \includegraphics{figures/rfcs-by-year-stream.pdf}
   \caption{Number of RFCs published per year}

### 2024-09-12 paper.tex
Tidy preamble
Initial import of text on documents from IMC 2021 and TMA 2023 papers

@@
 \usepackage[british]{babel}
-\usepackage{ifpdf}
-\ifpdf
 \usepackage{microtype}
-\fi
 \usepackage{amsmath}
@@
 
-% Define a simple \todo{...} macro:
 \newcommand{\todo}[1]{\textbf{\textcolor{red}{To do: #1}}}
+\newcommand{\pb}[1]{\vspace{0.75ex}\noindent{\textbf{#1}}}
 
@@
 %==================================================================================================
-\section{}
+\section{Background and Datasets}
+
+% The following is from our IMC 2021 paper:
+
+We start by presenting an overview of the IETF, and the publication process
+for RFCs, before outlining the data sources we use within this paper. We
+also highlight the ethical considerations of accessing and processing this
+data.
+
+%--------------------------------------------------------------------------------------------------
+\subsection{An IETF Primer}
+
+% The following is from our IMC 2021 paper:
+
+\pb{The IETF} is an open standards organisation, which develops
+Internet standards via contributions and collaborations across a number of
+voluntary stakeholders, including academics, consultants, industry
+representatives, governments, and civil society organisations.  Through
+extensive collaboration across contributors, the IETF, and associated
+organisations, develops Internet standards and other documents. These are
+published by the RFC Editor (\url{https://www.rfc-editor.org}) in four
+\emph{publication streams}: the IETF stream, the Internet Research Task
+Force (IRTF) stream, the Internet Architecture Board (IAB) stream, and the
+Independent Submission stream.
+
+There is also a fifth, legacy stream, comprising RFCs published prior to
+the adoption of separate publication streams in July 2007 \cite{rfc4844}.
+While the IETF is an open standards forum that develops technical standards
+and operational guidelines for the Internet, the IRTF is an associated
+organisation that promotes longer-term research, and the IAB provides
+long-range technical direction for Internet development. The Independent
+Submission stream ``allows RFC publication for some documents that are
+outside the official IETF/IAB/IRTF process but are relevant to the Internet
+community'' (\url{https://www.rfc-editor.org/about/independent}).
+
+\pb{The standards development process} is an inherently collaborative
+activity.  Most day-to-day work is conducted on public mailing lists, in
+conjunction with three plenary meetings and numerous interim working group
+meetings per year. 
+The mailing lists are broadly split into three categories: announcement
+lists, where replies are not allowed; non-working group lists, for
+discussing topics that do not relate to the work within an IETF working
+group or IRTF research group; and working group and area lists, where
+technical discussions take place.
+
+The process of RFC publication begins with the submission of an
+\emph{Internet-Draft}. Whereas anyone can post a draft, not all drafts
+become RFCs. After a draft is first posted, multiple revisions might take
+place resulting in multiple versions of the draft. Each new draft is
+announced on one or more mailing lists related to the topic of the draft,
+soliciting feedback and encouraging discussion. Drafts are initially posted
+by individuals. For publication under the IETF stream, drafts must then be
+adopted by a working group, where, via further revision, the draft may
+ultimately be published as an RFC. The process of managing drafts, and the
+degree and type of peer review conducted prior to their submission to the
+RFC Editor for publication, differs between streams.
+
+Once the technical development of the draft is complete its publication is
+managed by the RFC Editor, who maintains the master archive of the RFC
+documents, along with an index of metadata pertaining to the RFCs and their
+authors. Finally, once an RFC has been published, deployment is voluntary,
+and therefore not all RFCs are widely implemented.
+
+%--------------------------------------------------------------------------------------------------
+\subsection{Data Sources}
+
+%--------------------------------------------------------------------------------------------------
+\subsection{Ethical Considerations}
+\label{sec:ethics}
+
+% The following is from our IMC 2021 paper:
+
+The data we analyse is extracted from public IETF archives and APIs.  We
+have taken steps to ensure ethical compliance.  To ensure that our access
+to these services does not cause operational problems for the IETF, we are
+in regular contact with the IETF Tools Team and Secretariat, as well as the
+operators of the Datatracker and mailing list archive.  We have extensively
+discussed our work with IETF leadership (IETF and IAB Chairs, the IETF
+Executive Director, and the IRTF Chair, who is a co-author on this paper)
+to ensure that our access falls within their acceptable use policies.
+
+Participation in the IETF is dependent on agreement to abide by the
+policies and procedures described at \url{https://www.ietf.org/about/note-well},
+including the privacy policy at \url{https://www.ietf.org/privacy-statement}.
+These make explicit provision that mailing list archives and the metadata
+contained in the Datatracker system will be made public, and it is this
+public data that we process to extract the aggregate statistics presented
+in Sections \ref{sec:documents}, \ref{sec:authors}, and \ref{sec:mail}, as
+well as the features analysed in Section \ref{sec:model}. We transfer and
+store data securely, and retain it only for the time needed to perform the
+analysis.  Since we operate entirely using the public APIs provided by
+IETF, we have no access to private data about individuals.
+
+In balancing these ethical considerations with the reproducibility of our
+work, we provide the tools needed to access the datasets from the relevant
+IETF sources, rather than the data itself.
+
+%==================================================================================================
+\section{Trends in RFC Publication}
+
+% \begin{figure}
+%   \centering
+%   \includegraphics{figures/rfcs-by-year-stream.pdf}
+%   \caption{Number of RFCs published per year}
+%   \label{fig:rfcs-by-year}
+% \end{figure}
+
+%--------------------------------------------------------------------------------------------------
+\subsection{Document Production and Complexity}
+
+% The following is from our IMC 2021 paper, Section 3.1:
+
+\pb{Growth of RFCs.}
+In total, 8,711 RFCs have been published through to the end of 2020.
+Figure~\ref{fig:rfcs_by_area} shows how publication trends, in terms of
+IETF areas and non-IETF streams, have changed over time. We identify three
+broad publication phases in the RFC series. First, in 1969 through 1974,
+RFCs are published at a rapid rate during the initial development of the
+ARPANET. Then, from 1975 through 1985, development slows. This reflects a
+relatively small community that is gaining real-world experience with the
+network and developing a small core of applications and protocols. Finally,
+with the creation of the IETF and the introduction of the National Science
+Foundation Network (NSFNET) in 1986, both the community, and the number of
+RFCs published, starts to expand rapidly. This is further driven by the
+opening of the network to commercial and public use in the mid-1990s, and
+continues to this day.
+
+As shown, the annual RFC publication rate was highest in 2005, at the peak of
+the standardisation efforts for SIP and related standards for voice-over-IP and
+Internet telephony. The rate of publication has slowed in recent years,
+following the completion of large work programmes relating to HTTP/2 and WebRTC.
 
 \begin{figure}
-  \centering
-  \includegraphics{figures/rfcs-by-year-stream.pdf}
-  \caption{Number of RFCs published per year}
-  \label{fig:rfcs-by-year}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/rfcs_areas.pdf}
+\caption{RFCs by area. \textit{Other} includes legacy RFCs, and RFCs from other non-IETF streams.}
+\label{fig:rfcs_by_area}
+\end{figure}
+
+\pb{Role of Working Groups (WGs).}
+From the creation of the IETF in 1986, the growing community with its interests
+in an increasing set of applications and protocols, has been split into working
+groups. These working groups are chartered with a focus on a well-defined
+programme of work, and exist within areas that have a broader focus.
+As shown in
+Figure~\ref{fig:rfcs_by_area}, the output of different areas has remained
+relatively stable over time. The most notable trends begin with the creation of
+the Real-time Applications and Infrastructure (rai) area from within the
+Transport (tsv) area, and its later merger with the Applications (app) area to
+become the Applications and Real-Time (art) area around 2014. Additionally, we also observe the significant growth in output of the Routing (rtg) area,
+owing to the ongoing development of standards for MPLS, service function
+chaining, and fat tree routing in data centres.
+
+To give a sense for the broader productivity of the IETF,
+Figure~\ref{fig:pub_wgs_yearly} shows the number of working groups that
+publish RFCs each year. This highlights how the structure of the IETF has
+grown to accommodate its larger community: in the early 1990s, fewer than
+20 working groups were actively publishing RFCs, while in recent years
+there has typically been at least 60 different publishing groups, with a
+peak of 97 active working groups, and other activities, in 2011.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/unique_wgs_per_year_areas.pdf}
+\caption{Number of publishing working groups. \textit{Other} includes legacy RFCs, IRTF research groups, and non-IETF streams.}
+\label{fig:pub_wgs_yearly}
+\end{figure}
+
+Figure~\ref{fig:rfcs_days_to_pub} plots the median number of days from the
+submission of an RFC's first draft, through to its publication as an RFC.
+This shows a clear trend: RFCs are taking longer to make their way through
+the standardisation and publication process. The median number of days to
+publication was 469 in 2001, rising to 1,170 in 2020. Further,
+Figure~\ref{fig:drafts_year} shows the median number of Internet-Drafts
+that are posted before an RFC is published. Days to publication and number
+of drafts are strongly correlated, suggesting that the time is spent making
+changes to the document. This may go some way towards explaining the
+decline in output of the IETF: each RFC is taking longer to produce, with
+more revisions before publication.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/day_counts_yearly.pdf}
+\caption{Days from first draft to RFC publication}
+\label{fig:rfcs_days_to_pub}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/draft_counts_yearly.pdf}
+\caption{Number of drafts per RFC}
+\label{fig:drafts_year}
+\end{figure}
+
+\pb{RFC Length.}
+One may conjecture that this slowdown is driven by longer RFCs that contain
+more material. To explore this, Figure~\ref{fig:pages_year} shows the
+median page count of RFCs. This shows that the increase in the duration of
+the standardisation process for RFCs cannot be attributed to RFCs becoming
+longer: median page counts have remained stable.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/page_counts_yearly_dt.pdf}
+\caption{RFC page counts}
+\label{fig:pages_year}
+\end{figure}
+
+\pb{Relationships between RFCs.}
+RFCs themselves may be becoming more complex as a result of them needing to
+describe their relationship with older standards. As the Internet has
+evolved and matured, applications and protocols must be maintained, and new
+standards need to interoperate with existing RFCs.
+Figure~\ref{fig:updates_year} highlights this, showing the proportion of
+RFCs that are published each year that update (i.e., extend or augment) or
+obsolete (i.e., replace) one or more previously published RFCs. Intuitively,
+this percentage has slowly increased as the IETF has matured: in 2020, more
+than 30\% of RFCs updated or made obsolete a previous RFC.
+Figure~\ref{fig:citations_year} expands on this, showing the median number
+of citations from each RFC to other Internet-Drafts and RFCs. This
+similarly shows that RFCs are increasingly referring to prior work.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/update_obsolete_yearly_pct.pdf}
+\caption{RFCs that update or obsolete previous RFCs}
+\label{fig:updates_year}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/cite_counts_yearly.pdf}
+\caption{Citations from RFCs to other Internet-Drafts and RFCs per RFC}
+\label{fig:citations_year}
+\end{figure}
+
+\pb{Use of requirements-setting language.}
+Figure~\ref{fig:keyword_usage_rates} further confirms the growing
+complexity of RFCs, showing how the use of keywords has evolved over time.
+Keywords are used in RFCs to indicate the normative requirements an RFC
+imposes on implementations. Figure~\ref{fig:keyword_usage_rates} shows the
+total number of occurrences of each of the RFC 2119~\cite{RFC2119} keywords
+(i.e., MUST, MUST NOT, REQUIRED, SHALL, SHALL NOT, SHOULD, SHOULD NOT,
+RECOMMENDED, MAY, OPTIONAL), divided by the page count of the RFC. As
+shown, the median number of keywords per page grew from 2001 through to
+2010, indicating a growing number of requirements being expressed in RFCs,
+before plateauing in recent years.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/keyword_usage_rate.pdf}
+\caption{Keyword occurrences per page}
+\label{fig:keyword_usage_rates}
+\end{figure}
+
+\todo{Should we include ``Academic impact of RFCs'' from IMC 2021?}
+
+\pb{Summary.}
+RFCs are taking longer to produce, and they go through a greater number of
+revisions before publication. Further, they increasingly update or
+reference previously published RFCs, and make greater use of
+requirements-setting keywords. Yet, while these measures indicate that the
+overall complexity of standards documents is increasing, none of these
+factors strongly correlate with the time taken to publish RFCs. This
+indicates that these factors aren't driving the increasing duration of the
+standardisation process. In the sections that follow, we explore trends in
+authorship and community interaction that may also impact the process.
+
+%--------------------------------------------------------------------------------------------------
+\subsection{Document Errata}
+
+% The following is from our TMA 2023 paper, Section III:
+
+\pb{Errata over Time.}
+Figure~\ref{fig:errata_per_year} presents the number of errata filed, on
+average per RFC, since 1969 based on the year of RFC publication.  The peak
+in the number of errata per RFC occurs in 1981. Only 29 RFCs were published
+that year, but they include major documents such as RFCs 791, 792, and 793
+(the original versions of the IP~\cite{rfc791}, ICMP~\cite{rfc792}, and
+TCP~\cite{rfc793} standards), with 17, 7, and 47 errata, respectively.
+These important protocols clearly garnered a great deal of scrutiny and
+revision.  The second highest peak occurs in 2006.  In contrast to the
+previous examples, this has the highest number of RFCs published per year
+(459), including RFC 4601 \cite{RFC4601} that has the most errata (114).
+Since this second peak, there has been a steady decrease in the number of
+errata filed.  This broadly correlates with the number of RFCs published
+per year, with Pearson coefficient 0.59 since 2007.  Table
+\ref{tab:top-10-rfcs-by-errata} lists the top RFCs by errata filing count.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/tma-2023/errata-by-year.pdf}
+\caption{Average errata filed per RFC for each year by RFC publication year,
+grouped by IETF area (acronyms expanded in Table~\ref{tab:errata-stats-area}).}
+\label{fig:errata_per_year}
+\end{figure}
+
+\input{tables-prev/tma-2023/top-rfcs-errata.tex}
+\input{tables-prev/tma-2023/errata-stats-area.tex}
+
+\pb{Errata Delay.}
+We next explore how long it takes for errata to be identified and filed.
+Figure~\ref{fig:errata_submission_days} presents a CDF of the number of
+days between RFC publication and the errata being filed, broken down based
+on IETF area. We see a wide range of delays. 7.3\% of errata are filed
+within the first 30 days, suggesting that many RFCs are published with
+issues that could have been identified prior to publication.  RFCs from the
+General (\emph{gen}) area--describing IETF policies and procedures--have
+the longest delay, with a median of 3,458 days, compared to the
+Applications and Real-time (\emph{art}) area with a median of 681
+days.\footnote{Errata are filed against RFCs within the \emph{subip} area
+within a median of 48 days, but this is skewed, with only 19 RFCs being
+published in that area.} Editorial errata are typically filed more quickly,
+with a median of 987 days, compared to a median of 1,138 days for technical
+errata.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/tma-2023/errata-submission-dates-area.pdf}
+\caption{Days from RFC publication and errata filing by IETF area (acronyms
+  expanded in Table~\ref{tab:errata-stats-area}).}
+\label{fig:errata_submission_days}
+\end{figure}
+
+\pb{Errata Status.}
+Figure~\ref{fig:errata_status} categorises the errata by status and
+publication year of the RFC to which they relate.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/tma-2023/errata-by-status.pdf}
+\caption{Errata filings by status, by publication year of the RFC.}
+\label{fig:errata_status}
+\end{figure}
+
+The largest share (42.5\%) of errata are \emph{verified}: errata that have
+been has been confirmed as necessary and accurate. This suggests that many
+errata are useful to the community. The next largest share (30.3\%) are
+those labelled \emph{hold for document update}.  These are errata that are
+not a necessary update to the RFC, but may be considered on future
+revisions. For example, \emph{erratum 6278} describes an oversight in RFC
+8610~\cite{rfc8610}; the solution to this is non-trivial, and so will be
+considered in the next version of the specification. Of the 930 RFCs that
+have \emph{hold for document update} errata filed against them, only 40\%
+have been updated or obsoleted by a subsequent RFC.  We flag that this may
+be a cause for concern, or at least a missed opportunity for improvements
+to standards.
+
+The third largest category (13\%) is \emph{rejected}, which covers errata
+that are invalid (like \emph{erratum 6323}, which was rejected because the
+original text was understood to be correct) or proposes a significant
+change to the RFC that should be done by publishing a new RFC (like
+\emph{erratum 5814}, which was rejected for proposing a significant change,
+rather than reporting an error). Such a large fraction of rejected
+submissions is unexpected and may flag issues with people's understanding
+of the errata process and its place within the wider standardisation
+process.  Finally, 14.2\% of errata are \emph{reported} but unverified.
+Again, we are surprised to see unverified errata from over a decade ago,
+suggesting the process should be expedited.
+
+\pb{Errata per RFC Area, Status, and Stream.}
+Figure~\ref{fig:errata_per_rfc} shows a CDF of the number of errata filed,
+per RFC, in each IETF area. Non-IETF RFCs, e.g., IRTF and independent stream
+RFCs, and legacy IETF RFCs, are labelled as ``None''.  We confirm errata in
+standards are common: of the 4,373 standards-track RFCs in our dataset,
+32.7\% have attracted at least one erratum filing.  However, there are
+three notable outliers.  First, RFCs published by the Sub-IP (subip) Area
+have very few errata, with only 5\% of subip RFCs attracting errata
+filings. This is because this temporary area -- established in 2001 and
+concluded in 2005 -- only published 19 RFCs, resulting in a far smaller
+sample than the other areas. For comparison, the next smallest area,
+General (gen), published 39 RFCs. \emph{gen} RFCs attract a greater number
+of errata on average, vs. \emph{subip} RFCs, likely due to their broader
+relevance.  Second, we see that both the Application (app) and Security
+(sec) Areas' RFCs are more likely to have errata filed for them, with
+35.9\% of Application and 39\% of Security RFCs attracting at least one
+erratum filing.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/tma-2023/errata-by-rfc-by-area.pdf}
+\caption{CDF of errata filed per RFC by grouped by IETF area (acronyms
+  expanded in Table~\ref{tab:errata-stats-area}).}
+\label{fig:errata_per_rfc}
+\end{figure}
+
+Finally, Table~\ref{tab:errata-stats-area} details the split between
+\emph{technical} and \emph{editorial} errata across each area. While there
+is broadly an even split, there are areas where one type of errata is more
+dominant. For example, in the Routing (rtg) area, 60.8\% of filings are
+editorial, while in the Applications (app) area, 59\% were technical.  It
+remains to determine why this is the case, and, in particular, to establish
+whether there is something inherent about the RFCs published by these areas
+that makes them more prone to containing errata, and to containing one type
+of errata vs. another. For example, in the Routing area, structured
+notation is frequently used to define routing entities; editorial errata
+are often filed in those definitions. Targeting such areas with improved
+alternate review procedures may be beneficial.
+
+Tables~\ref{tab:errata-stats-stream} and \ref{tab:errata-stats-status}
+further categorise errata by the stream and status, at the time of
+publication, of each RFC. As expected, the majority of errata are filed
+against IETF RFCs and \emph{Proposed Standards} since these make up the
+majority of RFCs that are published. However, there are notable differences
+in the average number of filings per RFC. \emph{Proposed Standards} (1.01
+errata per RFC), \emph{Draft Standards} (1.93), and \emph{Internet
+Standards} (2.17) attract a far higher number of errata per RFC than
+\emph{Informational} (0.52) or \emph{Experimental} (0.39) documents. This
+may be due to the additional readership and attention that standards-track
+documents receive, and because they are more likely to be the basis for
+future work and protocol extensions.
+
+\input{tables-prev/tma-2023/errata-stats-stream.tex}
+\input{tables-prev/tma-2023/errata-stats-status.tex}
+
+\todo{Should we include ``Impact of Citations'' from TMA 2023?}
+
+\pb{Errata Location.}
+Finally, we investigate the location of errata within RFCs.
+Figure~\ref{fig:errata_location_percent_count} presents the number of
+errata occurring at each decile of the documents, for the 2,552 filings
+where accurate location information is available, and after the copyright
+notice and other boilerplate has been removed.  We see that technical
+errata dominate over editorial in almost all places, except for the very
+beginning where the \textit{Introduction} is located. Moreover, it shows
+that the most technical errata are near the middle of the document where
+the most complex content is.  We explore where errata occur, with
+Figure~\ref{fig:errata_section_wise_counts} showing section titles for
+errata appearing in at least $10$ documents. Sections such as the
+\emph{Introduction} or \textit{References} are dominated by editorial
+errata while more technical sections, like \emph{IANA Considerations} (i.e.,
+parameter registrations), \emph{Security Considerations} or
+\emph{Definitions}, have a larger proportion of technical errata. In
+addition, we see that sections labelled \emph{Appendix} attract a
+significant proportion of technical errata. While appendices vary in their
+content, they are widely used to provide pseudocode and test vectors, or to
+describe algorithms. This suggests that it may be useful to target review
+efforts on appendices and other dense technical content where errata are
+more likely.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/tma-2023/location-percent.pdf}
+\caption{Errata counts by percentile location in document (0 is the beginning; 100 is the end).}
+\label{fig:errata_location_percent_count}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/tma-2023/section-title-counts.pdf}
+\caption{Errata counts by section title for the more frequent section titles.}
+\label{fig:errata_section_wise_counts}
 \end{figure}
@@
 %==================================================================================================
-\section{}
-
-%==================================================================================================
-\section{}
+\section{Exploring Authorship}
 
@@
 %==================================================================================================
-% Set the bibliography style. Choose one of the following, depending on the
-% document class being used:
-%
-%   \bibliographystyle{abbrv}                 When using article class
-%   \bibliographystyle{IEEEtran}              When using IEEE style
-%   \bibliographystyle{ACM-Reference-Format}  When using ACM style
-
-\bibliographystyle{ACM-Reference-Format}
-
-% Load the bibliography file(s) for this paper:
-\bibliography{example}
+\bibliographystyle{abbrv}
+\bibliography{paper}
 

### 2024-09-12 tables-prev/tma-2023/errata-stats-stream.tex
Initial import of text on documents from IMC 2021 and TMA 2023 papers

\begin{table*}
\centering
\footnotesize
\begin{tabular}{lr|rrrr|rr}
\toprule
\textbf{Stream} & \textbf{\#} & \textbf{Verified} & \textbf{Held} & \textbf{Rejected} & \textbf{Reported} & \textbf{Technical} & \textbf{Editorial} \\
\midrule
IETF (6619) & 5797  & 2348  & 1815  & 798  & 836  & 3034  & 2763 \\
IAB (124) & 55  & 25  & 13  & 3  & 14  & 23  & 32 \\
Independent (376) & 330  & 235  & 32  & 28  & 35  & 172  & 158 \\
Legacy (1929) & 510  & 226  & 182  & 39  & 63  & 198  & 312 \\
IRTF (97) & 67  & 37  & 8  & 9  & 13  & 44  & 23 \\
\midrule All & 6759  & 2871  & 2050  & 877  & 961  & 3471  & 3288 \\
\bottomrule
\end{tabular}
\caption{Errata statistics by stream; the ``\emph{Editorial}'' stream has no documents, and is not shown.}
\label{tab:errata-stats-stream}
\end{table*}

### 2024-09-12 tables-prev/tma-2023/top-rfcs-errata.tex
Initial import of text on documents from IMC 2021 and TMA 2023 papers

\begin{table*}
\centering
\footnotesize
\begin{tabular}{rlrrr}
\toprule
\textbf{RFC} & \textbf{Title} & \textbf{Year} & \textbf{Area} & \textbf{Filing count} \\
\midrule
4601 & Protocol Independent Multicast - Sparse Mode (PIM-SM): Protocol Specification (Revised) & 2006 & rtg & 114 \\
4880 & OpenPGP Message Format & 2007 & sec & 52 \\
793 & Transmission Control Protocol & 1981 & None & 47 \\
4634 & US Secure Hash Algorithms (SHA and HMAC-SHA) & 2006 & None & 44 \\
5661 & Network File System (NFS) Version 4 Minor Version 1 Protocol & 2010 & tsv & 42 \\
1345 & Character Mnemonics and Character Sets & 1992 & app & 41 \\
8446 & The Transport Layer Security (TLS) Protocol Version 1.3 & 2018 & sec & 40 \\
5545 & Internet Calendaring and Scheduling Core Object Specification (iCalendar) & 2009 & app & 35 \\
3261 & SIP: Session Initiation Protocol & 2002 & rai & 33 \\
5905 & Network Time Protocol Version 4: Protocol and Algorithms Specification & 2010 & int & 32 \\
\bottomrule
\end{tabular}
\caption{Top 10 RFCs by errata filing count}
\label{tab:top-10-rfcs-by-errata}
\end{table*}

### 2024-09-21 paper.tex
Updates to Trends in RFC Publication

@@
 %--------------------------------------------------------------------------------------------------
-\subsection{Document Production and Complexity}
+\subsection{RFC Production and Complexity}
+
+RFC production trends show clear evidence that the IETF is a mature
+standards development organisation. The peak of activity was in the 1990s
+and early 2000s, with the initial broad deployment of the Internet, and the
+organisation has since entered more of a development and maintenance mode
+of operation. There are clear signs that the complexity of the installed
+base of protocols is slowing development of new technologies.
 
@@
 
-\pb{Growth of RFCs.}
+\pb{RFC Publication Rate:}
 In total, 8,711 RFCs have been published through to the end of 2020.
@@
 
-As shown, the annual RFC publication rate was highest in 2005, at the peak of
-the standardisation efforts for SIP and related standards for voice-over-IP and
-Internet telephony. The rate of publication has slowed in recent years,
-following the completion of large work programmes relating to HTTP/2 and WebRTC.
+As shown, the annual RFC publication rate was highest in 2005, at the peak
+of the standardisation efforts for SIP and related standards for voice-over-IP
+and Internet telephony. The rate of publication has slowed in recent years,
+following the completion of large work programmes relating to HTTP/2 and
+WebRTC.
 
@@
 
-\pb{Role of Working Groups (WGs).}
-From the creation of the IETF in 1986, the growing community with its interests
-in an increasing set of applications and protocols, has been split into working
-groups. These working groups are chartered with a focus on a well-defined
-programme of work, and exist within areas that have a broader focus.
-As shown in
-Figure~\ref{fig:rfcs_by_area}, the output of different areas has remained
-relatively stable over time. The most notable trends begin with the creation of
-the Real-time Applications and Infrastructure (rai) area from within the
-Transport (tsv) area, and its later merger with the Applications (app) area to
-become the Applications and Real-Time (art) area around 2014. Additionally, we also observe the significant growth in output of the Routing (rtg) area,
-owing to the ongoing development of standards for MPLS, service function
-chaining, and fat tree routing in data centres.
+\pb{Role of Working Groups:}
+From the creation of the IETF in 1986, the growing community with its
+interests in an increasing set of applications and protocols, has been
+split into working groups. These working groups are chartered with a focus
+on a well-defined programme of work, and exist within areas that have a
+broader focus.  As shown in Figure~\ref{fig:rfcs_by_area}, the output of
+different areas has remained relatively stable over time. The most notable
+trends begin with the creation of the Real-time Applications and
+Infrastructure (rai) area from within the Transport (tsv) area, and its
+later merger with the Applications (app) area to become the Applications
+and Real-Time (art) area around 2014. Additionally, we also observe the
+significant growth in output of the Routing (rtg) area, owing to the
+ongoing development of standards for MPLS, service function chaining, and
+fat tree routing in data centres.
 
@@
 %--------------------------------------------------------------------------------------------------
-\subsection{Document Errata}
+\subsection{RFC Errata}
 

### 2024-09-21 paper.tex
Consistently use : in \pb{}

@@
 
-\pb{RFC Length.}
+\pb{RFC Length:}
 One may conjecture that this slowdown is driven by longer RFCs that contain
@@
 
-\pb{Relationships between RFCs.}
+\pb{Relationships between RFCs:}
 RFCs themselves may be becoming more complex as a result of them needing to
@@
 
-\pb{Use of requirements-setting language.}
+\pb{Use of requirements-setting language:}
 Figure~\ref{fig:keyword_usage_rates} further confirms the growing
@@
 
-\pb{Summary.}
+\pb{Summary:}
 RFCs are taking longer to produce, and they go through a greater number of
@@
 
-\pb{Errata over Time.}
+\pb{Errata over Time:}
 Figure~\ref{fig:errata_per_year} presents the number of errata filed, on
@@
 
-\pb{Errata Delay.}
+\pb{Errata Delay:}
 We next explore how long it takes for errata to be identified and filed.
@@
 
-\pb{Errata Status.}
+\pb{Errata Status:}
 Figure~\ref{fig:errata_status} categorises the errata by status and
@@
 
-\pb{Errata per RFC Area, Status, and Stream.}
+\pb{Errata per RFC Area, Status, and Stream:}
 Figure~\ref{fig:errata_per_rfc} shows a CDF of the number of errata filed,
@@
 
-\pb{Errata Location.}
+\pb{Errata Location:}
 Finally, we investigate the location of errata within RFCs.

### 2024-09-21 paper.tex
Add separators
Add content from IMC 2021 paper, §3.1, to Trends in Participant Demographics

@@
 
+%..................................................................................................
 \pb{The IETF} is an open standards organisation, which develops
@@
 
+%..................................................................................................
 \pb{The standards development process} is an inherently collaborative
@@
 
+%..................................................................................................
 \pb{RFC Publication Rate:}
@@
 
+%..................................................................................................
 \pb{Role of Working Groups:}
@@
 
+%..................................................................................................
 \pb{RFC Length:}
@@
 
+%..................................................................................................
 \pb{Relationships between RFCs:}
@@
 
+%..................................................................................................
 \pb{Use of requirements-setting language:}
@@
 
+%..................................................................................................
 \pb{Summary:}
@@
 
+%..................................................................................................
 \pb{Errata over Time:}
@@
 
+%..................................................................................................
 \pb{Errata Delay:}
@@
 
+%..................................................................................................
 \pb{Errata Status:}
@@
 
+%..................................................................................................
 \pb{Errata per RFC Area, Status, and Stream:}
@@
 
+%..................................................................................................
 \pb{Errata Location:}
@@
 %==================================================================================================
-\section{Exploring Authorship}
+\section{Trends in Participant Demographics}
+
+% The following is from our IMC 2021 paper, Section 3.2:
+
+The IETF Datatracker maintains metadata about document authors, including
+their names, email addresses, affiliations, and location information. This
+dataset does not cover the entire RFC corpus, with metadata available for
+authors of RFCs published from 2001, and where it has been provided.
+Country data is available for around 70\% of authors, while affiliation
+information is given for around 80\%. In this section, we look at the
+authorship of all of the RFCs published each year, using the available
+metadata. An author is counted once in a year for each affiliation or
+location they hold.
+
+%..................................................................................................
+\pb{Geographic Distribution of Authors:}
+Figures~\ref{fig:author_countries_normalised} and
+\ref{fig:author_continents_normalised} show the proportion of authors from
+countries and continents, respectively. The IETF has signalled that it
+wishes to encourage greater geographical diversity~\cite{rfc7704,
+ietfblog:diversity}. Without an explicit goal, we frame our findings within
+the context of global population distribution.  We find that North America,
+while still disproportionately over represented, is becoming less dominant.
+75\% of authors were from North America in 2001, and this has declined to
+44\% in 2020. At the same time, representation of both Europe and Asia has
+grown, from 17\% to 40\% and 6\% to 14\%, respectively.  However, Africa
+and South America remain heavily underrepresented, with only $\approx$0.5\%
+of authors coming from either continent in 2020.  This suggests that, if
+the IETF is to become more geographically representative, further efforts
+are needed.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/authors/top5_countries_normalised.pdf}
+\caption{Authorship countries (normalised)}
+\label{fig:author_countries_normalised}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/authors/continents_normalised.pdf}
+\caption{Authorship continents (normalised)}
+\label{fig:author_continents_normalised}
+\end{figure}
+
+%..................................................................................................
+\pb{Author Affiliations:}
+Figure~\ref{fig:author_affiliations_normalised} shows the top 10
+affiliations by proportion of authors each year. Affiliation data is
+gathered from the Datatracker, and processed to normalise affiliation
+names, removing common variations in spelling, and amalgamating known
+subsidiaries and merged companies. For example, Huawei and Futurewei are
+combined as Huawei, and Sun Microsystems is merged with Oracle.
+
+From Figure~\ref{fig:author_affiliations_normalised}, we observe several
+interesting trends.  First, Cisco remains a consistent employer of IETF
+contributors, with around 12\% of authors affiliated with the company in
+2020, and having been the single largest affiliation across all years in
+the dataset. We can also see the rise of Huawei beginning in 2005, with
+7.1\% of authors being affiliated with the company in 2020, having peaked
+at 9.7\% in 2018. Google has a similar trajectory, first appearing in the
+dataset in 2006, with 3.8\% of authors being affiliated with it in 2020.
+
+We also observe the decline of a number of affiliations.  Microsoft and
+Nokia, having peaked with 3.3\% and 3.6\% of authors, had 0.7\% and 1.7\%
+of authors in 2020, respectively, with the absolute number of authors from
+both companies also declining.
+
+That we are able to observe changing trends in author affiliations, and in
+particular, that participants from new companies contribute, indicates that
+the IETF has broader relevance beyond a narrow set of companies. While we
+do not demonstrate any causal link, it is encouraging that commercially
+successful companies opt to enable their employees to actively participate
+in the IETF. Care is needed to ensure that this relevance is maintained,
+however: we observe that the author pool has grown less diverse in terms of
+companies that are represented. 35.4\% of authors came from the top 10
+affiliations in the dataset in 2020, compared with 25.6\% in 2001.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/authors/top5_affiliations_normalised.pdf}
+\caption{Authorship affiliations (normalised)}
+\label{fig:author_affiliations_normalised}
+\end{figure}
+
+%..................................................................................................
+\pb{Academia and consultants:}
+Academic affiliations are those where the affiliation name contains
+``University'', ``Institute'', or ``College'', while consultancy
+affiliations are those that contain ``Consultant``. Affiliation data has
+been normalised to remove common abbreviations (e.g., ``U.'' for
+``University'') and to translate non-English affiliations.  As shown in
+Figure~\ref{fig:author_affiliations_normalised}, we find that an increasing
+number of authors come from academic affiliations, growing from 8.1\% of
+authors in 2001, to 13.6\% in 2020, having peaked at 16.5\% in 2009. The
+number of consultants, by our measure, has remained stable, accounting for
+2\% of authors in 2020.  The remaining authors are largely from industrial
+affiliations.
+
+Figure~\ref{fig:author_affiliations_normalised_acad} shows the top 10
+academic affiliations in the dataset, and the percentage of academic
+authors that have those affiliations over time. In general, academic
+affiliations are each typically held by a small number of authors. We can
+see a number of trends in academic authorship, with fewer authors from
+Columbia University, MIT, and ISI in recent years, and the rise of Tsinghua
+University and University Carlos III of Madrid.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/authors/top5_affiliations_normalised_acad.pdf}
+\caption{Academic affiliations (normalised)}
+\label{fig:author_affiliations_normalised_acad}
+\end{figure}
+
+%..................................................................................................
+\pb{Arrival of new authors:}
+Finally, Figure~\ref{fig:author_new} shows the percentage of authors in each
+year that have not previously authored an RFC. Given that the dataset used here
+begins in 2001, 100\% of authors are new in that year. The more stable trend in
+recent years likely highlights the churn in RFC authorship, with around 30\%
+of authors each year having never previously authored an RFC.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/authors/repeat_authors.pdf}
+\caption{Percentage of new authors per year.}
+\label{fig:author_new}
+\end{figure}
+
+%..................................................................................................
+\pb{Summary:}
+The trends highlighted in this section  indicate a pool of authors that is
+diversifying and changing over time, with a growing proportion of authors
+from outside of North America, contributions from new companies and
+affiliations, and relatively high authorship churn. Despite this, we find
+that certain regions and groups are not well represented in the IETF (e.g.,
+contributors from Africa), and that the authorship pool is becoming
+increasingly centralised, with a third of authors coming from the top 10
+affiliations. This suggests, if the organisation is concerned with being
+more representative, further efforts are needed.
 

### 2024-09-21 paper.tex
First version of Characterising Participant Interactions, from IMC 2021, plus structure outline for later parts

@@
 %..................................................................................................
+% IMC 2021 paper Section 3.2 (last part)
 \pb{Arrival of new authors:}
@@
 of authors each year having never previously authored an RFC.
+\todo{Not clear this fits here}
 
@@
 more representative, further efforts are needed.
+
+%==================================================================================================
+\section{Organisational Dynamics of the IETF}
+
+%--------------------------------------------------------------------------------------------------
+\subsection{Characterising Participant Interactions}
+
+% IMC 2021 paper Section 3.3
+
+The operations of the IETF is largely underpinned by mailing list
+interactions, which are used to discuss and finalise the drafts that
+eventually become RFCs. Our data, which starts from 1995, confirms their
+vital role, with 1,153 mailing lists, containing 2,439,240 emails from
+74,646 unique email addresses.
+
+\pb{Volume of Discussion:}
+Figure~\ref{fig:pid_count_emailing_yearly} presents the number of emails
+sent across the last 25 years, showing that email volumes have grown
+significantly with time, plateauing at around 130,000 annual messages since
+around 2010. The figure also shows how the number of unique Person ID
+counts (as described in \S\ref{sec:data}) observed every year in the mail
+lists, showing a decreasing trend in the number of contributors.
+
+Figure~\ref{fig:emailvol_by_year_catg} shows the relative email volume by
+type. We see that an increasing fraction of messages originate from
+automated addresses. This is largely as a result of the growing use of
+GitHub, and other version control systems, for managing drafts. There were
+122 active IETF working groups at the time of writing: of these, 17 listed
+a GitHub repository in their metadata. The QUIC working group, as one
+example, has replaced the typical email list discussions with GitHub
+issues: indeed, this is a significant part of the surge observed in 2016.
+Given that most working groups use mailing lists to manage their activity,
+we do not further analyse interactions that take place on GitHub in this
+paper. As a result, Figure~\ref{fig:emailvol_by_year_catg} understates the
+volume of interactions: the plateau observed in recent years is at least
+somewhat attributable to the shift to GitHub and similar services. As this
+shift continues, it will become important for future work to consider these
+interactions.
+\todo{We have more discussion of GitHub?}
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/pid_count_emailing_yearly.pdf}
+\caption{Number of Person IDs exchanging emails each year}
+\label{fig:pid_count_emailing_yearly}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/frequency_emails_yearly_categories2.pdf}
+\caption{Number of messages exchanged per year across different categories:
+  messages mapped to the Datatracker (Datatracker Person-ID); messages by
+  automated email addresses (Automated); messages by role-based addresses
+  (Role-based); and messages not mapped to the Datatracker (New Person-ID)}
+\label{fig:emailvol_by_year_catg}
+\end{figure}
+
+%..................................................................................................
+\pb{Discussion of drafts:}
+The mailing lists are, other than plenary or interim meetings, the primary
+means for the discussion of drafts.  To measure how often drafts are
+discussed, we next identify mentions of drafts in mailing list messages.
+We therefore extract any mention of a draft (beginning \texttt{draft-}) or
+RFC (i.e., ``RFC'' followed by a number).  Figure~\ref{fig:draft_mentions}
+presents the number of drafts mentions in the emails per year. Separate
+mentions of the same draft are counted as different mentions, as we want to
+observe the entire volume of mentions.  We observe a strong increase in the
+number of mentions over time.  This is largely driven by the growing number
+of drafts being published. In fact, we find a Pearson correlation of 0.89
+between the number of drafts published and the number of mentions. This
+speaks to the influence of emails that mention drafts in driving draft
+production.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/yearly-draft-mention-volume.pdf}
+\caption{Number of draft mentions in each year found in the mail lists}
+\label{fig:draft_mentions}
+\end{figure}
+
+%..................................................................................................
+\pb{Contribution duration.}
+We now look into the longevity of contributors, in terms of the number of
+years that they actively participate in any IETF mailing lists. We define
+the \textit{contribution duration} of a participant as the length of time
+that they have contributed to the mailing lists.  To do this, we look at
+the contributors who first contribute to mailing lists between the years
+2000 to 2013. We limit our analysis to 2013, since the longevity of
+contributors that first participate more recently cannot be determined.
+For each year, in the period 2000 to 2013, we look at those who first
+contribute in each year and the number of years that they then go on to
+remain active in the mailing lists. For example, a participant who first
+sent an e-mail to an IETF list in 2010, and last sent an e-mail to an IETF
+list in 2018, will have a contribution duration of 9 years.
+
+We generate Gaussian Mixture Models for observing different clusters of the
+maximum duration that contributors remain active for.  These models
+identified three broad clusters that contributors can be categorised
+within:
+
+\begin{itemize}
+    \item \emph{Young contributors} who leave within 1 year of joining.
+    \item \emph{Mid-age contributors} who go onto remain associated for
+      more than 1 year, but less than 5 years.
+    \item \emph{Senior contributors} who go onto remain associated for 5 or
+      more years.
+\end{itemize}
+
+We use these contribution categories defined above to characterise the
+interactions between contributors belonging to different categories, as
+well as the authors of each RFC. We look at outgoing and incoming
+interactions with RFC authors in the period between the first draft and
+publication of the RFC. If this period is less than two years, we look at
+the activities of authors for two years before the RFC was published.
+Interactions are defined from the viewpoint of the RFC authors:
+
+\begin{itemize}
+    \item \textit{Outgoing interaction}, where an RFC author responds to an
+      email from other contributors (i.e., email sent).
+    \item \textit{Incoming interaction}, where contributors respond to an
+      email by the author (i.e., email received).
+\end{itemize}
+
+For each interaction type, we count the number of messages and number of
+unique contributors involved in those interactions. For each RFC, we
+analyse the interaction activities based on three measures of author
+contribution duration:
+
+\begin{itemize}
+    \item \emph{Junior-most author}, i.e., the author with the lowest
+      contribution duration at the time of publication.
+    \item \emph{Senior-most author}, i.e., the author with the highest
+      contribution duration at time of publication.
+    \item \emph{Mean contribution duration} of all of the authors of the
+      RFC at the time of publication.
+\end{itemize}
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/age_authors_RFCs.pdf}
+\caption{Contribution duration distribution of authors of RFCs: junior-most
+  author of each RFC, senior-most author of each RFC, and mean
+  contribution-duration of all authors of each RFC}
+\label{fig:age_dist_rfc_authors}
+\end{figure}
+
+Figure~\ref{fig:age_dist_rfc_authors} shows the distribution of
+contribution duration of each of these three measures. This shows that the
+majority of junior-most authors have participated for less than 5 years in
+the IETF, whereas the majority of senior-most authors have participated in
+the IETF for more than 10 years. In fact, 35\% of authors exceed 15 years
+of IETF participation, showing that RFCs tend to be authored by a mix of
+seniority levels.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/rfc_authors_degree.pdf}
+\caption{CDF showing drift in annual degree (interaction with their
+  network) of RFC authors over the period 2000-2020.}
+\label{fig:degree_dist_rfc_authors}
+\end{figure}
+
+%..................................................................................................
+\pb{Evolution of interactions:}
+Figure~\ref{fig:degree_dist_rfc_authors} shows the gradual drift in annual
+degree (i.e., the number of people interacted with) of RFC authors. The
+degree of authors has substantially increased over time. For instance, in
+the year 2000 only 5.5\% of the authors had a degree of over 25, whereas by
+the year 2015 almost a quarter of the authors had a degree over 25.  This
+confirms that, on average, more recent RFCs generate greater discussion.
+This may explain increasing publication times: authors spend more time
+interacting with other participants, and these discussions are likely to
+lead to more drafts.
+\todo{Replace with material from one of the later papers?}
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/junior_senior_authors_seniormem_indegree.pdf}
+\caption{CDF showing number of senior contributors sending messages
+  (in-degree) to junior and senior authors.}
+\label{fig:junior_senior_indegree_seniormem}
+\end{figure}
+
+We can also contrast the interaction patterns of the junior vs.\ senior
+authors. Figure~\ref{fig:junior_senior_indegree_seniormem} presents the
+CDFs of the in-degree across both junior and senior authors.  It shows that
+the incoming interactions from senior contributors to junior authors are
+significantly less than the incoming interactions from senior contributors
+to senior authors. Nearly 55\% of junior authors receive messages from
+fewer than 10 senior contributors, whereas nearly 65\% of senior authors
+receive messages from more than 10 senior contributors.  This speaks to the
+differing roles played by these sub-populations: senior authors act as hubs
+through which substantial volumes of interaction flow.
+\todo{Replace with material from one of the later papers?}
+
+%..................................................................................................
+\pb{Summary:}
+Emails play a vital role in underpinning RFC publication with approximately
+130,000 emails sent per year. We have shown that the seniority of both
+participants and RFC authors fundamentally changes the volume of
+interactions that they have. These trends are likely to have implications
+for the IETF community, especially as it tries to encourage new
+participants.
+
+%--------------------------------------------------------------------------------------------------
+\subsection{Influence}
+
+% ICWSM 2022 paper
+% - Measuring influence
+% - Behaviour of influential participants
+% - Impact of influential participants
+
+%--------------------------------------------------------------------------------------------------
+\subsection{Hierarchy}
+%..................................................................................................
+% ICWSM 2024 paper
+% - RQ1: Are higher levels of the hierarchy growing or becoming more
+%   centralised? Are they associated with an increased domination of the
+%   conversation? 
+% - RQ2: What is the association between the organisation hierarchy and
+%   general communication patterns?
+% - RQ3: How do people with differing roles communicate with each other?
+%   Does information flow “up” or “down” the hierarchy? 
+% - RQ4: What is the impact that individuals have on their direct contacts’
+%   activity over time? How does this vary across hierarchy level?
+
+%..................................................................................................
+% LEDA
+% - a taxonomy of dialogue acts (DAs) and a labeled dataset of emails
+
+%==================================================================================================
+\section{Understanding Participant Behaviour}
+
+% Tracing Linguistic Markers
+% - RQ1: How do linguistic traits differ between more and less influential
+%   participants? 
+% - RQ2: How do linguistic traits vary for participants at different levels
+%   of the organisation hierarchy? 
+% - RQ3: How does linguistic behaviour of participants change as they gain
+%   influence?
+
+% Frontiers in Psychology paper
+% - Use of Language, discussing how it shifts as people become more senior
+
+%==================================================================================================
+\section{Success Factors}
+
+\subsection{For authors}
+% ICWSM 2022 paper Section 4
+
+\subsection{For documents}
+
+% IMC 2021 paper Section 4
 

### 2024-09-22 paper.tex
Add some to-do notes
Add §5.2 on Influence, from ICWSM 2022 paper

@@
        with Computational Methods}
-\author{Colin Perkins\\University of Glasgow}
+\author{Colin Perkins \and
+        Prashant Khare \and
+        Mladen Karan \and
+        Stephen McQuistin \and
+        Waleed Iqbal \and
+        Junaid Qadir \and
+        Matthew Purver \and
+        Patrick Healey \and
+        Ignacio Castro \and
+        Gareth Tyson}
+
 \maketitle
@@
 \section{Trends in RFC Publication}
+\label{sec:trends-documents}
 
@@
 \end{figure}
+
+\todo{Add draft publication rate as mentioned in Section \ref{sec:org-dynamics:participants}}
 
@@
 \section{Organisational Dynamics of the IETF}
+\label{sec:org-dynamics}
 
 %--------------------------------------------------------------------------------------------------
-\subsection{Characterising Participant Interactions}
+\subsection{Interactions}
+\label{sec:org-dynamics:participants}
 
@@
 production.
+\todo{Add discussion of number of drafts published over time in Section
+\ref{sec:trends-documents}. Does this tie into the discussion of increasing
+complexity of draft publication?}
 
@@
 \subsection{Influence}
+\label{sec:org-dynamics:influence}
 
@@
 
+% This is modified from the Introduction to the ICWSM 2022 paper:
+
+As we showed in Section \ref{sec:org-dynamics:participants}, protocol
+standardisation is an inherently social process with most day-to-day work
+happening on public mailing lists, aided by meetings, video conference
+calls, and open document and code repositories.  We are specifically
+interested in better understanding how influence is distributed across
+stakeholders and how it might affect the standardisation process.  This is
+of critical societal importance: the IETF has a major impact on global
+Internet technologies, and understanding the social processes involved
+would give us insight into not only the driving forces behind
+standardisation, but also its resilience to the loss of major participants.
+In the following we consider the following \emph{research questions}: 
+\begin{enumerate}
+    \item How centralised is the active IETF community, and to what extent
+      is it reliant on a small core of participants? 
+    \item How do the most influential participants behave? 
+    \item How does influence (determined by mailing list participation)
+      relate to wider impacts throughout the IETF?
+    \item Does the organisational affiliation of participants also
+      influence the innovation (adoption of new work) within IETF?
+\end{enumerate}
+
+% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+% ICWSM 2022 paper Section 3
+\subsubsection{The IETF as a Social Graph}
+
+\label{subsec:measuring_influence}
+
+Much of the interaction between IETF participants occurs on public mailing
+lists.  For each year, we build a social graph based on the \emph{active
+community} of participants (nodes) who have interactions (edges) with any
+other participant in the previous 5 years.
+
+In Section \ref{sec:org-dynamics:participants}, we demonstrated that there
+are three categories of participants in IETF - (a) \textit{young
+contributors} -- who leave within one year of their first year mailing list
+contribution; (b) \textit{mid-age contributors} -- participants who stay
+active for up-to 5 years; and (c) \textit{senior contributors} -- who
+remain active for more than 5 years. Following this, a 5 year period window
+is chosen to observe interactions. There are no direct
+participant-to-participant email exchanges in our dataset: the emails and
+responses we capture are sent to the public mailing lists, and we observe
+an interaction between two participants when one replies to an email sent
+by another on any mailing list. This yields a social graph based on
+1,049,793 emails (out of 2.1M) from 22,138 unique participants across 840
+mailing lists.
+
+%..................................................................................................
+% ICWSM 2022 paper Section 3.1
+\pb{Participation}
+To examine the variation in participation between IETF participants, in
+Figure~\ref{fig:degree_cdf} we plot the cumulative degree distribution of
+the social graph -- i.e., the cumulative number of people a participant
+interacts with.  We observe a core group that always interacts with
+substantially more people than the rest of the community, with around 80\%
+of the participants having degree  $<$15, but 5\%-10\% having degree $>$40.
+However,  this difference decreases over time: in 2005-2009 around 6\% of
+participants had degree $>$40, but by the period 2015-2019 this increases
+to $\sim10\%$.  This shows that while there is always a core of more active
+participants, the degree of participation has spread  out over time.
+\todo{Repetitive with Section \ref{sec:org-dynamics:participants}?}
+
+\begin{figure}
+\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/degree_CDF.pdf}
+\caption{Cumulative degree distribution of the email graph for different year-periods.}
+\label{fig:degree_cdf}
+\end{figure}
+
+To understand the structure of the community, and its reliance on specific
+groups, we analyse the connected components of this graph. Each connected
+component reflects a maximal set of nodes such that each pair of nodes is
+connected by a path. We compute the size of the Largest Connected Component
+(LCC) and the Number of Connected Components (NCC) for each year in
+Figure~\ref{fig:lcc_size_yearly}.  The NCC peaks in 2003 before declining,
+and broadly aligns with the variation in the number of meeting
+attendees,\footnote{\url{https://datatracker.ietf.org/stats/meeting/overview/}}
+suggesting a lag between participating for the first time and integrating
+into the wider community.  In contrast, the size of the LCC increases until
+2006 before stabilising.  Overall, we observe that the IETF community has
+become less fragmented.
+
+\begin{figure}
+\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lcc_yearly_no_components.pdf}
+\caption{Size of Largest Connected Component (LCC) and Number of Connected Components (NCC)}
+\label{fig:lcc_size_yearly}
+\end{figure}
+
+%..................................................................................................
+% ICWSM 2022 paper Section 3.1
+\pb{Influence}
+Relying on a small group to interconnect the community could undermine the
+resilience of the IETF.  To study the influence of participants and their
+role in interconnecting the wider community, we compute the betweenness
+centrality of each participant in a given time
+period~\cite{kourtellis2013identifying, weitzel2012measuring,
+sole2014centrality}. We also considered other graph based influence metrics
+such as eigenvector centrality: eigenvector centrality reflects the
+importance of a node as per its neighbours, while betweenness centrality is
+based on shortest paths which is independent of the influence of
+neighbours.
+
+However, we found a very strong correlation between the two measures, 
+in line with similar experiments \cite{valente2008correlated,he2016correlation}:
+the Spearman's rank correlation of participants ranked by betweenness
+centrality and eigenvector centrality ranged between 0.51-0.72, with a
+strong statistical significance ($p < 0.01$) for the period 2000-2019. We
+therefore use just betweenness centrality in our analysis; this is often
+studied and acknowledged as a measure of influence in social and complex
+networks, particularly built on online communication
+\cite{hagen2018crisis,chen2012identifying, ghalmane2019centrality}.
+
+Figure~\ref{fig:purge_nodes_largest_connected_component_number_components}
+then shows the effect of removing the most influential participants (in 1\%
+increments from most to least, moving left-to-right on the x-axis) on the
+size of the LCC.  Worryingly, we find that removal of just 20-25\% of the
+most influential participants causes the LCC to shrink by 90\%.  However,
+we also find that this impact has decreased over time: for instance, in
+2000-2004, removing the top $\sim$5\% influential participants, reduces the
+size of LCC by more than half, whereas in 2015-2019, it takes the removal
+of the top $\sim$15\% of participants to have the same effect.  This shows
+that the community has become \emph{more} cohesive and resilient over time,
+and the IETF can now sustain a larger amount of churn while maintaining a
+well-connected social graph.
+
+\begin{figure}
+\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lcc_proportion_granular_yearly.pdf}
+\caption{Impact of removing participants by their influence on the size of the LCC.}
+\label{fig:purge_nodes_largest_connected_component_number_components}
+\end{figure}
+
+% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+% ICWSM 2022 paper Section 3.2
+\subsubsection{Behaviour of Influential Participants}
+\label{subsec:influencer_behaviour}
+
+\todo{This is somewhat repetitive with Section \ref{sec:org-dynamics:participants}}
+
+We now characterise the behaviour of the most influential participants, in
+terms of the volume of emails sent, length of time active within the
+community, and topics discussed.
+
+%..................................................................................................
+\pb{Email volume}
+Figure~\ref{fig:emailcount_top_percentile} shows that each participant in
+the top 10\% most influential participants sends on average around
+0.05\%-0.08\% of total emails in a given period.  Collectively, the top
+10\% most influential participants account for 43.75\%, on average, of the
+total emails, a substantially larger proportion than the others.
+
+This dynamic seems stable over time, making differences even more acute.
+At the same time, the overall number of emails sent increases up to 2010
+and remains roughly stable from then onward.  Along with the results from
+\S\ref{subsec:measuring_influence}, this shows a worrying, if slowly
+improving, dependence of the IETF on a small number of influential
+participants.
+
+\begin{figure}
+\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lineplot_proportion_emailscount_top_percentile.pdf}
+\caption{Proportion of emails sent (\%) by participants in each period 
+(with 95\% confidence interval) according to their betweenness-centrality percentile (x-axis).
+Y2-axis (purple) shows yearly count of emails.}
+\label{fig:emailcount_top_percentile}
+\end{figure}
+
+%..................................................................................................
+\pb{Cross-area review} 
+In addition to sending more emails, we test if influential participants
+engage with different parts of the community more. The work of the IETF is
+divided into several areas (e.g., Applications \& Real-time, Security, and
+Routing). This eases administration, but might also act as a barrier to
+broader discussion and review. For example, WebRTC standards developed in
+the Applications and Real-time area might contain elements that could
+benefit from expertise from the Security area, but participants in one area
+might not review work in another.
+
+Figure~\ref{fig:areas_participation_top_percentile} shows the mean number
+of areas where IETF participants are active (derived from the mailing lists
+they use).  We see that influential participants engage in more areas of
+the IETF, on average, and that cross-area engagement has improved over
+time, indicating that the community itself has matured. This indicates that
+influential participants benefit the IETF in enabling cross-area discussion
+and review, and can bridge administrative divisions.
+
+\begin{figure}
+\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lineplot_avg_areas_top_percentile.pdf}
+\caption{Mean number of areas participated in (with 95\% confidence
+  interval) according to their betweenness-centrality percentile in the
+  x-axis, ranked from  top (0-10) to bottom percentile (91-100).}
+\label{fig:areas_participation_top_percentile}
+\end{figure}
+
+%..................................................................................................
+\pb{Participation duration}
+We next consider for how long participants remain associated with the IETF
+(measured as the delta between first and last emails sent).
+Figure~\ref{fig:age_top_percentile} shows the mean participation duration
+distribution for participants, ranked by influence.  Participation duration
+increases over time, with the most influential participants typically being
+those who have been active for longer. While this might be expected, it
+might be a good sign for the IETF community in that it implies that the
+most influential participants are also the most experienced.
+
+\begin{figure}
+\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lineplot_age_top_percentile.pdf}
+\caption{Mean participation duration (with 95\% confidence interval)
+  according to their betweenness-centrality percentile in the x-axis,
+  ranked from  top (0-10) to bottom percentile (91-100).}
+\label{fig:age_top_percentile}
+\end{figure}
+
+We extend this observation by considering what happens to participants once
+they become influential. Figure \ref{fig:heatmap_top10_overlap} shows what
+proportion of the top 10\% influential participants in any given year
+(\emph{y-axis}) was also among the top 10\% influential participants in any
+other year (\emph{x-axis}). This shows that a significant majority of
+participants that become influential, continue to be influential for a
+number of years: the top 10\% are influential for at least 6-7 years on
+average.
+
+\begin{figure}
+\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/heatmap_top10_percent_overlap.pdf}
+\caption{Yearly overlap (\%) of top 10\% influencers.}
+\label{fig:heatmap_top10_overlap}
+\end{figure}
+
+Further, Figure \ref{fig:rate_new_influential_participants} shows that
+breaking into the top 10\% of influencers requires an increasing number of
+years of participation. This may be beneficial, showing that the IETF is
+maturing and is capable of retaining influential and experienced
+participants, but it may also point to an increasingly ossified structure
+that is not welcoming to newcomers.
+
+\begin{figure}
+\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/new_influencers_top10_age_longevityInTop10.pdf}
+\caption{Percentage of new influential participants breaking into top 10\%
+  influencers (black); the average years of participation before entering
+  top 10\% (purple); and minimum \#  years new influencers remain in top
+  10\% (with 95\% confidence intervals).}
+\label{fig:rate_new_influential_participants}
+\end{figure}
+
+%..................................................................................................
+\pb{Topics of Discussion} 
+The topics discussed on WG mailing lists are a good indicator of the focus
+of technical work, and it might be expected that influential participants
+will set the direction of that work.  To explore this, we use the Latent
+Dirichlet Allocation (LDA) \cite{blei2003latent,hoffman2010online} model
+from \texttt{gensim} \cite{rehurek_lrec} to induce 100 topics on the entire
+set of email texts. 
+Each topic is a distribution over words: e.g., a \emph{security} topic might
+have high probability for \emph{cypher}, \emph{rsa}, \emph{auth}, and
+related words. We can also use the model to obtain a vector for any input
+text as a sparse distribution over all topics, e.g., finding that a text is
+comprised of 30\% security and 70\% video streaming topics. With this, we
+generate a vector for each participant in a given time period by
+concatenating all messages sent by that participant in the period and
+feeding it into the LDA model. The result can be loosely interpreted as a
+distribution of topics on which each participant works.
+
+Manual inspection of the induced topics reveals some interesting trends. We
+find that topics related to \emph{routing} and \emph{email} protocols are
+seeing a steady decline in popularity, while topics like \emph{streaming}
+and \emph{cryptography} are becoming more prominent. This reflects wider
+trends in standardisation, as these efforts have adjusted to the public's
+increased awareness of privacy, especially in light of the Edward Snowden
+leaks \cite{RFC7258}. We also find that as videoconferencing became
+increasingly popular, so did the WebRTC protocol that frequently underpins
+them.
+
+We next measure the topical diversity of a participant by observing the
+entropy of their topic vector, their \emph{topic entropy}, defined as
+follows:
+\begin{equation}
+    \eta = - \sum_{i=1}^{|T|} p_i \log p_i \nonumber
+\end{equation}
+where $T = [p_1, ..., p_N]$ is a topic vector, which defines a probability
+distribution over $N$ topics, each $p_i$ is a probability of the $i$-th
+topic and $\sum_i p_i = 1$.
+
+Figure \ref{fig:topic_diversity} shows topic entropy distributions of
+participants in different time periods and influence percentiles. While we
+initially experimented with measuring diversity by simply counting the
+number of topics that account for the majority ($ \geq 95\% $) of a
+participant's topic distribution probability mass, we found that most
+participants engage in a relatively small number of topics, regardless of
+influence. However, after measuring diversity using entropy, we identify
+that the activity of the more influential participants tends to be more
+evenly spread across the topics they participate in\footnote{An example
+with 3 topics: a distribution of $[0.8, 0.1, 0.1]$ is less evenly spread
+than $[0.3, 0.4, 0.3]$.}. This difference becomes more pronounced over
+time, implying that increased topic entropy (i.e.,~more evenly participating
+in different topics) is an increasingly salient property of influencers.
+This is aligned with our earlier findings on cross-area review, showing the
+growing number of areas in which influencers participate.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/icwsm-2022/topic-div.png}
+\caption{Mean topic entropy (with 95\% confidence interval) according to
+  their betweenness-centrality percentile in the x-axis, ranked from  top
+  (0-10) to bottom percentile (91-100).}
+\label{fig:topic_diversity}
+\end{figure}
+
+% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+\subsubsection{Impact of Influential Participants}
+\label{subsec:impact_influence}
+
+% ICWSM 2022 paper Section 3.3
+
+While these top participants are influential within the \emph{mailing
+lists}, it is unclear how this influence translates into document
+authorship and into gaining  leadership roles within  the organisation.
+
+%..................................................................................................
+\pb{Draft authorship}
+The output of the IETF is technical standards and other documents published
+in the RFC series. RFCs are developed by WGs from a sequence of Internet-Drafts,
+with individuals acting as named authors.  Having identified the most
+influential participants in the mailing list community, we can determine
+whether or not these same participants are also the most active authors.
+
+Figure~\ref{fig:drafts_initiation_top_percentile} shows the distribution of
+the proportion of documents authored by mailing list participants, sorted
+by their influence rank. We observe a growing proportion of the community
+involved in draft authoring, skewed towards influential participants who
+tend to write more drafts.  During 2010-2019, each participant in the top
+10 percentile, authored 0.125\% to 0.175\% of the total number of drafts in
+that period. We find that 32\%-42\% of the total drafts, in different years
+during this period, were authored by the top 10\% most influential mailing
+list participants.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/icwsm-2022/documents_proportion_published_top_percentile.pdf}
+\caption{Average drafts authored (with 95\% confidence interval) according
+  to their betweenness-centrality percentile in the x-axis, ranked from
+  top (0-10) to bottom percentile (91-100).  Right-y axis (purple) shows
+  yearly drafts per year.}
+\label{fig:drafts_initiation_top_percentile}
+\end{figure}
+
+%..................................................................................................
+\pb{Co-authorship and email graph correlation}
+Figure~\ref{fig:drafts_initiation_top_percentile} showed that influential
+mailing list participants tend to author more drafts than others.
+We next ask if they are also influencers in the draft co-authorship graph.
+
+Thus, we create a draft co-authorship graph where each author is a node,
+and draft co-authorship is an edge. We then measure influence with
+betweenness centrality of the authors in each time period.
+Table~\ref{tbl:correlation_sprmn} compares the top 20\% of influencers of
+the mailing lists with those of the co-authorship graph.  We find a
+significant overlap between both groups, ranging from 48.2\% to 67.2\%.
+There is a significant ($p < 0.05$) positive correlation between the
+rankings of the overlapping members of each community: participants that
+are influential in the mailing lists are also likely to be influential in
+draft authorship.
+
+\begin{table}
+\small 
+\begin{tabular}{llllll}
+\toprule
+\multicolumn{6}{c}{Top 20\% draft authors \& all email participants}\\ \midrule
+
+Years   & \multicolumn{2}{c}{Sub-Network Size} & Overlap  & \multicolumn{2}{c}{Spearman's $r_{s}$} \\ \cline{2-3}  \cline{5-6}
+    &   Co-author & Email  &  &   $r_{s}$   & $p$-value \\ \midrule
+
+2000-04   &   398 &   1390 &   48.24\%  &  .323 &   4.75e-06 \\
+2005-09   &   427 &   1662 &   67.21\% &   .332   &   7.39e-09    \\
+2010-14   &   728 &   1639 &   63.05\% &   .299   &   5.73e-11  \\
+2015-19   &   915 &   1370 &   55.85\%  &   .337   &   4.13e-15  \\ %\hline
+
+\bottomrule
+\end{tabular}
+\caption{Overlap in the co-author and email graph.}
+\label{tbl:correlation_sprmn}
+\end{table}
+
+We also look at the top 20\% participants from the co-authorship network
+who are \emph{not} part of the top 20\% influencers in the email network in
+Figure \ref{fig:non_overlapping_coauthornetwork_authors_emailnetwork}.
+We observe that not all the prolific authors are that engaged in the email
+discussion: 40\%-50\% of non-overlapping authors are ranked between 20th to
+40th percentile of influence in the email network. These non-overlapping
+authors are typically more junior with respect to their participation
+duration.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/lineplot_non_overlapping_authors_emailnetwork_percentile.pdf}
+\caption{Percentage of Top 20\% authors that are not 20\% influencers
+  according to their betweenness-centrality (email graph) percentile in the
+  x-axis, ranked from  top (0-10) to bottom percentile (91-100).}
+\label{fig:non_overlapping_coauthornetwork_authors_emailnetwork}
+\end{figure}
+
+%..................................................................................................
+\pb{Leadership roles} 
+WG chairs are selected from the community, and we might expect those
+selected to be influential in the community.
+Figure~\ref{fig:wg_chair_before_after} shows that this is indeed the case:
+87.5\% of WG chairs are in the top 20\% of mailing list influencers and
+67.4\% are in the top 20\% of document authors in the year before they
+became chairs. This also shows the impact of taking up a leadership role:
+influence in both the mailing list and authorship communities grows in the
+year after participants become a WG chair for the first time.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures-prev/icwsm-2022/WG_chair_percentile_CDF_before_after.pdf}
+ \caption{CDF of the percentile of betweenness-centrality of WG chairs in
+  the email and co-authorship graph, one year before and after becoming
+  chairs.}
+\label{fig:wg_chair_before_after}
+\end{figure}
+
+%..................................................................................................
+\pb{Organisations}  While participants in the IETF contribute as
+individuals, they are usually affiliated with an organisation.  To study
+the potential influence of organisations, for each draft we obtain the
+authors' affiliations from the Datatracker. If this is not available, we
+use the domain name of the authors' email addresses, which we map to the
+relevant organisation (e.g., \texttt{@cisco.com} maps to Cisco). For generic
+email addresses, such as \texttt{@gmail.com}, we use the participant name
+if no affiliation is available.
+
+Figure \ref{fig:coauthor_affiliation_network2019_otheryears} shows the ten
+most frequent affiliated organisations of the top 20\% authors in the
+period 2015-2019, and the number of authors affiliated with these
+organisations in  each  period.  While the dominance of Cisco is clear,
+other organisations, such as Huawei, have gained a larger presence over
+time.  In general, a small number of organisations employ a significant
+fraction of the influential participants in the standards process. If we
+look at the period 2015-2019, for example, out of the 915 authors ranked in
+the top 20\% of most influential authors, 342 belong to the ten most
+frequently affiliated organisations, and nearly 253 are from Cisco, Huawei,
+Ericsson, or Juniper alone.
+
+\begin{figure}
+\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/2019_affiliations_in_otheryears_coauthor_network.pdf}
+\caption{Authors in the most common affiliations of top 20\% authors
+  (co-authorship graph) in 2015-2019.}
+\label{fig:coauthor_affiliation_network2019_otheryears}
+\end{figure}
+
+While authors from the same organisation often co-author drafts together,
+collaborations between authors from different organisations are also
+common.  Figure~\ref{fig:heatmap_2019_coauthordraft_most_frequent_org}
+shows the most common collaborations between authors from the top 15 most
+frequently affiliated organisations.  Collaborations between authors at
+competing organisations, such as between authors from Cisco, Huawei,
+Ericsson, and Juniper, occur frequently. We also computed that over the
+years, joint collaboration (authors from multiple affiliations) has
+increased: 772 co-authored drafts in the period 2000-2004 vs. 3083
+co-authored drafts in the period 2015-2019.
+
+\begin{figure}
+\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/heatmap_2019_top10_affiliation_collaboration.pdf}
+\caption{Number of drafts co-authored by top 15 most frequently affiliated
+  organisations. We only include the top 20\% authors (co-authorship graph)
+  of 2015-2019.}
+\label{fig:heatmap_2019_coauthordraft_most_frequent_org}
+\end{figure}
+
+Further inspection of the collaboration trends reveals that the volume of
+collaborations between authors from Huawei and other organisations appears
+to be unaffected by the recent addition of Huawei to the U.S. Entity List
+\cite{bis:2019:entity} as the trends in Figure~\ref{fig:heatmap_2019_coauthordraft_most_frequent_org}
+also hold for 2020-2021.\footnote{\tiny\url{https://mailarchive.ietf.org/arch/msg/ietf-announce/0ywjgSS4LlO0DaWDoLJLRHxJdUk/}}
+
+% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+\subsubsection{Discussion}
+
+% ICWSM 2022 paper section 4.3
+
+We conclude that the IETF is still reasonably centralised, but that this
+has improved over the years. However, removing around 20\%-25\% most
+influential participants still fragments the entire network
+(\S\ref{subsec:measuring_influence}, Figure
+\ref{fig:purge_nodes_largest_connected_component_number_components}). The
+most influential people are the ones with highest degree of engagements
+within the community who also tend to be more involved in draft authoring
+activities (\S\ref{subsec:influencer_behaviour}, Figure
+\ref{fig:drafts_initiation_top_percentile}). However, we also observe that
+over the years more people (even the ones less influential in the email
+network) are getting involved in draft authoring activities (Figure
+\ref{fig:drafts_initiation_top_percentile}). This shows that these
+activities, while centralised to a certain degree, are still open to
+contributions from the wider IETF community.
+
+We observe that the influential participants show a much higher level of
+engagement with the community (\S\ref{subsec:influencer_behaviour}). A
+considerable proportion of total emails is sent by the top 10\% of
+influential participants (Figure \ref{fig:emailcount_top_percentile}).
+Compared to the rest of participants, they are active in more areas (Figure
+\ref{fig:areas_participation_top_percentile}), are active within the IETF
+for a longer period of time (Figure \ref{fig:age_top_percentile}), and
+participate in a more diverse set of topics (Figure
+\ref{fig:topic_diversity}). Finally, their influence extends from the
+mailing lists to other activities such as draft authorship.
+
+A significant overlap and correlation is observed between
+influential authors from co-authorship network and email networks
+(\S\ref{subsec:impact_influence}, Table \ref{tbl:correlation_sprmn}). This
+shows that a large set of authors exhibit an ability to co-author drafts as
+well as engage with participants on the email networks, thereby translating
+their influence from email networks to co-authorship and vice versa.
+
+A considerable portion of influential participants ($\sim$30\%) are
+affiliated to one of the more prominent organisations (e.g., \emph{Cisco}
+or \emph{Ericsson}) (\S\ref{subsec:impact_influence}, Figure
+\ref{fig:coauthor_affiliation_network2019_otheryears}). Participants from
+different organisations do considerably collaborate (Figure
+\ref{fig:heatmap_2019_coauthordraft_most_frequent_org}). Moreover, the
+level of such collaboration is found to have increased with time.
+
 %--------------------------------------------------------------------------------------------------
 \subsection{Hierarchy}
-%..................................................................................................
+
 % ICWSM 2024 paper
@@
 % ICWSM 2022 paper Section 4
+
+% ICWSM 2022 paper section 4.3
+The statistical analysis shows that influence of draft authors in the email
+networks does impact the possibility of a draft getting adopted
+(\S\ref{subsec:results}, Table \ref{tbl:res} \& \ref{tbl:resstat}). This
+might hint at the ability of participants who hold a domain expertise to be
+able to engage better with the community. Several WG chairs are already in
+the top percentile influential category in both the email and co-authorship
+networks before taking up these leadership roles, which further elevates
+after taking up such leadership roles (Figure
+\ref{fig:wg_chair_before_after}).
+
+Most importantly the statistical analysis shows that being affiliated with
+a prominent organisation positively impacts the chances of a draft in
+getting adopted by a WG, thereby directly driving the innovation process
+(Table \ref{tbl:res} \& \ref{tbl:resstat}).
 

### 2024-09-22 paper.tex
Tweak section title
Tidy (no content changes)

@@
        with Computational Methods}
-\author{Colin Perkins \and
+\author{Mladen Karan \and
         Prashant Khare \and
-        Mladen Karan \and
-        Stephen McQuistin \and
         Waleed Iqbal \and
         Junaid Qadir \and
+        Stephen McQuistin \and
+        Patrick Healey \and
         Matthew Purver \and
-        Patrick Healey \and
         Ignacio Castro \and
-        Gareth Tyson}
+        Gareth Tyson \and
+        Colin Perkins}
 
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/rfcs_areas.pdf}
-\caption{RFCs by area. \textit{Other} includes legacy RFCs, and RFCs from other non-IETF streams.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/rfcs_areas.pdf}
+  \caption{
+    RFCs by area. \textit{Other} includes legacy RFCs, and RFCs from other
+    non-IETF streams.
+  }
   \label{fig:rfcs_by_area}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/unique_wgs_per_year_areas.pdf}
-\caption{Number of publishing working groups. \textit{Other} includes legacy RFCs, IRTF research groups, and non-IETF streams.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/unique_wgs_per_year_areas.pdf}
+  \caption{
+    Number of publishing working groups. \textit{Other} includes legacy
+    RFCs, IRTF research groups, and non-IETF streams.
+  }
   \label{fig:pub_wgs_yearly}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/day_counts_yearly.pdf}
-\caption{Days from first draft to RFC publication}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/day_counts_yearly.pdf}
+  \caption{
+    Days from first draft to RFC publication
+  }
   \label{fig:rfcs_days_to_pub}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/draft_counts_yearly.pdf}
-\caption{Number of drafts per RFC}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/draft_counts_yearly.pdf}
+  \caption{
+    Number of drafts per RFC
+  }
   \label{fig:drafts_year}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/page_counts_yearly_dt.pdf}
-\caption{RFC page counts}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/page_counts_yearly_dt.pdf}
+  \caption{
+    RFC page counts
+  }
   \label{fig:pages_year}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/update_obsolete_yearly_pct.pdf}
-\caption{RFCs that update or obsolete previous RFCs}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/update_obsolete_yearly_pct.pdf}
+  \caption{
+    RFCs that update or obsolete previous RFCs
+  }
   \label{fig:updates_year}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/cite_counts_yearly.pdf}
-\caption{Citations from RFCs to other Internet-Drafts and RFCs per RFC}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/cite_counts_yearly.pdf}
+  \caption{
+    Citations from RFCs to other Internet-Drafts and RFCs per RFC
+  }
   \label{fig:citations_year}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/documents/keyword_usage_rate.pdf}
-\caption{Keyword occurrences per page}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/keyword_usage_rate.pdf}
+  \caption{
+    Keyword occurrences per page
+  }
   \label{fig:keyword_usage_rates}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/tma-2023/errata-by-year.pdf}
-\caption{Average errata filed per RFC for each year by RFC publication year,
-grouped by IETF area (acronyms expanded in Table~\ref{tab:errata-stats-area}).}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/tma-2023/errata-by-year.pdf}
+  \caption{
+    Average errata filed per RFC for each year by RFC publication year,
+    grouped by IETF area (acronyms expanded in Table~\ref{tab:errata-stats-area}).
+  }
   \label{fig:errata_per_year}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/tma-2023/errata-submission-dates-area.pdf}
-\caption{Days from RFC publication and errata filing by IETF area (acronyms
-  expanded in Table~\ref{tab:errata-stats-area}).}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/tma-2023/errata-submission-dates-area.pdf}
+  \caption{
+    Days from RFC publication and errata filing by IETF area (acronyms
+    expanded in Table~\ref{tab:errata-stats-area}).
+  }
   \label{fig:errata_submission_days}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/tma-2023/errata-by-status.pdf}
-\caption{Errata filings by status, by publication year of the RFC.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/tma-2023/errata-by-status.pdf}
+  \caption{
+    Errata filings by status, by publication year of the RFC.
+  }
   \label{fig:errata_status}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/tma-2023/errata-by-rfc-by-area.pdf}
-\caption{CDF of errata filed per RFC by grouped by IETF area (acronyms
-  expanded in Table~\ref{tab:errata-stats-area}).}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/tma-2023/errata-by-rfc-by-area.pdf}
+  \caption{
+    CDF of errata filed per RFC by grouped by IETF area (acronyms expanded
+    in Table~\ref{tab:errata-stats-area}).
+  }
   \label{fig:errata_per_rfc}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/tma-2023/location-percent.pdf}
-\caption{Errata counts by percentile location in document (0 is the beginning; 100 is the end).}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/tma-2023/location-percent.pdf}
+  \caption{
+    Errata counts by percentile location in document (0 is the beginning;
+    100 is the end).
+  }
   \label{fig:errata_location_percent_count}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/tma-2023/section-title-counts.pdf}
-\caption{Errata counts by section title for the more frequent section titles.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/tma-2023/section-title-counts.pdf}
+  \caption{
+    Errata counts by section title for the more frequent section titles.
+  }
   \label{fig:errata_section_wise_counts}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/authors/top5_countries_normalised.pdf}
-\caption{Authorship countries (normalised)}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/authors/top5_countries_normalised.pdf}
+  \caption{
+    Authorship countries (normalised)
+  }
   \label{fig:author_countries_normalised}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/authors/continents_normalised.pdf}
-\caption{Authorship continents (normalised)}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/authors/continents_normalised.pdf}
+  \caption{
+    Authorship continents (normalised)
+  }
   \label{fig:author_continents_normalised}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/authors/top5_affiliations_normalised.pdf}
-\caption{Authorship affiliations (normalised)}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/authors/top5_affiliations_normalised.pdf}
+  \caption{
+    Authorship affiliations (normalised)
+  }
   \label{fig:author_affiliations_normalised}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/authors/top5_affiliations_normalised_acad.pdf}
-\caption{Academic affiliations (normalised)}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/authors/top5_affiliations_normalised_acad.pdf}
+  \caption{
+    Academic affiliations (normalised)
+  }
   \label{fig:author_affiliations_normalised_acad}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/authors/repeat_authors.pdf}
-\caption{Percentage of new authors per year.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/authors/repeat_authors.pdf}
+  \caption{
+    Percentage of new authors per year.
+  }
   \label{fig:author_new}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/pid_count_emailing_yearly.pdf}
-\caption{Number of Person IDs exchanging emails each year}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/emails/pid_count_emailing_yearly.pdf}
+  \caption{
+    Number of Person IDs exchanging emails each year
+  }
   \label{fig:pid_count_emailing_yearly}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/frequency_emails_yearly_categories2.pdf}
-\caption{Number of messages exchanged per year across different categories:
-  messages mapped to the Datatracker (Datatracker Person-ID); messages by
-  automated email addresses (Automated); messages by role-based addresses
-  (Role-based); and messages not mapped to the Datatracker (New Person-ID)}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/emails/frequency_emails_yearly_categories2.pdf}
+  \caption{
+    Number of messages exchanged per year across different categories:
+    messages mapped to the Datatracker (Datatracker Person-ID); messages
+    by automated email addresses (Automated); messages by role-based
+    addresses (Role-based); and messages not mapped to the Datatracker
+    (New Person-ID)
+  }
   \label{fig:emailvol_by_year_catg}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/yearly-draft-mention-volume.pdf}
-\caption{Number of draft mentions in each year found in the mail lists}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/emails/yearly-draft-mention-volume.pdf}
+  \caption{
+    Number of draft mentions in each year found in the mail lists
+  }
   \label{fig:draft_mentions}
@@
 \begin{itemize}
-    \item \textit{Outgoing interaction}, where an RFC author responds to an
-      email from other contributors (i.e., email sent).
-    \item \textit{Incoming interaction}, where contributors respond to an
-      email by the author (i.e., email received).
+  \item \textit{Outgoing interaction}, where an RFC author responds
+    to an email from other contributors (i.e., email sent).
+
+  \item \textit{Incoming interaction}, where contributors respond to
+    an email by the author (i.e., email received).
 \end{itemize}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/age_authors_RFCs.pdf}
-\caption{Contribution duration distribution of authors of RFCs: junior-most
-  author of each RFC, senior-most author of each RFC, and mean
-  contribution-duration of all authors of each RFC}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/emails/age_authors_RFCs.pdf}
+  \caption{
+    Contribution duration distribution of authors of RFCs:
+    junior-most author of each RFC, senior-most author of
+    each RFC, and mean contribution-duration of all authors
+    of each RFC
+  }
 \label{fig:age_dist_rfc_authors}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/rfc_authors_degree.pdf}
-\caption{CDF showing drift in annual degree (interaction with their
-  network) of RFC authors over the period 2000-2020.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/emails/rfc_authors_degree.pdf}
+  \caption{
+    CDF showing drift in annual degree (interaction with their network)
+    of RFC authors over the period 2000-2020.
+  }
   \label{fig:degree_dist_rfc_authors}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/junior_senior_authors_seniormem_indegree.pdf}
-\caption{CDF showing number of senior contributors sending messages
-  (in-degree) to junior and senior authors.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/emails/junior_senior_authors_seniormem_indegree.pdf}
+  \caption{
+    CDF showing number of senior contributors sending messages (in-degree)
+    to junior and senior authors.
+  }
   \label{fig:junior_senior_indegree_seniormem}
@@
 % ICWSM 2022 paper Section 3
-\subsubsection{The IETF as a Social Graph}
+\subsubsection{Measuring Influence: The IETF as a Social Graph}
 
@@
 \begin{figure}
-\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/degree_CDF.pdf}
-\caption{Cumulative degree distribution of the email graph for different year-periods.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/degree_CDF.pdf}
+  \caption{
+    Cumulative degree distribution of the email graph for different
+    year-periods.
+  }
   \label{fig:degree_cdf}
@@
 \begin{figure}
-\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lcc_yearly_no_components.pdf}
-\caption{Size of Largest Connected Component (LCC) and Number of Connected Components (NCC)}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/lcc_yearly_no_components.pdf}
+  \caption{
+    Size of Largest Connected Component (LCC) and Number of Connected
+    Components (NCC)
+  }
   \label{fig:lcc_size_yearly}
@@
 \begin{figure}
-\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lcc_proportion_granular_yearly.pdf}
-\caption{Impact of removing participants by their influence on the size of the LCC.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/lcc_proportion_granular_yearly.pdf}
+  \caption{
+    Impact of removing participants by their influence on the size
+    of the LCC.
+  }
   \label{fig:purge_nodes_largest_connected_component_number_components}
@@
 \begin{figure}
-\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lineplot_proportion_emailscount_top_percentile.pdf}
-\caption{Proportion of emails sent (\%) by participants in each period 
-(with 95\% confidence interval) according to their betweenness-centrality percentile (x-axis).
-Y2-axis (purple) shows yearly count of emails.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/lineplot_proportion_emailscount_top_percentile.pdf}
+  \caption{
+    Proportion of emails sent (\%) by participants in each period (with
+    95\% confidence interval) according to their betweenness-centrality
+    percentile (x-axis).  Y2-axis (purple) shows yearly count of emails.
+  }
   \label{fig:emailcount_top_percentile}
@@
 \begin{figure}
-\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lineplot_avg_areas_top_percentile.pdf}
-\caption{Mean number of areas participated in (with 95\% confidence
-  interval) according to their betweenness-centrality percentile in the
-  x-axis, ranked from  top (0-10) to bottom percentile (91-100).}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/lineplot_avg_areas_top_percentile.pdf}
+  \caption{
+    Mean number of areas participated in (with 95\% confidence interval)
+    according to their betweenness-centrality percentile in the x-axis,
+    ranked from top (0-10) to bottom percentile (91-100).
+  }
   \label{fig:areas_participation_top_percentile}
@@
 \begin{figure}
-\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/lineplot_age_top_percentile.pdf}
-\caption{Mean participation duration (with 95\% confidence interval)
-  according to their betweenness-centrality percentile in the x-axis,
-  ranked from  top (0-10) to bottom percentile (91-100).}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/lineplot_age_top_percentile.pdf}
+  \caption{
+    Mean participation duration (with 95\% confidence interval) according
+    to their betweenness-centrality percentile in the x-axis, ranked from
+    top (0-10) to bottom percentile (91-100).
+  }
   \label{fig:age_top_percentile}
@@
 \begin{figure}
-\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/heatmap_top10_percent_overlap.pdf}
-\caption{Yearly overlap (\%) of top 10\% influencers.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/heatmap_top10_percent_overlap.pdf}
+  \caption{
+    Yearly overlap (\%) of top 10\% influencers.
+  }
   \label{fig:heatmap_top10_overlap}
@@
 \begin{figure}
-\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/new_influencers_top10_age_longevityInTop10.pdf}
-\caption{Percentage of new influential participants breaking into top 10\%
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/new_influencers_top10_age_longevityInTop10.pdf}
+  \caption{
+    Percentage of new influential participants breaking into top 10\%
     influencers (black); the average years of participation before entering
     top 10\% (purple); and minimum \#  years new influencers remain in top
-  10\% (with 95\% confidence intervals).}
+    10\% (with 95\% confidence intervals).
+  }
   \label{fig:rate_new_influential_participants}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/icwsm-2022/topic-div.png}
-\caption{Mean topic entropy (with 95\% confidence interval) according to
-  their betweenness-centrality percentile in the x-axis, ranked from  top
-  (0-10) to bottom percentile (91-100).}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/topic-div.png}
+  \caption{
+    Mean topic entropy (with 95\% confidence interval) according to their
+    betweenness-centrality percentile in the x-axis, ranked from top (0-10)
+    to bottom percentile (91-100).
+  }
   \label{fig:topic_diversity}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/icwsm-2022/documents_proportion_published_top_percentile.pdf}
-\caption{Average drafts authored (with 95\% confidence interval) according
-  to their betweenness-centrality percentile in the x-axis, ranked from
-  top (0-10) to bottom percentile (91-100).  Right-y axis (purple) shows
-  yearly drafts per year.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/documents_proportion_published_top_percentile.pdf}
+  \caption{
+    Average drafts authored (with 95\% confidence interval) according to
+    their betweenness-centrality percentile in the x-axis, ranked from top
+    (0-10) to bottom percentile (91-100).  Right-y axis (purple) shows
+    yearly drafts per year.
+  }
   \label{fig:drafts_initiation_top_percentile}
@@
   \end{tabular}
-\caption{Overlap in the co-author and email graph.}
+  \caption{
+    Overlap in the co-author and email graph.
+  }
   \label{tbl:correlation_sprmn}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/lineplot_non_overlapping_authors_emailnetwork_percentile.pdf}
-\caption{Percentage of Top 20\% authors that are not 20\% influencers
-  according to their betweenness-centrality (email graph) percentile in the
-  x-axis, ranked from  top (0-10) to bottom percentile (91-100).}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/lineplot_non_overlapping_authors_emailnetwork_percentile.pdf}
+  \caption{
+    Percentage of Top 20\% authors that are not 20\% influencers according
+    to their betweenness-centrality (email graph) percentile in the x-axis,
+    ranked from  top (0-10) to bottom percentile (91-100).
+  }
   \label{fig:non_overlapping_coauthornetwork_authors_emailnetwork}
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{figures-prev/icwsm-2022/WG_chair_percentile_CDF_before_after.pdf}
- \caption{CDF of the percentile of betweenness-centrality of WG chairs in
-  the email and co-authorship graph, one year before and after becoming
-  chairs.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/WG_chair_percentile_CDF_before_after.pdf}
+  \caption{
+    CDF of the percentile of betweenness-centrality of WG chairs in the
+    email and co-authorship graph, one year before and after becoming
+    chairs.
+  }
   \label{fig:wg_chair_before_after}
@@
 \begin{figure}
-\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/2019_affiliations_in_otheryears_coauthor_network.pdf}
-\caption{Authors in the most common affiliations of top 20\% authors
-  (co-authorship graph) in 2015-2019.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/2019_affiliations_in_otheryears_coauthor_network.pdf}
+  \caption{
+    Authors in the most common affiliations of top 20\% authors
+    (co-authorship graph) in 2015-2019.
+  }
   \label{fig:coauthor_affiliation_network2019_otheryears}
@@
 \begin{figure}
-\includegraphics[width=0.47\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/heatmap_2019_top10_affiliation_collaboration.pdf}
-\caption{Number of drafts co-authored by top 15 most frequently affiliated
-  organisations. We only include the top 20\% authors (co-authorship graph)
-  of 2015-2019.}
+  \centering
+  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/heatmap_2019_top10_affiliation_collaboration.pdf}
+  \caption{
+    Number of drafts co-authored by top 15 most frequently affiliated
+    organisations. We only include the top 20\% authors (co-authorship
+    graph) of 2015-2019.
+  }
   \label{fig:heatmap_2019_coauthordraft_most_frequent_org}

### 2024-09-22 paper.tex
Use consistent figure width

@@
 
+\newlength{\figureWidthOneColumn}
+\setlength{\figureWidthOneColumn}{0.40\textwidth}
+
 \newcommand{\todo}[1]{\textbf{\textcolor{red}{To do: #1}}}
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/rfcs_areas.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/documents/rfcs_areas.pdf}
   \caption{
@@
 
-\todo{Add draft publication rate as mentioned in Section \ref{sec:org-dynamics:participants}}
+\todo{Plot RFC publication count as a cumulative distribution}
+
+\todo{Add draft publication count as mentioned in Section \ref{sec:org-dynamics:participants}}
 
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/unique_wgs_per_year_areas.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/documents/unique_wgs_per_year_areas.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/day_counts_yearly.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/documents/day_counts_yearly.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/draft_counts_yearly.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/documents/draft_counts_yearly.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/page_counts_yearly_dt.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/documents/page_counts_yearly_dt.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/update_obsolete_yearly_pct.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/documents/update_obsolete_yearly_pct.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/cite_counts_yearly.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/documents/cite_counts_yearly.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/documents/keyword_usage_rate.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/documents/keyword_usage_rate.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/tma-2023/errata-by-year.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/errata-by-year.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/tma-2023/errata-submission-dates-area.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/errata-submission-dates-area.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/tma-2023/errata-by-status.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/errata-by-status.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/tma-2023/errata-by-rfc-by-area.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/errata-by-rfc-by-area.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/tma-2023/location-percent.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/location-percent.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/tma-2023/section-title-counts.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/section-title-counts.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/authors/top5_countries_normalised.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/authors/top5_countries_normalised.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/authors/continents_normalised.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/authors/continents_normalised.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/authors/top5_affiliations_normalised.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/authors/top5_affiliations_normalised.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/authors/top5_affiliations_normalised_acad.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/authors/top5_affiliations_normalised_acad.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/authors/repeat_authors.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/authors/repeat_authors.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/emails/pid_count_emailing_yearly.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/emails/pid_count_emailing_yearly.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/emails/frequency_emails_yearly_categories2.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/emails/frequency_emails_yearly_categories2.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/emails/yearly-draft-mention-volume.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/emails/yearly-draft-mention-volume.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/emails/age_authors_RFCs.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/emails/age_authors_RFCs.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/emails/rfc_authors_degree.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/emails/rfc_authors_degree.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/imc-2021/emails/junior_senior_authors_seniormem_indegree.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/emails/junior_senior_authors_seniormem_indegree.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/degree_CDF.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/degree_CDF.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/lcc_yearly_no_components.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/lcc_yearly_no_components.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/lcc_proportion_granular_yearly.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/lcc_proportion_granular_yearly.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/lineplot_proportion_emailscount_top_percentile.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/lineplot_proportion_emailscount_top_percentile.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/lineplot_avg_areas_top_percentile.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/lineplot_avg_areas_top_percentile.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/lineplot_age_top_percentile.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/lineplot_age_top_percentile.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/heatmap_top10_percent_overlap.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/heatmap_top10_percent_overlap.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/new_influencers_top10_age_longevityInTop10.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/new_influencers_top10_age_longevityInTop10.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/topic-div.png}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/topic-div.png}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/documents_proportion_published_top_percentile.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/documents_proportion_published_top_percentile.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/lineplot_non_overlapping_authors_emailnetwork_percentile.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/coauthor_affiliation_network/lineplot_non_overlapping_authors_emailnetwork_percentile.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/WG_chair_percentile_CDF_before_after.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/WG_chair_percentile_CDF_before_after.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/2019_affiliations_in_otheryears_coauthor_network.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/coauthor_affiliation_network/2019_affiliations_in_otheryears_coauthor_network.pdf}
   \caption{
@@
   \centering
-  \includegraphics[width=0.40\textwidth]{figures-prev/icwsm-2022/coauthor_affiliation_network/heatmap_2019_top10_affiliation_collaboration.pdf}
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/coauthor_affiliation_network/heatmap_2019_top10_affiliation_collaboration.pdf}
   \caption{

### 2024-09-22 paper.tex
Tweaks

@@
 % \end{figure}
+
+\todo{Add introduction to trends in RFC publication; research questions}
 
@@
 
+%--------------------------------------------------------------------------------------------------
+\subsection{Summary}
+
+\todo{summarise results about RFC publication trends}
+
 %==================================================================================================
@@
 
-%..................................................................................................
-\pb{Geographic Distribution of Authors:}
+%--------------------------------------------------------------------------------------------------
+\subsection{Geographic Distribution of Authors}
 Figures~\ref{fig:author_countries_normalised} and
@@
 
-%..................................................................................................
-\pb{Author Affiliations:}
+%--------------------------------------------------------------------------------------------------
+\subsection{Author Affiliations}
 Figure~\ref{fig:author_affiliations_normalised} shows the top 10
@@
 
-%..................................................................................................
-\pb{Summary:}
+%--------------------------------------------------------------------------------------------------
+\subsection{Summary}
 The trends highlighted in this section  indicate a pool of authors that is
@@
 
+\todo{Add introduction to organisational dynamics section}
+
 %--------------------------------------------------------------------------------------------------
-\subsection{Interactions}
+\subsection{Characterising Participant Interactions}
 \label{sec:org-dynamics:participants}
@@
 %--------------------------------------------------------------------------------------------------
-\subsection{Influence}
+\subsection{Characterising Influence and Impact}
 \label{sec:org-dynamics:influence}
@@
 
-% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-\subsubsection{Discussion}
+%--------------------------------------------------------------------------------------------------
+\subsubsection{Summary}
 

### 2024-09-22 paper.tex
Add ICWSM 2024 paper research questions

@@
         Prashant Khare \and
+        Matthew Russell Barnes \and
         Waleed Iqbal \and
@@
         Stephen McQuistin \and
+        Richard G. Clegg \and
         Patrick Healey \and
@@
 standardisation, but also its resilience to the loss of major participants.
-In the following we consider the following \emph{research questions}: 
+In the following we consider the following research questions: 
 \begin{enumerate}
@@
 
-%--------------------------------------------------------------------------------------------------
+% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 \subsubsection{Summary}
@@
 %--------------------------------------------------------------------------------------------------
-\subsection{Hierarchy}
+\subsection{Hierarchy and Communication Dynamics}
+\label{sec:org-dynamics:hierarchy}
 
@@
 %   activity over time? How does this vary across hierarchy level?
+
+In the following, we quantitatively determine the effect that organisational
+hierarchy levels have on communication patterns throughout the IETF. 
+% We model communication as a network where a node represents an individual
+% and a directed edge represents a communication sent from one individual to
+% another~\cite{panzarasa2009patterns,viswanath2009evolution,klimt2004enron}.
+% By repeating the analysis across time, we can build a full temporal
+% network~\cite{masuda2016guide} of communication interactions between people
+% in the system.In this study we consider only the network structure and the
+% status at that time of the individuals, not the content of messages. 
+We study how the email communication is impacted by the hierarchical roles
+of individuals. According to the IETF mission statement~\cite{rfc3935}, the
+IETF is made up of volunteers who collaborate to develop consensus-based
+technical standards. This collaboration should be evident in their
+communication practises.
+
+%..................................................................................................
+% ICWSM 2024 paper section 2
+
+In the following we consider the following research questions: 
+\begin{enumerate}
+  \item
+    Are higher levels of the hierarchy growing or becoming more
+    centralised? Are they associated with an increased domination
+    of the conversation?
+    (\S\ref{sec:org-dynamics:hierarchy:rq1})
+
+  \item
+    What is the association between the organisation hierarchy and general
+    communication patterns?
+    (\S\ref{sec:org-dynamics:hierarchy:rq2})
+
+  \item
+    How do people with differing roles communicate with each other?  Does
+    information flow ``up" or ``down" the hierarchy?
+    (\S\ref{sec:org-dynamics:hierarchy:rq3})
+
+  \item
+    What is the impact that individuals have on their direct contacts'
+    activity over time? How does this vary across hierarchy level?
+    (\S\ref{sec:org-dynamics:hierarchy:rq4})
+
+\end{enumerate}
+
+The first hypothesis focuses on the impact of the
+``steepness"~\cite{anderson2010functions} of the hierarchy. This refers to
+the fewer people in higher hierarchy levels the more steep the hierarchy
+is, i.e a centralised control of decision making in higher levels. The
+opposite of this is a ``diffused"~\cite{hussain2019voice} hierarchy where
+the large amount of people in higher levels stifle lower level voice by
+creating ``bystanders" who find it hard to speak up. In general, the
+conclusions about steep hierarchies are confused, where some papers
+conclude a positive and some a negative effect of steep hierarchies. We
+explore what this means for the IETF in RQ1.
+
+The second hypothesis regards the ``power
+distance"~\cite{li2021does,duan2018authoritarian,guo2020inclusive} of
+individuals, which is the number of hierarchy levels between people who are
+communicating. The focus is on the impact this has on people's voice or
+lack thereof when the distance is large versus small. The conclusions from
+this research area usually suggest that even small power distances have a
+large effect of reducing the voice of the lower of the two levels. We
+tackle this for the IETF in three different ways outlined in RQ2-4.
+
+The IETF being a voluntary organisation gives a new perspective on these
+hypotheses, as the existing literature focuses on commercial organisations.
+In contrast to the above conclusions, we demonstrate that despite the
+hierarchy becoming more ``diffused", participants in the higher levels in
+the IETF hierarchy perform a ``facilitator" role, promoting discussion with
+those they interact with and receiving a benefit themselves as a result. We
+also show that higher level participants tend to be a focus of discussion
+with remarks directed ``upward" toward them whereas Regular Participants
+(RPs) engage more in group-discussion with less focus on an individual.
+This suggests that the IETF does not conform to the notion that power
+distance has a negative effect on communication, nor that a more diffused
+hierarchy reduces the voice of lower levels.
+
+% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+\subsubsection{Methodology}
+\label{sec:org-dynamics:hierarchy:methodology}
+
+% From our ICWSM 2024 paper, section 4:
+
+% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+\subsubsection{Centralisation and Dominance}
+\label{sec:org-dynamics:hierarchy:rq1}
+
+%..................................................................................................
+% From our ICWSM 2024 paper, Section 2:
+
+\pb{RQ1: are higher levels of the hierarchy growing or becoming more
+centralised? are they associated with an increased domination of the
+conversation?}
+
+The effect the steepness of an organisation's hierarchy has on the
+communication within is not well understood. This RQ is important because
+once we know whether the IETF hierarchy is becoming steeper, we can look at
+the effects this is having on the communication patterns of participants.
+
+In this RQ, we look at the number of individuals and/or roles that inhabit
+each hierarchy level. We also consider whether individuals in higher levels
+are more active in the conversation and quantify the evolution of the
+mailing list activity within each level of the IETF hierarchy. We find that
+the middle level, Working Group Chairs (WGCs), is gaining in overall
+proportion of activity, whilst the lowest level, RPs, is decreasing. The
+top level, Area Directors (ADs), remains constant in proportion. The
+increase in WGC activity coincides with an increase of WGCs per Working
+Groups (WGs).
+
+%..................................................................................................
+% From our ICWSM 2024 paper, Section 5.1:
+
+% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+\subsubsection{Organisational Hierarchy and Communication Patterns}
+\label{sec:org-dynamics:hierarchy:rq2}
+
+%..................................................................................................
+% From our ICWSM 2024 paper, Section 2:
+
+\pb{RQ2: What is the association between the organisation hierarchy and
+general communication patterns?}
+
+The communication patterns that different levels in the hierarchy
+experience may be  indicative of the cooperation between
+levels~\cite{li2021does,duan2018authoritarian}. For instance, if
+participants at higher levels send lots of email while receiving little,
+this suggests a lack of receptiveness of higher levels to the suggestions
+of lower levels, and low confidence of lower levels to voice their
+opinions. In contrast, we hypothesise that collaboration between layers is
+high, as outlined in the IETF mission statement, which will look more like
+the opposite of this example.
+
+In this vein, we compare the ways that different roles communicate, looking
+at the tendency for communications to be purely inbound, outbound or use
+more complex patterns. We count three edge motifs for different hierarchy
+levels as described in section~\ref{sec:motif}. We find that higher levels
+consistently experience a higher proportion of incoming communication than
+RPs, and all levels send outward communication at similar proportions over
+time. Also, the proportion of people who begin email threads are split by
+hierarchy level. Higher levels show a disproportionate tendency to
+originate email threads.
+
+%..................................................................................................
+% From our ICWSM 2024 paper, Section 5.2:
+
+% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+\subsubsection{Communication up and down the Hierarchy}
+\label{sec:org-dynamics:hierarchy:rq3}
+
+%..................................................................................................
+% From our ICWSM 2024 paper, Section 2:
+
+\pb{RQ3: How do people with differing roles communicate with each other?
+Does information flow ``up" or ``down" the hierarchy?}
+
+Our third hypothesis is that the cooperative nature of the IETF will be
+evident in the effect that power distance~\cite{li2021does} has on
+communication throughout the organisation. The raw proportion of the
+communication between hierarchy levels helps us to understand the
+confidence of lower levels to exercise their voice. This proportion, for a
+cooperative and voluntary organisation should be higher for lower hierarchy
+levels, showing a confidence in their voice.
+
+We categorise edges based on the hierarchy levels they originate from and
+are sent to. Ratios are then calculated of ``upward" versus ``downward"
+communication between RPs, WGCs and ADs. We find that the lower the level,
+the more of a skew to upward communication.  Communication tends to flow
+``up'' the hierarchy more than ``down''.
+
+%..................................................................................................
+% From our ICWSM 2024 paper, Section 5.3:
+
+% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+\subsubsection{Mobility Within the Hierarchy}
+\label{sec:org-dynamics:hierarchy:rq4}
+
+%..................................................................................................
+% From our ICWSM 2024 paper, Section 2:
+
+\pb{RQ4: What is the impact that individuals have on their direct contacts'
+activity over time? How does this vary across hierarchy level?}
+
+An important question~\cite{li2021does,guo2020inclusive} for any
+organisation is whether higher level participants encourage those at lower
+levels in the hierarchy to properly engage. The IETF is cooperative and
+voluntary by design, therefore we hypothesise a high level of encouragement
+from higher levels to lower. We measure whether activity in one time period
+is associated with activity in a subsequent time period both for
+individuals and for neighbourhoods. If those who communicate with higher
+levels receive a boost in their communication as a result, this is an
+indication of such encouragement.
+
+In particular, we perform Mobility Taxonomy analysis to determine the
+effect individual IETF participants have on their future number of
+connections, and the average number of connections of their neighbours.  We
+find that WGCs have an increased tendency, versus RPs, to remain active
+(\textit{Mobility}), an increased tendency for their neighbours to be
+active subsequent to high WGC activity (\textit{Philanthropy}) and for them
+to gain activity after their neighbours are active (\textit{Community}).
+We interpret this as WGCs filling a ``facilitator" role on the mailing
+lists.
+
+%..................................................................................................
+% From our ICWSM 2024 paper, Section 5.4:
 

### 2024-09-22 paper.tex
Add content from ICWSM 2024 paper, section 4

@@
 \newlength{\figureWidthOneColumn}
-\setlength{\figureWidthOneColumn}{0.40\textwidth}
+\setlength{\figureWidthOneColumn}{0.4\textwidth}
+
+\newlength{\figureWidthTwoColumn}
+\setlength{\figureWidthTwoColumn}{0.8\textwidth}
 
@@
 
+For \textbf{RQ1}, to determine the steepness or diffusion of the hierarchy,
+the proportion of nodes and activity are determined in each hierarchy
+level. Figure~\ref{fig:activity_proportions}a shows the number of active
+nodes in each level. The bottom level, RPs, are the largest contingent
+consisting of 90--86\% of active participants; this proportion is
+decreasing over time. The next level is WGCs which manage the WGs and are
+6--10\% of active participants; this proportion is rising over time. The
+top level is ADs which manage many WGs within an area and consist of less
+than 1\% of the total active population of the IETF; this stays constant
+over time. Therefore, the WGC level is becoming more diffused over time as
+there are about 15 RPs for every one WGC in the organisation in 2013 versus
+8 RPs per WGC in 2021.
+
+The proportion of activity of nodes in the last year is also plotted and
+split into hierarchy level in Figure~\ref{fig:activity_proportions}b. It is
+clear that individuals at higher levels contribute a disproportionate share
+to communication and this share is increasing over time, which is
+consistent with findings in \cite{khare2022web}. Therefore, less of a
+proportion of communication is flowing from the RPs at the same time as the
+number of WGC roles is increasing. This further suggests that the IETF
+hierarchy is becoming more diffused, in terms of mailing list
+communication, which may make cooperation between hierarchy levels more
+difficult.
+
+\begin{figure}[t]
+  \centering
+  \small{a)} \\
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/active_node_proportions_year_sliding_window.png} \\
+  \small{b)} \\
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/activity_proportions_year_sliding_window.png}
+  \caption{
+    Proportions of RPs, WGCs and ADs by a) number of active individuals and
+    b) activity (number of emails sent to mailing lists). 
+  }
+  \label{fig:activity_proportions}
+\end{figure}
+
+Figure~\ref{fig:wgc_over_time}a shows the amount of nodes that inhabit the
+WGC hierarchy level over our time period. There is a $35\%$ rise in the
+amount of WGC roles whereas the amount of individuals that fill the WGC
+roles remains mostly constant.
+
+Figure~\ref{fig:wgc_over_time}b contains two estimates of the number of WGs
+over time. The estimated number of WGs is likely the most accurate in the
+middle 50\% of both plotted lines. Their discrepancy is the least (at most
+10 WGs) from late 2014 to early 2019. The start of the COVID-19 pandemic
+may also affect mailing list activity in 2020, and similar trends are
+visible in the data in \cite{mcquistin2021characterising}. A full
+explanation on how the data is estimated can be found in
+section~\ref{sec:WG_activity}.
+
+Figure~\ref{fig:wgc_over_time}c shows how many WGCs exist per WG, using
+both estimates of the WG amount from Figure~\ref{fig:wgc_over_time}b. It is
+clear there is a slow a rise in WGCs per WG over time. The ratio rises from
+between 1.7-1.9 to 2.0-2.2 within the middle 50\%, late 2014 to early
+2019.This perhaps represents a growing understanding in the IETF that it is
+desirable to have two or more chairs per WG in case of
+illness/unavailability or to avoid conflicts of interest. Therefore, this
+plot again shows that the hierarchy is becoming more diffused, even if the
+raw amount of individuals in the WGC roles has remained constant. The
+multiple roles per individual may result in even more difficulty of RPs to
+use their voice.
+
+An AD version of Figure~\ref{fig:wgc_over_time} is available, but it is
+omitted to save space as all three plots stay mostly constant throughout
+the period. There are about 15 individual ADs split amongst 15 roles in 7-8
+areas with about 2 ADs per Area. The set of IETF Areas is broadly fixed,
+while WGs are created and closed relatively frequently.
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/over_time.png}
+  \caption{
+    Overview of WGC numbers and WG numbers over time. a) shows the number
+    of WGC roles (some individuals hold multiple WGC roles) and the number
+    of individuals who are WGCs b) shows two different estimates of the
+    number of WGs (see section~\ref{sec:method}) and c) shows the mean
+    number of WGCs per WG. Notice the non-zero-based y-axis exaggerates the
+    variation in these figures.
+  }
+  \label{fig:wgc_over_time}
+\end{figure}
+
+An AD version of Figure~\ref{fig:wgc_over_time} is available, but it is
+omitted to save space as all three plots stay mostly constant throughout
+the period. There are about 15 individual ADs split amongst 15 roles in 7-8
+areas with about 2 ADs per Area. The set of IETF Areas is broadly fixed,
+while WGs are created and closed relatively frequently.
+
+Table~\ref{tab:WGC_email} shows the emails sent and received for WGC and
+ADs on mailing lists in the year before they first take that role, and the
+year after. WGCs and ADs are both active in using the mailing lists, ADs
+much more so. However, they don't become more active to any major degree
+after taking the roles. This may suggest that the individuals who take on
+higher hierarchy level roles are already contributing at a higher level.
+
+\begin{table}[!tbp]
+\centering
+    \begin{tabular}{c||c|c}
+        Status & Sent & Received \\
+         \hline \hline 
+        Before WGC & $67.26\pm99.80$ & $70.11\pm103.37$ \\
+        \hline
+        After WGC & $77.20\pm102.85$ & $85.90\pm107.38$ \\
+        \hline
+        Before AD & $218.84\pm226.92$ & $230.50\pm214.75$ \\
+        \hline
+        After AD & $209.21\pm186.46$ & $219.16\pm189.85$ \\
+    \end{tabular}
+        \caption{Mean number of emails received for one year before/after becoming a WGC (AD) for the first time. The error is one standard deviation.}
+    \label{tab:WGC_email}
+\end{table}
+
+\pb{Summary}
+In answer to RQ1 the picture is mixed, the number of individuals with WGC
+roles has remained broadly constant but the number of WGC roles has
+increased, indicating that individuals willing to become WGC have taken on
+more such roles. The number of the higher level AD roles has, by design,
+remained broadly constant over the time studied. The proportion of active
+individuals with WGC roles has grown from 6\% to 10\% and the proportion of
+the conversation taken by those roles has grown from 25\% to 35\% over the
+period studied. Therefore the hierarchy has become more diffused, which may
+mean RPs experience more difficulty in expressing their voice.
+
 % - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
@@
 
+To quantify the effect of being in each hierarchy level has on network wide
+communication for \textbf{RQ2}, we calculate the proportion of three node
+motifs, lasting at most a month, in which each node participates over a
+year period, see section~\ref{sec:motif}. These motifs are counted,
+categorised into the three hierarchy levels and then proportions are taken
+of each motif type.
+
+Figure~\ref{fig:motif_proportions} shows the proportions of each level for
+three node ``Inward Star", ``Outward Star", ``Mixed Star", and
+``Triangles". The Inward and Mixed Star motifs show about a $4\%$ and
+$10\%$ increase in proportion for higher levels in the hierarchy versus for
+RPs. However, about a $10\%$ increase for RPs is seen for Triangles versus
+WGCs and ADs, and Outward Star proportions remain similar for all levels of
+the hierarchy.
+
+The larger proportion for WGCs and ADs of Inward and Mixed Star motifs
+whilst the Outward Star motifs remain similar for all levels suggests that
+higher levels receive more direct communication. The higher proportion of
+RP Triangle motifs suggests their discussion is more of a group activity,
+whereas WGCs and ADs have a larger proportion of one-on-one conversations.
+Therefore, we interpret the general communication patterns within the IETF
+as a discussion amongst RPs interspersed with questions sent to and
+announcements from WGCs and ADs. This indicates some collaboration between
+layers and confidence of some RPs' to voice their opinions upwards.
+
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=\figureWidthTwoColumn]{figures-prev/icwsm-2024/motif_strata_comparison.png}
+  \caption{
+    Communication patterns analysed by three edge temporal motifs for RPs,
+  }
+  \label{fig:motif_proportions}
+\end{figure*}
+
+Figure~\ref{fig:origin_emails} seems to corroborate this. We calculate the
+proportion of mailing list threads which nodes in each hierarchy level
+originate, using a year window, pushed forward by a month each calculation.
+We see that WGCs and ADs send a disproportionate number of originating
+emails to the mailing lists. WGCs send 30--50\% whereas they are 6--10\% of
+active individuals, and ADs 5--10\% versus 0.5\%.
+
+Therefore, the larger proportion of inward motifs for higher levels is in
+part due to WGCs and ADs disproportionately originating email threads.
+Other IETF participants will then reply to the thread, boosting the
+proportion of inward motifs WGCs and ADs appear in. This may suggest the
+higher levels are good ``condensers of communication" as they receive more
+than they send.
+
+In answer to the RQ, combining the interpretations of the two plots, the
+discussion seems to happen in the following way: The ADs or WGCs are
+commonly the originators of discussion threads. Where conversation includes
+a WGC or AD the conversation tends to be directed around them and usually
+inward towards the WGC or AD. The RPs are more likely to engage in
+discussions amongst themselves (triangular communication patterns) when
+higher level individuals are not present.
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/Origin_email_proportions.png}
+  \caption{
+    Proportion of ``origin" (non reply) emails sent by RPs, WGCs and ADs,
+    this should be interpreted in conjunction with
+    figure~\ref{fig:activity_proportions} showing the proportion of active
+    individuals in each class.
+  }
+  \label{fig:origin_emails}
+\end{figure}
+
 % - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
@@
 % From our ICWSM 2024 paper, Section 5.3:
+
+For \textbf{RQ3}, we determine the direction communication tends to flow
+through the hierarchy by looking at inbound and outbound communication
+flows. The network's edges are categorised based on their source and target
+node's hierarchy level. For each of the inter-level communication
+combinations the proportion of communication in the last year going in the
+``upward" hierarchy direction is calculated. For instance, if there are six
+emails from RPs to WGCs and twelve from WGCs to RPs then the proportion
+upward is $\frac{6}{6+12}=\frac{1}{3}$.
+
+In Figure~\ref{fig:inter_strata}, these proportions are plotted, and the
+year long window is again pushed forward by one month to increase the
+resolution of changes. A proportion of higher than $0.5$ shows more
+communication flowing up the hierarchy than down. All three levels are
+mostly upward in direction, with the lowest two levels (RP$\rightarrow$
+WGC) having the largest skew in the proportion. The other two lines
+(RP$\rightarrow$AD and WGC$\rightarrow$AD) show similar periodicity, with
+the WGC$\rightarrow$AD line dropping as far below $0.5$ as above showing a
+large shift in inter-level communication patterns higher up the hierarchy.
+
+The direct measurement of message flow between the hierarchy levels seems
+to bolster our interpretation that the cooperative design of the IETF is
+working. This analysis shows that RPs are communicating much more
+preferentially upwards in the hierarchy, despite the decrease in overall
+share of RP activity. The increase within the WGC level does not have any
+noticeable effect on their level of communication, if anything there is a
+negative relationship. This again suggests both WGCs and ADs perform a kind
+of ``condenser of communication" or ``facilitator" role. In other words,
+this analysis suggests the IETF has a ``bottom-up" communication style,
+where the higher levels encourage the lower levels' voice.  This is the
+desired pattern of communication for a volunteer organisation that is
+looking to grow and encourage participants to engage and gradually take on
+leadership roles.
+
+Overall for RQ3 then, the pattern observed was communication flowing up the
+hierarchy and this corresponds to the answer found in RQ2. WGCs and ADs
+were more likely to receive communication with individuals than send
+communication to individuals whereas both were more likely than RPs to send
+out messages to the list in general.
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/up_hierarchy_communication_ratios.png}
+  \caption{
+    Proportion of communications that are ``up" the hierarchy for different
+    groups. A proportion above 0.5 indicates that the majority of
+    communications are ``upward".
+  }
+  \label{fig:inter_strata}
+\end{figure}
 
@@
 % From our ICWSM 2024 paper, Section 5.4:
+
+Finally, to determine how individuals and their neighbours affect each
+other's activity levels in subsequent time periods, for \textbf{RQ4}, we
+take correlations of degree between different time windows using the
+Mobility Taxonomy~\cite{barnes2023measuring} aspects called
+\textit{Mobility}, \textit{Neighbour Mobility}, \textit{Philanthropy} and
+\textit{Community} explained in the previous section. In
+Figure~\ref{fig:mob_tax}, the time window chosen is one year which is split
+into two snapshots graphs of six months, the time window then moves forward
+one month and the process is repeated. The mid point of the year time
+window is plotted. WGCs and RPs are plotted but the small number of ADs
+mean that the correlations are too noisy to show meaningful effects.
+
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=\figureWidthTwoColumn]{figures-prev/icwsm-2024/hierarchy_status_comparison.png}
+  \caption{
+    Connection between roles and activity for RP and WGC viewed in terms of
+    a) Mobility b) Neighbour Mobility c) Philanthropy and d) Community over
+    time. See section~\ref{subsec:mob_tax} for definitions of these terms.
+  }
+  \label{fig:mob_tax}
+\end{figure*}
+
+The \textit{Mobility} of both RPs and WGCs is plotted in
+Figure~\ref{fig:mob_tax}a. The high amount of correlation between the
+degree of an individual in the first half of the time window with the
+second half shows that all users have a tendency to become more active in
+subsequent time periods if they are highly active in the previous period,
+but this is more pronounced in WGCs. This is found to follow a period of
+about one year for both RPs and WGC, suggesting that a person is likely to
+be as active as there were a year ago. The \textit{Neighbour Mobility}
+plot, Figure~\ref{fig:mob_tax}b, shows a similar tendency for
+neighbourhoods, but with less of a split.
+
+The \textit{Philanthropy} plot, Figure~\ref{fig:mob_tax}c, shows a higher
+correlation for WGCs than RPs between an individual's degree in the first
+snapshot with and their ND in the second. This means that individuals who
+directly interact with WGCs become more active on mailing lists
+subsequently. The \textit{Community} plot, Figure~\ref{fig:mob_tax}d, shows
+the correlation between the ND in the first snapshot with the degree in the
+second.  The higher correlation for WGCs suggests they benefit from
+interactions with highly active neighbours by themselves becoming more
+active in subsequent time periods. These trends for \textit{Philanthropy}
+and \textit{Community} show a reversal in early 2016 that coincides with a
+period where the mean number of WGCs per WG and the number of WGC roles was
+rapidly rising, see figure~\ref{fig:wgc_over_time}. A merging of the
+Applications area and Real-time Applications \& Infrastructure area in May
+2015 may be a cause of this reversal of correlation, as the assignment of
+WGs to areas was changed which may cause more non-reciprocal WGC
+communication than normal.
+
+Increased \textit{Mobility} for WGCs suggests that they gain advantage from
+their role with their activity leading to similar communication activity in
+the future. Moreover, they experience an increased effect of
+\textit{Philanthropy} and \textit{Community}. The former suggests that if a
+WGC is engaged in a higher amount of discussion, then in the future the
+individuals who they were communicating with will also be highly active.
+The latter suggests that when WGCs are surrounded by people who are active
+in discussion, they are likely to be more active themselves in the future
+too. All of this has less of an effect for RPs who show a lower correlation
+in terms of all three aspects. Our interpretation of this is that active
+WGCs are ``facilitating" discussion in the WG mailing lists; both having
+their activity boosted by their neighbours and, in turn, boosting the
+activities of their neighbours. This aligns well with the hypothesis that
+the cooperative and voluntary design of the IETF lends itself well to
+higher levels encouraging those lower to engage in discussion.
+
+Overall the answer to RQ4 is that the WGCs have a positive effect on their
+immediate contacts and those individuals that directly discuss with WGCs
+are more likely to engage more in discussion in subsequent time periods.
+Conversely though the WGCs who discuss topics with individuals who are
+active in discussion are, themselves, more likely to engage more fully in
+discussion. This points to something like a virtuous cycle of WGCs
+encouraging and being encouraged by their direct contacts.
+
+% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+\subsubsection{Summary}
+
+% From our ICWSM 2024 paper, Section 6:
+
+\todo{summarise hierarchy results}
 

### 2024-09-22 paper.tex
Tidy RQs for §5.3

@@
 
+% RQ1: Are higher levels of the hierarchy growing or becoming more
+% centralised? Are they associated with an increased domination of
+% the conversation?
+
 %..................................................................................................
 % From our ICWSM 2024 paper, Section 2:
-
-\pb{RQ1: are higher levels of the hierarchy growing or becoming more
-centralised? are they associated with an increased domination of the
-conversation?}
 
@@
 
+% RQ2: What is the association between the organisation hierarchy and
+% general communication patterns?
+
 %..................................................................................................
 % From our ICWSM 2024 paper, Section 2:
-
-\pb{RQ2: What is the association between the organisation hierarchy and
-general communication patterns?}
 
@@
 
+% RQ3: How do people with differing roles communicate with each other?
+% Does information flow ``up" or ``down" the hierarchy?
+
 %..................................................................................................
 % From our ICWSM 2024 paper, Section 2:
-
-\pb{RQ3: How do people with differing roles communicate with each other?
-Does information flow ``up" or ``down" the hierarchy?}
 
@@
 
+% RQ4: What is the impact that individuals have on their direct contacts'
+% activity over time? How does this vary across hierarchy level?
+
 %..................................................................................................
 % From our ICWSM 2024 paper, Section 2:
-
-\pb{RQ4: What is the impact that individuals have on their direct contacts'
-activity over time? How does this vary across hierarchy level?}
 

### 2024-09-22 paper.tex
Add acknowledgements
Cite our prior work

@@
 % in the area.
+
+This paper summarises and expands on our prior work
+\cite{mcquistin:2021:characterising,khare:2022:web-we-weave,
+healey:2023:power,mcquistin:2023:errare,khare:2023:tracing,
+karan:2023:leda,healey:2023:power-frontiers,barnes:2024:temporal}
 
@@
 
+This work has been supported in part by the UK Engineering and Physical
+Sciences Research Council under grants EP/S033564/1 and EP/S036075/1
+(``Streamlining Social Decision Making for Improved Internet Standards'').
+
 %==================================================================================================

### 2024-09-22 paper.tex
Reorganise later sections
Fix PDF metadata

@@
 
-%..................................................................................................
-% LEDA
-% - a taxonomy of dialogue acts (DAs) and a labeled dataset of emails
-
 %==================================================================================================
@@
 
+% LEDA
+% - a taxonomy of dialogue acts (DAs) and a labeled dataset of emails
+
 %==================================================================================================
@@
 
-\subsection{For authors}
+\subsection{What Makes a Successful Author?}
 % ICWSM 2022 paper Section 4
@@
 
-\subsection{For documents}
+\subsection{What Makes a Successful Document?}
 
@@
   \pdfinfo{
-    /Title        (...)
+    /Title        (An Empirical Analysis of the Internet Engineering Task Force with Computational Methods)
     /Author       (...)
-    /Subject      (...)
-    /Keywords     (..., ..., ...)
-    /CreationDate (D:20150827110616Z)
-    /ModDate      (D:20150827110616Z)
+    /Subject      (Internet Standards)
+    /Keywords     (IETF, Internet, Standardisation)
+    /CreationDate (D:20240922164700Z)
+    /ModDate      (D:20240922164700Z)
     /Creator      (LaTeX)

### 2024-09-27 paper.tex
Add Section 2.2 on Data Sources

@@
 
+% The following is adapted from our IMC 2021 paper and our IFIP TMA 2023
+% paper, Section II:
+
+We rely on a number of data sources to conduct our study.
+
+%..................................................................................................
+\pb{IETF Datatracker:}
+The Datatracker\footnote{\url{https://datatracker.ietf.org/}} is the main
+administrative database used by the IETF, IRTF, and IAB to coordinate their
+work. It contains information about contributors, working groups,
+documents, and meetings. The Datatracker provides comprehensive metadata
+about authors, and the evolution of drafts as they work their way through
+the standardisation process. The Datatracker was introduced in the early
+2000s and its functionality has gradually been extended over time. It
+contains limited historical data about RFCs produced before its creation,
+but the IETF has back-filled some data from other records.  We have
+extracted relevant metadata from the Datatracker, using its programmatic
+REST API, for all RFCs published since 2001. This gives us data pertaining
+to 4,512 authors, as well as richer metadata for 5,707 RFCs.
+
+%..................................................................................................
+\pb{RFC Editor:}
+The RFC Editor maintains an index of all RFC publications, alongside
+metadata related to each document, including its standardisation status,
+publication stream, authors, and errata. We gather all entries for RFCs
+published through the end of 2020, giving a total of 8,711 RFCs.
+
+RFCs are not changed after publication, but the RFC Editor maintains a
+public database of reported errata. The errata filing are separated into
+\emph{technical errata}, mistakes in the technical content of the RFC that
+are likely to result in incorrect, non-conforming, implementations of the
+standards, and \emph{editorial errata} including spelling and punctuation
+errors that do not otherwise impact the technical content.
+Errata filed against RFCs published on the IETF stream are checked for
+correctness by the RFC authors, working group chairs, and area directors.
+IRTF errata are checked by the authors and the Internet Research Steering
+Group. The IAB and Independent Submissions Editor check errata for their
+streams. We collected\footnote{RFC errata are available at
+\url{https://www.rfc-editor.org/errata.php}. We thank the RFC Editor for
+making the underlying database available to us in machine-readable form for
+this analysis.} the 6,759 errata reports submitted from January 2001 to the
+end of December 2022 for analysis.
+
+%..................................................................................................
+\pb{Email archives and entity resolution:}
+The IETF maintains reasonably complete email archives relating to working
+group discussions, meeting, and other activities.\footnote{\url{https://mailarchive.ietf.org/}
+with public IMAP access also available.} We gather all available messages
+contained within this archive, covering 2,439,240 messages, sent from
+74,646 unique email addresses, across 1,153 lists. This snapshot was taken
+on April 18th 2021.
+
+A key challenge is attributing each email to an individual contributor.
+Although each, naturally, contains a \texttt{From:} header, we find that some
+contributors use multiple addresses. Beyond this, it is necessary to map
+each email to the respective person in the Datatracker and RFC Editor
+datasets.  Thus, we perform entity resolution on email senders, and match
+email senders with their Datatracker profiles. We assign each sender a
+unique identifier, which is associated with a set of name and address
+variations from the Datatracker.
+
+Entity resolution takes place in multiple stages.  First, we check if the
+sender's email address has a Datatracker profile. If so, we associate all
+of their Datatracker metadata with that person's unique identifier, and
+label the message as having been sent by that identifier.  Next, if an
+email's sender does not appear in the Datatracker, we check, based on their
+name, if they have already been assigned a Datatracker record. If so, the
+message is labelled as having been sent by that identifier, and the set of
+names and addresses associated with that identifier is updated to include
+the email's sender name and address.
+
+These two stages -- matching with the Datatracker, and merging previously
+seen names and addresses -- accounts for the majority (60\%) of messages.
+If an email's sender is not in the Datatracker, and the name and address
+has not been previously seen, a new person identifier is generated. This
+accounts for a small portion (~10\%) of messages. This is reasonable, given
+that Datatracker profiles have become necessary for many of the IETF's
+day-to-day activities.
+
+As a final processing step, we label each identifier as either representing
+a \emph{contributor}, a \emph{role-based} address, or an \emph{automated}
+sender. A contributor refers to any standard user participating in the
+IETF; role-based identifiers reflect addresses that are used by the holders
+of a particular organisational role, such as the IETF chair; and automated
+identifiers are those addresses that are system-specific, such as GitHub or
+IETF notification and announcement addresses. Role-based and automated
+identifiers account for the remaining 30\% of messages.
+
+Note, within the mail archive, spam-indicating headers are present for most
+messages since 2009, and we confirm pre-filtering is performed by the IETF
+mail servers.  As an extra validation step, we ran a spam filter over all
+the archived messages and discarded those marked as spam. Both sources
+indicate there is very little spam (less than 1\%), so it should not have
+significant impact on our findings.
+
+% \pb{Manually labelled dataset:} 
+% We also use data collated by Nikkhah et al.~\cite{nikkhah2017statistical}.
+% This dataset labels the ``success'', alongside other features like scope
+% and value, of 251 RFCs.  155 of these also appear in our set of RFCs that
+% have Datatracker metadata available. We later use this to explore the
+% features that best predict the success of an RFC.
+
+% \pb{Microsoft Academic citation data:}
+% We use the Microsoft Academic Knowledge API to identify indexed academic
+% work that cites each RFC. We use the Microsoft Academic Graph instead of
+% other sources (\eg Google Scholar) as it provides time-stamped citations.
+
+%..................................................................................................
+\pb{Reproducibility and data access:}
+We have developed a library, \texttt{ietfdata}, that fetches the RFC
+Index, communicates with the IETF Datatracker API, and retrieves messages
+from the IETF IMAP mail archives. This library appropriately regulates
+access, caches data to minimise the impact on the infrastructure, and
+performs necessary post-processing. This library is available as open
+source and Ongoing development is coordinated on GitHub.%
+\footnote{\url{https://github.com/glasgow-ipl/ietfdata}}
+
 %--------------------------------------------------------------------------------------------------
@@
 Executive Director, and the IRTF Chair, who is a co-author on this paper)
-to ensure that our access falls within their acceptable use policies.
+and the RFC Editor to ensure that our access falls within their acceptable
+use policies.
 

### 2024-09-27 paper.tex
Update Section 2.3 (Ethics)

@@
 
-% The following is from our IMC 2021 paper:
+% The following is adapted from our IMC 2021 paper:
 
@@
 in regular contact with the IETF Tools Team and Secretariat, as well as the
-operators of the Datatracker and mailing list archive.  We have extensively
+operators of the Datatracker and mailing list archive.  We have
 discussed our work with IETF leadership (IETF and IAB Chairs, the IETF
@@
 
-Participation in the IETF is dependent on agreement to abide by the
-policies and procedures described at \url{https://www.ietf.org/about/note-well},
-including the privacy policy at \url{https://www.ietf.org/privacy-statement}.
-These make explicit provision that mailing list archives and the metadata
-contained in the Datatracker system will be made public, and it is this
-public data that we process to extract the aggregate statistics presented
-in Sections \ref{sec:documents}, \ref{sec:authors}, and \ref{sec:mail}, as
-well as the features analysed in Section \ref{sec:model}. We transfer and
-store data securely, and retain it only for the time needed to perform the
-analysis.  Since we operate entirely using the public APIs provided by
-IETF, we have no access to private data about individuals.
+Participants in the IETF must agree to abide by the policies and procedures
+described at \url{https://www.ietf.org/about/note-well}, including the privacy
+policy at \url{https://www.ietf.org/privacy-statement}. These make explicit
+provision that mailing list archives and the metadata contained in the
+Datatracker system will be made public, and it is this public data that
+we process to extract the aggregate statistics presented in Sections
+\ref{sec:trends-documents}, \ref{sec:trends-demographics}, \ref{sec:org-dynamics},
+\ref{sec:behaviour} as well as the features analysed in Section
+\ref{sec:success-factors}. We transfer and store data securely, and retain
+it only for the time needed to perform the analysis.  Since we operate
+entirely using the public APIs provided by IETF, we have no access to
+private data about individuals.
 
@@
 \section{Trends in Participant Demographics}
+\label{sec:trends-demographics}
 
@@
 \section{Understanding Participant Behaviour}
+\label{sec:behaviour}
 
@@
 \section{Success Factors}
+\label{sec:success-factors}
 

### 2024-09-27 paper.tex
Update placeholder about internet-drafts

@@
 RFCs published, starts to expand rapidly. This is further driven by the
-opening of the network to commercial and public use in the mid-1990s, and
-continues to this day.
-
-As shown, the annual RFC publication rate was highest in 2005, at the peak
+opening of the network to commercial and public use in the mid-1990s.
+The annual RFC publication rate was highest in 2005, at the peak
 of the standardisation efforts for SIP and related standards for voice-over-IP
@@
 
-\todo{Add draft publication count as mentioned in Section \ref{sec:org-dynamics:participants}}
+In addition to the published RFCs, we track work-in-progress documents,
+known as Internet-drafts, that are intended for eventual publication as
+RFCs.
+\todo{Add internet-draft counts over time as mentioned in Section \ref{sec:org-dynamics:participants}}
 

### 2024-09-28 paper.tex
Add Introduction to Section 3

@@
 
-\todo{Add introduction to trends in RFC publication; research questions}
+To begin, we review trends in IETF standards publication. We consider the
+rate of publication of documents in the RFC series, factors affecting the
+publication rate, and the occurrence of errata reports, and answer the
+following research questions:
+\begin{enumerate}
+  \item What factors affect the publication rate of RFCs and is the IETF
+    standards development process getting faster or slower over time?
+    (\S\ref{sec:trends-documents:production})
+
+  \item Are the standards produced by the IETF broadly correct? Do the
+    reported errata indicate problems with the standards development
+    process in general, or in particular areas, and is the IETF getting
+    better or worse at producing correct RFCs over time?
+    (\S\ref{sec:trends-documents:errata})
+
+\end{enumerate}
 
@@
 \subsection{RFC Production and Complexity}
+\label{sec:trends-documents:production}
 
@@
 \subsection{RFC Errata}
+\label{sec:trends-documents:errata}
 
@@
 standardisation, but also its resilience to the loss of major participants.
-In the following we consider the following research questions: 
+In this section we consider the following research questions: 
 \begin{enumerate}

### 2024-09-28 paper.tex
Update §3.1

@@
   \caption{
-    RFCs by area. \textit{Other} includes legacy RFCs, and RFCs from other
-    non-IETF streams.
+    Number of RFCs published over time, subdivided by IETF Area.
+    The areas are 
+    \emph{art} -- applications and real-time;
+    \emph{app} -- applications;
+    \emph{int} -- internetworking;
+    \emph{ops} -- operations and management;
+    \emph{rai} -- real-time applications and infrastructure;
+    \emph{rtg} -- routing;
+    \emph{sec} -- security; and
+    \emph{tsv} -- transport and services.
+    The \emph{other} category includes legacy RFCs, pre-dating the creation
+    of the IETF, and RFCs published in non-IETF publication streams such as
+    the IRTF, IAB, and the Independent stream.
   }
@@
 \pb{Role of Working Groups:}
-From the creation of the IETF in 1986, the growing community with its
+From the creation of the IETF in 1986 the growing community, with its
 interests in an increasing set of applications and protocols, has been
-split into working groups. These working groups are chartered with a focus
-on a well-defined programme of work, and exist within areas that have a
-broader focus.  As shown in Figure~\ref{fig:rfcs_by_area}, the output of
+split into a number of \emph{working groups}.\footnote{There are 126
+active working groups at the time of writing, organised into 7 areas.
+The number of groups varies over time, but is typically in the range
+100-150.} These working groups are each chartered with a well-defined
+programme of work and exist within \emph{areas} that have a broader focus.
+
+As shown in Figure~\ref{fig:rfcs_by_area}, the output of
 different areas has remained relatively stable over time. The most notable
@@
 grown to accommodate its larger community: in the early 1990s, fewer than
-20 working groups were actively publishing RFCs, while in recent years
-there has typically been at least 60 different publishing groups, with a
-peak of 97 active working groups, and other activities, in 2011.
+20 working groups were published an RFC each year, while in recent years
+there has typically been at least 60 working group publishing RFCs, with
+a peak of 97 working groups publishing RFCs in 2011 (the number of groups
+publishing RFCs is less than the total number of groups, because it takes
+time for the work of a new group to reach maturity).
 
@@
   \caption{
-    Number of publishing working groups. \textit{Other} includes legacy
-    RFCs, IRTF research groups, and non-IETF streams.
+    Number of groups publishing RFCs over time, subdivided by IETF Area.
+    The areas are as described in Figure~\ref{fig:rfcs_by_area}.
   }
@@
 
+%..................................................................................................
+\pb{Factors Affecting Publication Rate:} 
+The standards development process is taking longer.
 Figure~\ref{fig:rfcs_days_to_pub} plots the median number of days from the
-submission of an RFC's first draft, through to its publication as an RFC.
-This shows a clear trend: RFCs are taking longer to make their way through
-the standardisation and publication process. The median number of days to
-publication was 469 in 2001, rising to 1,170 in 2020. Further,
-Figure~\ref{fig:drafts_year} shows the median number of Internet-Drafts
-that are posted before an RFC is published. Days to publication and number
-of drafts are strongly correlated, suggesting that the time is spent making
-changes to the document. This may go some way towards explaining the
-decline in output of the IETF: each RFC is taking longer to produce, with
-more revisions before publication.
+submission of the first draft of a document through to its publication as
+an RFC. This shows a clear trend: RFCs are taking longer to make their way
+through the standardisation and publication process. The median number of
+days to publication was 469 in 2001, rising to 1,170 in 2020.
 
@@
 
+Similarly, Figure~\ref{fig:drafts_year} shows the median number of
+revisions each document undergoes before being published as an RFC. Days to
+publication and number of revisions are strongly correlated, suggesting
+that the time is spent making changes to the document. This may go some way
+towards explaining the decline in output of the IETF: each RFC is taking
+longer to produce, with more revisions before publication.
+
 \begin{figure}
@@
 
-%..................................................................................................
-\pb{RFC Length:}
-One may conjecture that this slowdown is driven by longer RFCs that contain
-more material. To explore this, Figure~\ref{fig:pages_year} shows the
-median page count of RFCs. This shows that the increase in the duration of
-the standardisation process for RFCs cannot be attributed to RFCs becoming
-longer: median page counts have remained stable.
+What factors affect the publication rate?  One may conjecture that this
+slowdown is driven by longer RFCs that contain more material. Yet from the
+median page count of RFCs, shown in Figure~\ref{fig:pages_year}, no clear
+trend being visible. The increase in the duration of the standardisation
+process for RFCs cannot be attributed to RFCs becoming longer: page counts
+have remained stable.
 
@@
 
-%..................................................................................................
-\pb{Relationships between RFCs:}
-RFCs themselves may be becoming more complex as a result of them needing to
-describe their relationship with older standards. As the Internet has
-evolved and matured, applications and protocols must be maintained, and new
-standards need to interoperate with existing RFCs.
-Figure~\ref{fig:updates_year} highlights this, showing the proportion of
-RFCs that are published each year that update (i.e., extend or augment) or
-obsolete (i.e., replace) one or more previously published RFCs. Intuitively,
-this percentage has slowly increased as the IETF has matured: in 2020, more
-than 30\% of RFCs updated or made obsolete a previous RFC.
+The publication rate might also be slowing because RFCs are becoming more
+complex. This could occur, for example, as a result of having to maintain
+compatibility with older RFCs as the Internet has evolved and matured.
+Figure~\ref{fig:updates_year} shows that this is the case, plotting the
+proportion of RFCs that are published each year that update (i.e., extend
+or augment) or obsolete (i.e., replace) one or more previously published
+RFCs.  This percentage has slowly increased as the IETF has matured: in
+2020, more than 30\% of RFCs updated or made obsolete a previous RFC.
+%
 Figure~\ref{fig:citations_year} expands on this, showing the median number
@@
 
-%..................................................................................................
-\pb{Use of requirements-setting language:}
-Figure~\ref{fig:keyword_usage_rates} further confirms the growing
-complexity of RFCs, showing how the use of keywords has evolved over time.
-Keywords are used in RFCs to indicate the normative requirements an RFC
-imposes on implementations. Figure~\ref{fig:keyword_usage_rates} shows the
-total number of occurrences of each of the RFC 2119~\cite{RFC2119} keywords
-(i.e., MUST, MUST NOT, REQUIRED, SHALL, SHALL NOT, SHOULD, SHOULD NOT,
-RECOMMENDED, MAY, OPTIONAL), divided by the page count of the RFC. As
+Figure~\ref{fig:keyword_usage_rates} further confirms the growing complexity
+of RFCs, showing how the use of normative language keywords has evolved
+over time.  Keywords are used in RFCs to indicate the normative requirements
+an RFC imposes on implementations. Figure~\ref{fig:keyword_usage_rates}
+shows the total number of occurrences of each of the RFC 2119~\cite{RFC2119}
+keywords (i.e., MUST, MUST NOT, REQUIRED, SHALL, SHALL NOT, SHOULD, SHOULD
+NOT, RECOMMENDED, MAY, OPTIONAL), divided by the page count of the RFC. As
 shown, the median number of keywords per page grew from 2001 through to
@@
 
-\todo{Should we include ``Academic impact of RFCs'' from IMC 2021?}
-
 %..................................................................................................
 \pb{Summary:}
-RFCs are taking longer to produce, and they go through a greater number of
-revisions before publication. Further, they increasingly update or
+RFCs are taking longer to produce, and they go through a greater number
+of revisions before publication. Further, they increasingly update or
 reference previously published RFCs, and make greater use of
-requirements-setting keywords. Yet, while these measures indicate that the
-overall complexity of standards documents is increasing, none of these
-factors strongly correlate with the time taken to publish RFCs. This
-indicates that these factors aren't driving the increasing duration of the
-standardisation process. In the sections that follow, we explore trends in
-authorship and community interaction that may also impact the process.
+requirements-setting keywords. These are all hallmarks of a maturing
+standards development organisation that must increasingly consider
+compatibility with the installed based in the development of its standards.
 

### 2024-09-29 paper.tex
Update §3.2

@@
 There is also a fifth, legacy stream, comprising RFCs published prior to
-the adoption of separate publication streams in July 2007 \cite{rfc4844}.
+the adoption of separate publication streams in July 2007 \cite{RFC4844}.
 While the IETF is an open standards forum that develops technical standards
@@
 
+% The following is from our TMA 2023 paper, Section I:
+
+Mistakes can and do occur in published RFCs.  These errors can be reported
+to the RFC Editor, and the RFC Editor makes the reports publicly available
+and coordinates with the IETF, and the other publication streams, on their
+verification. These errata can clarify editorial concerns as well as correct
+substantive technical errors. Since the presence of significant errors in
+its specification can undermine the success of a protocol, studying the
+nature of these errata and their subsequent fixes is important.
+
+In the following, we analyse the 6,759 errata filed with the RFC Editor
+between 2001–2022, inclusive, documenting 3,288 editorial issues and 3,471
+technical issues, and covering 2,240 RFCs, to understand the frequency
+and causes of problems in RFCs.
+
 % The following is from our TMA 2023 paper, Section III:
@@
 that year, but they include major documents such as RFCs 791, 792, and 793
-(the original versions of the IP~\cite{rfc791}, ICMP~\cite{rfc792}, and
-TCP~\cite{rfc793} standards), with 17, 7, and 47 errata, respectively.
+(the original versions of the IP~\cite{RFC791}, ICMP~\cite{RFC792}, and
+TCP~\cite{RFC793} standards), with 17, 7, and 47 errata, respectively.
 These important protocols clearly garnered a great deal of scrutiny and
@@
 errata filed.  This broadly correlates with the number of RFCs published
-per year, with Pearson coefficient 0.59 since 2007.  Table
-\ref{tab:top-10-rfcs-by-errata} lists the top RFCs by errata filing count.
+per year, with Pearson coefficient 0.59 since 2007.
+Table \ref{tab:top-10-rfcs-by-errata} lists the top RFCs by errata filing
+count.
 
@@
   \caption{
-    Average errata filed per RFC for each year by RFC publication year,
-    grouped by IETF area (acronyms expanded in Table~\ref{tab:errata-stats-area}).
+    Average number of errata filed per RFC in each year, by RFC publication year,
+    grouped by IETF area. The areas are as described in Figure~\ref{fig:rfcs_by_area}.
   }
@@
   \caption{
-    Days from RFC publication and errata filing by IETF area (acronyms
-    expanded in Table~\ref{tab:errata-stats-area}).
+    Cumulative distribution of the number of days from RFC publication to
+    errata filing by IETF area.
+    The areas are as described in Figure~\ref{fig:rfcs_by_area}.
   }
@@
 \pb{Errata Status:}
-Figure~\ref{fig:errata_status} categorises the errata by status and
-publication year of the RFC to which they relate.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/errata-by-status.pdf}
-  \caption{
-    Errata filings by status, by publication year of the RFC.
-  }
-  \label{fig:errata_status}
-\end{figure}
-
+Figure~\ref{fig:errata_status} categorises the errata by verification
+status and publication year of the RFC to which they relate. 
 The largest share (42.5\%) of errata are \emph{verified}: errata that have
 been has been confirmed as necessary and accurate. This suggests that many
-errata are useful to the community. The next largest share (30.3\%) are
-those labelled \emph{hold for document update}.  These are errata that are
-not a necessary update to the RFC, but may be considered on future
-revisions. For example, \emph{erratum 6278} describes an oversight in RFC
-8610~\cite{rfc8610}; the solution to this is non-trivial, and so will be
-considered in the next version of the specification. Of the 930 RFCs that
-have \emph{hold for document update} errata filed against them, only 40\%
-have been updated or obsoleted by a subsequent RFC.  We flag that this may
-be a cause for concern, or at least a missed opportunity for improvements
-to standards.
+errata identify real problems with RFCs and are useful to the community.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/errata-by-status.pdf}
+  \caption{
+    Errata filings by status, by publication year of the RFC.
+  }
+  \label{fig:errata_status}
+\end{figure}
+
+The next largest share (30.3\%) are those labelled \emph{hold for document
+update}.  These are errata that are not a necessary update to the RFC, but
+may be considered on future revisions. For example, \emph{erratum 6278}
+describes an oversight in RFC 8610~\cite{RFC8610}; the solution to this is
+non-trivial, and so will be considered in the next version of the
+specification. Of the 930 RFCs that have \emph{hold for document update}
+errata filed against them, only 40\% have been updated or obsoleted by a
+subsequent RFC.  We flag that this may be a cause for concern, or at least
+a missed opportunity for improvements to standards.
 
@@
 of the errata process and its place within the wider standardisation
-process.  Finally, 14.2\% of errata are \emph{reported} but unverified.
-Again, we are surprised to see unverified errata from over a decade ago,
-suggesting the process should be expedited.
+process.
+
+Finally, 14.2\% of errata are \emph{reported} but unverified.  Again, we
+are surprised to see unverified errata from over a decade ago, suggesting
+the process should be expedited.
 
@@
 32.7\% have attracted at least one erratum filing.  However, there are
-three notable outliers.  First, RFCs published by the Sub-IP (subip) Area
-have very few errata, with only 5\% of subip RFCs attracting errata
-filings. This is because this temporary area -- established in 2001 and
-concluded in 2005 -- only published 19 RFCs, resulting in a far smaller
-sample than the other areas. For comparison, the next smallest area,
-General (gen), published 39 RFCs. \emph{gen} RFCs attract a greater number
-of errata on average, vs. \emph{subip} RFCs, likely due to their broader
-relevance.  Second, we see that both the Application (app) and Security
-(sec) Areas' RFCs are more likely to have errata filed for them, with
-35.9\% of Application and 39\% of Security RFCs attracting at least one
-erratum filing.
+three notable outliers:
+\begin{enumerate}
+  \item RFCs published by the Sub-IP (subip) Area have very few errata,
+    with only 5\% of subip RFCs attracting errata filings. This temporary
+    area -- established in 2001 and concluded in 2005 -- only published 19
+    RFCs, resulting in a far smaller sample than the other areas. For
+    comparison, the next smallest area, General (gen), published 39 RFCs.
+    \emph{gen} RFCs attract a greater number of errata on average, vs.
+    \emph{subip} RFCs, likely due to their broader relevance.
+
+  \item We see that both the Application (app) and Security (sec) Area
+    RFCs are more likely to have errata filed for them than other areas,
+    with 35.9\% of Application and 39\% of Security RFCs attracting at
+    least one erratum filing.
+
+  \item Table~\ref{tab:errata-stats-area} details the split between
+    \emph{technical} and \emph{editorial} errata across each area.  While
+    there is broadly an even split, there are areas where one type of
+    errata is more dominant. For example, in the Routing (rtg) area, 60.8\%
+    of filings are editorial, while in the Applications (app) area, 59\%
+    were technical.  It remains to determine why this is the case, and, in
+    particular, to establish whether there is something inherent about the
+    RFCs published by these areas that makes them more prone to containing
+    errata, and to containing one type of errata vs. another. For example,
+    in the Routing area, structured notation is frequently used to define
+    routing entities; editorial errata are often filed in those
+    definitions. Targeting such areas with improved alternate review
+    procedures may be beneficial.
+
+\end{enumerate}
 
@@
   \caption{
-    CDF of errata filed per RFC by grouped by IETF area (acronyms expanded
-    in Table~\ref{tab:errata-stats-area}).
+    Cumulative distribution of errata filed per RFC by grouped by IETF
+    area.  The areas are as described in Figure~\ref{fig:rfcs_by_area}.
   }
@@
 \end{figure}
-
-Finally, Table~\ref{tab:errata-stats-area} details the split between
-\emph{technical} and \emph{editorial} errata across each area. While there
-is broadly an even split, there are areas where one type of errata is more
-dominant. For example, in the Routing (rtg) area, 60.8\% of filings are
-editorial, while in the Applications (app) area, 59\% were technical.  It
-remains to determine why this is the case, and, in particular, to establish
-whether there is something inherent about the RFCs published by these areas
-that makes them more prone to containing errata, and to containing one type
-of errata vs. another. For example, in the Routing area, structured
-notation is frequently used to define routing entities; editorial errata
-are often filed in those definitions. Targeting such areas with improved
-alternate review procedures may be beneficial.
 

### 2024-09-29 paper.tex
Remove unnecessary todo
Tweak Figure captions

@@
   \caption{
-    Citations from RFCs to other Internet-Drafts and RFCs per RFC
+    Number of citations to other Internet-Drafts and RFCs per RFC in each
+    year.
   }
@@
   \caption{
-    Keyword occurrences per page
+    Number of occurrences of normative language keywords per page in RFCs
+    in each year.
   }
@@
 
-\todo{Should we include ``Impact of Citations'' from TMA 2023?}
+% \todo{Should we include ``Impact of Citations'' from TMA 2023?}
 

### 2024-09-29 paper.tex
Complete §3.3

@@
 
-\todo{summarise results about RFC publication trends}
+On RFC Production and Complexity (\S\ref{sec:trends-documents:production})
+we see evidence that Internet standards are maturing and that the pace of
+innovation has slowed. The rate of standards publication peaked in the
+mid-2000s, and has gradually declined since, and the number of IETF working
+groups has also started to decline. The standards that are being produced
+are taking longer, with more revisions prior to publication, and make
+increasing use of normative language and references to prior work; all
+evidence of complexity and the burden of backwards compatibility. The signs
+are that the IETF has developed, and is maintaining and extending, the core
+Internet standards, but has not broken out of that niche to new topic areas.
+
+The RFC errata (\S\ref{sec:trends-documents:errata}) process seems to be
+broadly effective at finding problems, although there are a concerning
+number of unresolved errata reports that are perhaps indicative of an
+organisation that doesn't effectively maintain supposedly completed work.
+The high prevalence of errata in security-related RFCs also deserves
+further scrutiny to determine whether it is due to a higher number of
+defects in these RFCs or because defects are more likely to be noticed
+because these RFCS receive more review.
 

### 2024-09-29 paper.tex
Structure of the paper

@@
 We structure the remainder of this paper as follows.
+Section \ref{sec:background} introduces the IETF and the Internet Standards
+development process, and reviews the data with which we work.
+Section \ref{sec:trends-documents} considers trends in document production,
+complexity, and correctness.
+Section \ref{sec:trends-demographics} moves on to consider the people who
+are involved in Internet standards development and how the demographics of
+the community have shifted over time.
+Section \ref{sec:org-dynamics} studies the interaction between participants,
+explores who has influence and impact in the IETF and how that influence
+is evident in their activities, and the dynamics of communications within
+the IETF.
+Section \ref{sec:language} considers communications within the IETF from
+the perspective of language to explore whether there are differences in
+linguistic patterns and language use in more influential participants,
+and to consider how the use of language reflects the consensus-driven
+nature of the standards development process.
+Section \ref{sec:success-factors} then considers factors that lead to
+success, exploring what makes a document likely to be completed and
+used, and what makes a successful participant.
+Finally, we review related work in Section \ref{sec:related} and conclude
+in Section \ref{sec:conclusions}.
 
@@
 \section{Background and Datasets}
+\label{sec:background}
 
@@
 \ref{sec:trends-documents}, \ref{sec:trends-demographics}, \ref{sec:org-dynamics},
-\ref{sec:behaviour} as well as the features analysed in Section
+\ref{sec:language} as well as the features analysed in Section
 \ref{sec:success-factors}. We transfer and store data securely, and retain
@@
 %==================================================================================================
-\section{Understanding Participant Behaviour}
-\label{sec:behaviour}
+\section{Language and Influence}
+\label{sec:language}
 
@@
 \section{Related Work}
+\label{sec:related}
 
@@
 \section{Conclusions}
+\label{sec:conclusions}
 

### 2024-09-29 paper.tex
Update §4.1

@@
 healey:2023:power,mcquistin:2023:errare,khare:2023:tracing,
-karan:2023:leda,healey:2023:power-frontiers,barnes:2024:temporal}
+karan:2023:leda,healey:2023:power-frontiers,barnes:2024:temporal}.
 
@@
 
-% The following is from our IMC 2021 paper, Section 3.2:
+In this section, we look at the authorship of published RFCs to explore
+how participation in the IETF has changed over time. We show that the
+geographic distribution of RFC authors has shifted over time, with the
+proportion of authors from the US decreasing and a corresponding rise in
+authors from Europe and China (\S\ref{sec:trends-demographics:geographic}).
+We similarly explore shifts in affiliation, showing increasing influence of
+Huawei and ongoing shifts in the fortunes of many prominent US and European
+technology companies (\S\ref{sec:trends-demographics:affiliation}). The
+ongoing influence of Cisco and strong input from academic institutions is
+also noted.
+
+% The following is from our IMC 2021 paper, Section 3.2.
+
+%--------------------------------------------------------------------------------------------------
+\subsection{Geographic Distribution of Authors}
+\label{sec:trends-demographics:geographic}
 
@@
 dataset does not cover the entire RFC corpus, with metadata available for
-authors of RFCs published from 2001, and where it has been provided.
-Country data is available for around 70\% of authors, while affiliation
-information is given for around 80\%. In this section, we look at the
-authorship of all of the RFCs published each year, using the available
-metadata. An author is counted once in a year for each affiliation or
-location they hold.
-
-%--------------------------------------------------------------------------------------------------
-\subsection{Geographic Distribution of Authors}
-Figures~\ref{fig:author_countries_normalised} and
-\ref{fig:author_continents_normalised} show the proportion of authors from
-countries and continents, respectively. The IETF has signalled that it
-wishes to encourage greater geographical diversity~\cite{rfc7704,
-ietfblog:diversity}. Without an explicit goal, we frame our findings within
-the context of global population distribution.  We find that North America,
-while still disproportionately over represented, is becoming less dominant.
-75\% of authors were from North America in 2001, and this has declined to
-44\% in 2020. At the same time, representation of both Europe and Asia has
-grown, from 17\% to 40\% and 6\% to 14\%, respectively.  However, Africa
-and South America remain heavily underrepresented, with only $\approx$0.5\%
-of authors coming from either continent in 2020.  This suggests that, if
-the IETF is to become more geographically representative, further efforts
-are needed.
+authors of RFCs published from 2001, and where it has been provided country
+data is available for around 70\% of authors, while affiliation information
+is provided for around 80\% of authors.
+
+Figure~\ref{fig:author_countries_normalised} shows how the proportion of
+RFC authors from different countries varies over time, with Figure
+\ref{fig:author_continents_normalised} breaking this down by continent.
+The IETF has signalled that it wishes to encourage greater geographical
+diversity~\cite{RFC7704,ietfblog:diversity}.  Without an explicit goal, we
+frame our findings within the context of global population distribution.
+We find that North America, while still disproportionately over represented,
+is becoming less dominant.  75\% of authors were from North America in 2001,
+and this has declined to 44\% in 2020. At the same time, representation of
+both Europe and Asia has grown, from 17\% to 40\% and 6\% to 14\%, respectively.
+However, Africa and South America remain heavily under-represented, with
+only $\approx$0.5\% of authors coming from either continent in 2020.  This
+suggests that, if the IETF is to become more geographically representative,
+further efforts are needed.
 
@@
 \subsection{Author Affiliations}
+\label{sec:trends-demographics:affiliation}
+
 Figure~\ref{fig:author_affiliations_normalised} shows the top 10

### 2024-09-29 paper.tex
Update §4.2 and §4.3

@@
 
-Figure~\ref{fig:author_affiliations_normalised} shows the top 10
-affiliations by proportion of authors each year. Affiliation data is
-gathered from the Datatracker, and processed to normalise affiliation
-names, removing common variations in spelling, and amalgamating known
+To explore trends in author affiliations, we gather affiliation data from
+the IETF Datatracker.  This data is processed to normalise affiliation
+names by removing common variations in spelling and to amalgamate known
 subsidiaries and merged companies. For example, Huawei and Futurewei are
@@
 
-From Figure~\ref{fig:author_affiliations_normalised}, we observe several
+%..................................................................................................
+\pb{Corporate Authors:}
+Figure~\ref{fig:author_affiliations_normalised} shows the top ten
+affiliations by proportion of RFC authors each year. We observe several
 interesting trends.  First, Cisco remains a consistent employer of IETF
@@
 
-That we are able to observe changing trends in author affiliations, and in
-particular, that participants from new companies contribute, indicates that
-the IETF has broader relevance beyond a narrow set of companies. While we
-do not demonstrate any causal link, it is encouraging that commercially
+These shifts in affiliation reflect changes in fortune and commercial
+success of the companies. They also show the ongoing importance of IETF
+standards, with new companies choosing to send participants: while we do
+not demonstrate any causal link, it is encouraging that commercially
 successful companies opt to enable their employees to actively participate
@@
 ``University'', ``Institute'', or ``College'', while consultancy
-affiliations are those that contain ``Consultant``. Affiliation data has
+affiliations are those that contain ``Consultant`` (we recognise that this
+definition of consultants is limited and incomplete). Affiliation data has
 been normalised to remove common abbreviations (e.g., ``U.'' for
@@
 authors in 2001, to 13.6\% in 2020, having peaked at 16.5\% in 2009. The
-number of consultants, by our measure, has remained stable, accounting for
-2\% of authors in 2020.  The remaining authors are largely from industrial
-affiliations.
+number of consultants, by our limited measure, has remained stable,
+accounting for 2\% of authors in 2020.  The remaining authors are largely
+from industrial affiliations.
 
@@
 Columbia University, MIT, and ISI in recent years, and the rise of Tsinghua
-University and University Carlos III of Madrid.
+University.
+
+While some academic affiliations are shared by large groups of participants
+(e.g., MIT, ISI, Tsinghua University), academic participation is often
+driven by a single research group or individual within an institution.
+This allows smaller institutions, such as the University of Bremen and
+the University of Auckland, that appear in Figure~\ref{fig:author_affiliations_normalised_acad},
+and others with single but prolific authors, to have outsize influence.
+Understanding whether this indicates academics pushing for standardisation
+of the results of government funded research to increase its impact, the
+outcomes of industry sponsored research, or academics acting as consultants
+for industry, is for further study.
 
@@
 \pb{Arrival of new authors:}
-Finally, Figure~\ref{fig:author_new} shows the percentage of authors in each
-year that have not previously authored an RFC. Given that the dataset used here
-begins in 2001, 100\% of authors are new in that year. The more stable trend in
-recent years likely highlights the churn in RFC authorship, with around 30\%
-of authors each year having never previously authored an RFC.
-\todo{Not clear this fits here}
+Figure~\ref{fig:author_new} shows the percentage of authors in each year
+that have not previously authored an RFC. Given that the dataset used here
+begins in 2001, 100\% of authors are new in that year. The more stable
+trend in recent years likely highlights the churn in RFC authorship, with
+around 30\% of authors each year having never previously authored an RFC.
+This is consistent with the presence of authors with new affiliations
+and from new countries and shows an ongoing influx of new participants.
 
@@
 The trends highlighted in this section  indicate a pool of authors that is
-diversifying and changing over time, with a growing proportion of authors
+slowly diversifying and changing over time, with a growing proportion of authors
 from outside of North America, contributions from new companies and
@@
 increasingly centralised, with a third of authors coming from the top 10
-affiliations. This suggests, if the organisation is concerned with being
+affiliations. This suggests that, if the IETF is concerned with being
 more representative, further efforts are needed.

### 2024-09-29 paper.tex
Add Introduction to §5

@@
 the community have shifted over time.
-Section \ref{sec:org-dynamics} studies the interaction between participants,
+Section \ref{sec:org-dyn} studies the interaction between participants,
 explores who has influence and impact in the IETF and how that influence
@@
 we process to extract the aggregate statistics presented in Sections
-\ref{sec:trends-documents}, \ref{sec:trends-demographics}, \ref{sec:org-dynamics},
+\ref{sec:trends-documents}, \ref{sec:trends-demographics}, \ref{sec:org-dyn},
 \ref{sec:language} as well as the features analysed in Section
@@
 RFCs.
-\todo{Add internet-draft counts over time as mentioned in Section \ref{sec:org-dynamics:participants}}
+\todo{Add internet-draft counts over time as mentioned in Section \ref{sec:org-dyn:participants}}
 
@@
 \section{Organisational Dynamics of the IETF}
-\label{sec:org-dynamics}
-
-\todo{Add introduction to organisational dynamics section}
+\label{sec:org-dyn}
+
+While the IETF does hold regular plenary and interim meetings, much of the
+interaction between participants takes place on the public mailing lists of
+the various working groups. In the following (\S\ref{sec:org-dyn:participants}),
+we characterise the mailing list interactions considering the volume of
+discussion, the number of drafts discussed, the duration of individual
+participation, and how interactions between participants change over time.
+
+We then consider influence and impact of participants (\S\ref{sec:org-dyn:influence}).
+We measure how centralised is the active IETF community and how reliant is
+it on a small core of participants; how the most influential participants
+behave; how influence, determined by mailing list participation, relates to
+wider impact in the IETF; and whether the organisational affiliation of
+participants influences adoption of new work.
+
+Finally, we consider the hierarchy and communication dynamics in the IETF
+(\S\ref{sec:org-dyn:hierarchy}). We ask whether participation in IETF
+leadership roles is growing or becoming more centralised onto a small group
+of participants; whether the formal organisational hierarchy is reflected
+in the patterns of communication between participants; how people in
+different roles communicate and whether information flows up or down the
+hierarchy; and whether contact with people in leadership positions
+influences mobility within the IETF.
 
@@
 \subsection{Characterising Participant Interactions}
-\label{sec:org-dynamics:participants}
+\label{sec:org-dyn:participants}
 
@@
 \subsection{Characterising Influence and Impact}
-\label{sec:org-dynamics:influence}
+\label{sec:org-dyn:influence}
 
@@
 
-As we showed in Section \ref{sec:org-dynamics:participants}, protocol
+As we showed in Section \ref{sec:org-dyn:participants}, protocol
 standardisation is an inherently social process with most day-to-day work
@@
 
-In Section \ref{sec:org-dynamics:participants}, we demonstrated that there
+In Section \ref{sec:org-dyn:participants}, we demonstrated that there
 are three categories of participants in IETF - (a) \textit{young
@@
 participants, the degree of participation has spread  out over time.
-\todo{Repetitive with Section \ref{sec:org-dynamics:participants}?}
+\todo{Repetitive with Section \ref{sec:org-dyn:participants}?}
 
@@
 
-\todo{This is somewhat repetitive with Section \ref{sec:org-dynamics:participants}}
+\todo{This is somewhat repetitive with Section \ref{sec:org-dyn:participants}}
 
@@
 \subsection{Hierarchy and Communication Dynamics}
-\label{sec:org-dynamics:hierarchy}
+\label{sec:org-dyn:hierarchy}
 
@@
     of the conversation?
-    (\S\ref{sec:org-dynamics:hierarchy:rq1})
+    (\S\ref{sec:org-dyn:hierarchy:rq1})
 
@@
     communication patterns?
-    (\S\ref{sec:org-dynamics:hierarchy:rq2})
+    (\S\ref{sec:org-dyn:hierarchy:rq2})
 
@@
     information flow ``up" or ``down" the hierarchy?
-    (\S\ref{sec:org-dynamics:hierarchy:rq3})
+    (\S\ref{sec:org-dyn:hierarchy:rq3})
 
@@
     activity over time? How does this vary across hierarchy level?
-    (\S\ref{sec:org-dynamics:hierarchy:rq4})
+    (\S\ref{sec:org-dyn:hierarchy:rq4})
 
@@
 \subsubsection{Methodology}
-\label{sec:org-dynamics:hierarchy:methodology}
+\label{sec:org-dyn:hierarchy:methodology}
 
@@
 \subsubsection{Centralisation and Dominance}
-\label{sec:org-dynamics:hierarchy:rq1}
+\label{sec:org-dyn:hierarchy:rq1}
 
@@
 \subsubsection{Organisational Hierarchy and Communication Patterns}
-\label{sec:org-dynamics:hierarchy:rq2}
+\label{sec:org-dyn:hierarchy:rq2}
 
@@
 \subsubsection{Communication up and down the Hierarchy}
-\label{sec:org-dynamics:hierarchy:rq3}
+\label{sec:org-dyn:hierarchy:rq3}
 
@@
 \subsubsection{Mobility Within the Hierarchy}
-\label{sec:org-dynamics:hierarchy:rq4}
+\label{sec:org-dyn:hierarchy:rq4}
 

### 2024-09-29 paper.tex
Update Volumen of Discussion in §5.1

@@
 \subsection{An IETF Primer}
+\label{sec:background:ietf}
 
@@
 \subsection{Data Sources}
+\label{sec:background:data}
 
@@
 interests in an increasing set of applications and protocols, has been
-split into a number of \emph{working groups}.\footnote{There are 126
+split into a number of \emph{working groups}.\footnote{There are 122
 active working groups at the time of writing, organised into 7 areas.
@@
 \pb{Volume of Discussion:}
-Figure~\ref{fig:pid_count_emailing_yearly} presents the number of emails
-sent across the last 25 years, showing that email volumes have grown
-significantly with time, plateauing at around 130,000 annual messages since
-around 2010. The figure also shows how the number of unique Person ID
-counts (as described in \S\ref{sec:data}) observed every year in the mail
-lists, showing a decreasing trend in the number of contributors.
-
-Figure~\ref{fig:emailvol_by_year_catg} shows the relative email volume by
-type. We see that an increasing fraction of messages originate from
-automated addresses. This is largely as a result of the growing use of
-GitHub, and other version control systems, for managing drafts. There were
-122 active IETF working groups at the time of writing: of these, 17 listed
-a GitHub repository in their metadata. The QUIC working group, as one
-example, has replaced the typical email list discussions with GitHub
-issues: indeed, this is a significant part of the surge observed in 2016.
-Given that most working groups use mailing lists to manage their activity,
-we do not further analyse interactions that take place on GitHub in this
-paper. As a result, Figure~\ref{fig:emailvol_by_year_catg} understates the
-volume of interactions: the plateau observed in recent years is at least
-somewhat attributable to the shift to GitHub and similar services. As this
-shift continues, it will become important for future work to consider these
-interactions.
-\todo{We have more discussion of GitHub?}
+Figure~\ref{fig:pid_count_emailing_yearly} shows the number of emails sent
+across the last 25 years (dashed red line), showing that email volumes have
+grown significantly with time, plateauing at around 130,000 messages per
+year since around 2010. The figure also shows (solid blue line) how the
+number of unique participants (with entity resolution performed as
+discussed in Section \ref{sec:background:data}) observed every year in the
+mail lists, showing a decreasing trend in the number of contributors since
+2007. We note that that the number of participants is broadly correlated
+with the rate of RFC publication (Figure \ref{fig:rfcs_by_area}), likely
+indicating that many participants disengage once the document they were
+working on is published.
 
@@
   \caption{
-    Number of Person IDs exchanging emails each year
+    Number of email messages per year and number of unique participants
+    exchanging emails per year.
   }
@@
 \end{figure}
+
+Figure~\ref{fig:emailvol_by_year_catg} breaks down the messages sent into
+four categories based on the sending address:
+\begin{enumerate}
+  \item \emph{Datatracker-mapped} addresses represent those participants
+    sending email in a given year that have an IETF Datatracker account
+    associated with their email address. Datatracker accounts are needed to
+    submit drafts and register for meetings, and to allow those in
+    leadership roles to perform administrative actions, so this indicates
+    the fraction of participants who are strongly engaged with the IETF.
+
+  \item \emph{Automated} addresses are those sent by automated systems
+    rather than directly by participants. This includes the automated
+    announcement of new draft submissions and meeting, last calls for
+    comments, etc., sent by the Datatracker, but there is also a growing
+    number messages sent by GitHub and other version control systems used
+    for managing drafts.
+    There were 122 active IETF working groups at the time of writing.
+    Of these, 17 listed a GitHub repository in their metadata. The QUIC
+    working group, as one example, has replaced the typical email list
+    discussions with GitHub issues: indeed, this is a significant part of
+    the surge observed in 2016.  Given that most working groups use mailing
+    lists to manage their activity, we do not further analyse interactions
+    that take place on GitHub.
+
+  \item \emph{Role-based} addresses represent those sent by people in
+    formal leadership or administrative roles in the context of their role.
+    There are a very limited number of these addresses, e.g., the IETF,
+    IRTF, and IAB Chairs, the IETF Executive Director. As seen, these
+    addresses are used sparingly. Note that working group chairs and area
+    directors use their individual email addresses when managing their
+    working group or area, and so appear as Datatracker mapped addresses.
+
+  \item \emph{Unmapped} addresses are those that do not fall into the other
+    categories. They typically represent participants that lurk on a
+    mailing list and provide occasional input, but who are not document
+    authors and do not attend meetings. This category has reduced over time
+    as more types of interaction require a Datatracker account and as
+    comments from occasional contributors are increasingly submitted via
+    issue trackers on GitHub and similar services.
+
+\end{enumerate}
+
+The increasing use of GitHub suggests that the data in
+Figure~\ref{fig:pid_count_emailing_yearly} likely understates the volume of
+interactions. The plateau observed in the number of messages sent in recent
+years is at least somewhat attributable to the shift to GitHub and similar
+services. As this shift continues, it will become important for future work
+to consider these interactions.
 

### 2024-09-29 paper.tex
Update Discussion of Draft Documents in §5.1

@@
 %..................................................................................................
-\pb{Discussion of drafts:}
+\pb{Discussion of draft documents:}
 The mailing lists are, other than plenary or interim meetings, the primary
-means for the discussion of drafts.  To measure how often drafts are
-discussed, we next identify mentions of drafts in mailing list messages.
-We therefore extract any mention of a draft (beginning \texttt{draft-}) or
+means for the discussion of draft documents. To measure how often drafts are
+discussed, we identify mentions of Internet-Drafts and RFCs in mailing list
+messages.  We extract any mention of a draft (beginning \texttt{draft-}) or
 RFC (i.e., ``RFC'' followed by a number).  Figure~\ref{fig:draft_mentions}
@@
 production.
-\todo{Add discussion of number of drafts published over time in Section
-\ref{sec:trends-documents}. Does this tie into the discussion of increasing
-complexity of draft publication?}
+\todo{Add and cross-reference a discussion of number of drafts published
+over time in Section \ref{sec:trends-documents}. Does this tie into the
+discussion of increasing complexity of draft publication?}
 
@@
   \caption{
-    Number of draft mentions in each year found in the mail lists
+    Number of mentions of Internet-Drafts and RFC in the IETF email
+    archive, per year.
   }

### 2024-09-29 paper.tex
Rework Duration of Participation in §5.1

@@
 %..................................................................................................
-\pb{Contribution duration.}
-We now look into the longevity of contributors, in terms of the number of
-years that they actively participate in any IETF mailing lists. We define
-the \textit{contribution duration} of a participant as the length of time
-that they have contributed to the mailing lists.  To do this, we look at
-the contributors who first contribute to mailing lists between the years
-2000 to 2013. We limit our analysis to 2013, since the longevity of
+\pb{Duration of Participation:}
+We now look into the duration of participation, in terms of the number of
+years that people actively participate in any IETF mailing lists.
+
+We define the \textit{contribution duration} of a participant as the length
+of time that they have contributed to the mailing lists.  To do this, we
+study the participants who first send email to mailing lists between the
+years 2000 to 2013. We limit our analysis to 2013, since the longevity of
 contributors that first participate more recently cannot be determined.
 For each year, in the period 2000 to 2013, we look at those who first
-contribute in each year and the number of years that they then go on to
+sent email in each year and the number of years that they then go on to
 remain active in the mailing lists. For example, a participant who first
@@
 
-We generate Gaussian Mixture Models for observing different clusters of the
-maximum duration that contributors remain active for.  These models
-identified three broad clusters that contributors can be categorised
-within:
-
-\begin{itemize}
-  \item \emph{Young contributors} who leave within 1 year of joining.
-  \item \emph{Mid-age contributors} who go onto remain associated for
-    more than 1 year, but less than 5 years.
-  \item \emph{Senior contributors} who go onto remain associated for 5 or
-    more years.
-\end{itemize}
-
-We use these contribution categories defined above to characterise the
-interactions between contributors belonging to different categories, as
-well as the authors of each RFC. We look at outgoing and incoming
-interactions with RFC authors in the period between the first draft and
-publication of the RFC. If this period is less than two years, we look at
-the activities of authors for two years before the RFC was published.
-Interactions are defined from the viewpoint of the RFC authors:
-
-\begin{itemize}
-  \item \textit{Outgoing interaction}, where an RFC author responds
-    to an email from other contributors (i.e., email sent).
-
-  \item \textit{Incoming interaction}, where contributors respond to
-    an email by the author (i.e., email received).
-\end{itemize}
-
-For each interaction type, we count the number of messages and number of
-unique contributors involved in those interactions. For each RFC, we
-analyse the interaction activities based on three measures of author
-contribution duration:
-
-\begin{itemize}
-  \item \emph{Junior-most author}, i.e., the author with the lowest
-    contribution duration at the time of publication.
-  \item \emph{Senior-most author}, i.e., the author with the highest
-    contribution duration at time of publication.
-  \item \emph{Mean contribution duration} of all of the authors of the
-    RFC at the time of publication.
-\end{itemize}
+First, to understand duration of participation, we first generate Gaussian
+Mixture Models for observing different clusters of the maximum duration of
+contribution. These models identified three broad clusters into which the
+participants can be categorised: \emph{young contributors} for whom the
+time between their first and last emails to the IETF lists in one year or
+less; \emph{mid-age contributors} who go onto remain active participants in
+the email lists for more than one year, but less than five years; and
+\emph{senior contributors} who remain participants in the mailing list for
+five or more years.
+
+Next, to begin to understand how contribution duration relates to RFC
+authorship, we look at the email interactions of the authors of each RFC.
+Specifically, we look at the distribution of the contribution duration of
+the authors of RFCs, considering the \emph{mean contribution duration} of
+all of the authors of the RFC at the time of publication, the contribution
+of the \emph{junior-most author} (the author with the lowest contribution
+duration at the time of publication), and the contribution of the
+\emph{senior-most author} (the author with the highest contribution
+duration at time of publication.  Figure~\ref{fig:age_dist_rfc_authors}
+shows the distribution of contribution duration of each of these three
+measures. This shows that the majority of junior-most authors have
+participated for less than 5 years in the IETF (i.e., they are young or
+mid-age contributors), whereas the majority of senior-most authors are
+senior contributors who have participated in the IETF for significantly
+more than 5 years (in fact, 35\% of authors exceed 15 years of IETF
+participation).  This shows that RFCs tend to be authored by a mix of
+seniority levels.
+
+% We look at outgoing and incoming interactions with RFC authors in the
+% period between the first draft and publication of the RFC. If this period
+% is less than two years, we look at the activities of authors for two
+% years before the RFC was published.  Interactions are defined from the
+% viewpoint of the RFC authors: \emph{outgoing interactions} where an RFC
+% author responds to an email from other contributors (i.e., email sent);
+% and \emph{incoming interactions} where contributors respond to an email
+% by the author (i.e., email received).
 
@@
 \end{figure}
-
-Figure~\ref{fig:age_dist_rfc_authors} shows the distribution of
-contribution duration of each of these three measures. This shows that the
-majority of junior-most authors have participated for less than 5 years in
-the IETF, whereas the majority of senior-most authors have participated in
-the IETF for more than 10 years. In fact, 35\% of authors exceed 15 years
-of IETF participation, showing that RFCs tend to be authored by a mix of
-seniority levels.
 

### 2024-09-29 paper.tex
Update Evolution of Interactions in §5.1

@@
 \pb{Evolution of interactions:}
-Figure~\ref{fig:degree_dist_rfc_authors} shows the gradual drift in annual
-degree (i.e., the number of people interacted with) of RFC authors. The
-degree of authors has substantially increased over time. For instance, in
-the year 2000 only 5.5\% of the authors had a degree of over 25, whereas by
-the year 2015 almost a quarter of the authors had a degree over 25.  This
-confirms that, on average, more recent RFCs generate greater discussion.
-This may explain increasing publication times: authors spend more time
-interacting with other participants, and these discussions are likely to
-lead to more drafts.
-\todo{Replace with material from one of the later papers?}
+The analysis in Section \ref{sec:trends-documents:production} showed that
+RFCs are taking longer to publish and show evidence of increasing complexity.
+Such increased complexity should be visible in the email discussion during
+preparation of the RFCs.
+
+Figure~\ref{fig:degree_dist_rfc_authors} explores this, showing the change
+in the number of other people RFC authors interacte with (i.e., the annual
+degree of their communication graph) over time. The amount of interactions
+RFC authors engage in has substantially increased over time. For instance,
+in the year 2000 only 5.5\% of the authors had a degree of over 25 (i.e.,
+they exchanged with more than 25 other people), whereas by the year 2015
+almost a quarter of the authors had a degree over 25. This confirms that,
+on average, more recent RFCs generate greater discussion. This helps to
+explain increasing publication times: as documents become more complex,
+authors spend more time interacting with other participants to resolve
+the complexity, and these discussions are likely to lead to more drafts.
 
@@
 
-We can also contrast the interaction patterns of the junior vs.\ senior
+We also contrast the interaction patterns of the junior vs.\ senior
 authors. Figure~\ref{fig:junior_senior_indegree_seniormem} presents the
-CDFs of the in-degree across both junior and senior authors.  It shows that
-the incoming interactions from senior contributors to junior authors are
-significantly less than the incoming interactions from senior contributors
-to senior authors. Nearly 55\% of junior authors receive messages from
-fewer than 10 senior contributors, whereas nearly 65\% of senior authors
-receive messages from more than 10 senior contributors.  This speaks to the
-differing roles played by these sub-populations: senior authors act as hubs
-through which substantial volumes of interaction flow.
-\todo{Replace with material from one of the later papers?}
+CDFs of the in-degree (i.e., number of messages received) across both
+junior and senior authors. It shows that the incoming interactions from
+senior contributors to junior authors are significantly less than the
+incoming interactions from senior contributors to senior authors. Nearly
+55\% of junior authors receive messages from fewer than 10 senior
+contributors, whereas nearly 65\% of senior authors receive messages from
+more than 10 senior contributors.  This speaks to the differing roles
+played by these sub-populations: senior authors act as hubs through which
+substantial volumes of interaction flow.
 

### 2024-09-29 paper.tex
Update intro to §5.2

@@
 standardisation is an inherently social process with most day-to-day work
-happening on public mailing lists, aided by meetings, video conference
-calls, and open document and code repositories.  We are specifically
+happening on public mailing lists.  We are specifically
 interested in better understanding how influence is distributed across

### 2024-09-29 paper.tex
Update 5.2.1

@@
 % - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+\subsubsection{Measuring Influence: The IETF as a Social Graph}
+\label{subsec:measuring_influence}
+
 % ICWSM 2022 paper Section 3
-\subsubsection{Measuring Influence: The IETF as a Social Graph}
-
-\label{subsec:measuring_influence}
-
-Much of the interaction between IETF participants occurs on public mailing
-lists.  For each year, we build a social graph based on the \emph{active
-community} of participants (nodes) who have interactions (edges) with any
-other participant in the previous 5 years.
-
-In Section \ref{sec:org-dyn:participants}, we demonstrated that there
-are three categories of participants in IETF - (a) \textit{young
-contributors} -- who leave within one year of their first year mailing list
-contribution; (b) \textit{mid-age contributors} -- participants who stay
-active for up-to 5 years; and (c) \textit{senior contributors} -- who
-remain active for more than 5 years. Following this, a 5 year period window
-is chosen to observe interactions. There are no direct
-participant-to-participant email exchanges in our dataset: the emails and
-responses we capture are sent to the public mailing lists, and we observe
-an interaction between two participants when one replies to an email sent
-by another on any mailing list. This yields a social graph based on
-1,049,793 emails (out of 2.1M) from 22,138 unique participants across 840
-mailing lists.
+
+To measure influence of participants in the IETF we extend the analysis in
+Section \ref{sec:org-dyn:participants} and, for each year, build a social
+graph based on the \emph{active community} of participants (nodes) who have
+email interactions (edges) with any other participant in the previous five
+years.
+
+We previously demonstrated that there are three categories of participants
+in IETF: \textit{young contributors} who leave within one year of their
+first year mailing list contribution; \textit{mid-age contributors} who
+stay active for up-to five years; and \textit{senior contributors} who
+remain active for more than five years (\S\ref{sec:org-dyn:participants}).
+Following this, a 5-year period window is chosen to observe interactions.
+There are no direct participant-to-participant email exchanges in our
+dataset: the emails and responses we capture are sent to the public mailing
+lists, and we observe an interaction between two participants when one
+replies to an email sent by another on any mailing list. This yields a
+social graph based on 1,049,793 emails (out of 2.1M; not all messages
+receive a reply) from 22,138 unique participants across 840 mailing lists.
 
@@
 To examine the variation in participation between IETF participants, in
-Figure~\ref{fig:degree_cdf} we plot the cumulative degree distribution of
-the social graph -- i.e., the cumulative number of people a participant
-interacts with.  We observe a core group that always interacts with
+Figure~\ref{fig:degree_cdf} we plot the cumulative degree distribution
+of the social graph (i.e., the cumulative number of people with which a
+participant interacts). We observe a core group that always interacts with
 substantially more people than the rest of the community, with around 80\%
-of the participants having degree  $<$15, but 5\%-10\% having degree $>$40.
-However,  this difference decreases over time: in 2005-2009 around 6\% of
+of the participants interacting with less than 15 others (i.e., having
+degree~$<15$), but 5\%-10\% interacting with more than 40 others.
+This difference decreases over time: in 2005-2009 only around 6\% of
 participants had degree $>$40, but by the period 2015-2019 this increases
-to $\sim10\%$.  This shows that while there is always a core of more active
-participants, the degree of participation has spread  out over time.
-\todo{Repetitive with Section \ref{sec:org-dyn:participants}?}
+to around 10\%.
+While there is always a core of more active participants, the amount of
+interaction has increased over time.
+This confirms the results in Figure \ref{fig:degree_dist_rfc_authors},
+for RFC authors, and shows that they also apply to interactions in the
+community at large.
 
@@
 2006 before stabilising.  Overall, we observe that the IETF community has
-become less fragmented.
+become less fragmented over time.
 
@@
     Size of Largest Connected Component (LCC) and Number of Connected
-    Components (NCC)
+    Components (NCC) of the email graph.
   }
@@
 period~\cite{kourtellis2013identifying, weitzel2012measuring,
-sole2014centrality}. We also considered other graph based influence metrics
-such as eigenvector centrality: eigenvector centrality reflects the
-importance of a node as per its neighbours, while betweenness centrality is
-based on shortest paths which is independent of the influence of
-neighbours.
-
-However, we found a very strong correlation between the two measures, 
-in line with similar experiments \cite{valente2008correlated,he2016correlation}:
-the Spearman's rank correlation of participants ranked by betweenness
+sole2014centrality}.%
+\footnote{
+  We considered other graph based influence metrics such as eigenvector
+  centrality: eigenvector centrality reflects the importance of a node as
+  per its neighbours, while betweenness centrality is based on shortest
+  paths which is independent of the influence of neighbours.  However, we
+  found a very strong correlation between the two measures, in line with
+  similar experiments \cite{valente2008correlated,he2016correlation}: the
+  Spearman's rank correlation of participants ranked by betweenness
   centrality and eigenvector centrality ranged between 0.51-0.72, with a
@@
   \cite{hagen2018crisis,chen2012identifying, ghalmane2019centrality}.
-
+}
 Figure~\ref{fig:purge_nodes_largest_connected_component_number_components}

### 2024-09-29 paper.tex
Update §5.2.2 (cross-area review; participation duration, etc)

@@
 % ICWSM 2022 paper Section 3.1
-\pb{Influence}
+\pb{Influence:}
 Relying on a small group to interconnect the community could undermine the
@@
 role in interconnecting the wider community, we compute the betweenness
-centrality of each participant in a given time
-period~\cite{kourtellis2013identifying, weitzel2012measuring,
-sole2014centrality}.%
+centrality of each participant in a given time period
+\cite{kourtellis2013identifying,weitzel2012measuring,sole2014centrality}%
 \footnote{
@@
 }
+to determine the most influential participants in the email graph.
 Figure~\ref{fig:purge_nodes_largest_connected_component_number_components}
@@
 % - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-% ICWSM 2022 paper Section 3.2
 \subsubsection{Behaviour of Influential Participants}
@@
 
-\todo{This is somewhat repetitive with Section \ref{sec:org-dyn:participants}}
-
-We now characterise the behaviour of the most influential participants, in
-terms of the volume of emails sent, length of time active within the
-community, and topics discussed.
-
-%..................................................................................................
-\pb{Email volume}
+% ICWSM 2022 paper Section 3.2
+
+We now characterise the behaviour of the most influential participants, as
+measured by betweenness centrality (\S\ref{subsec:measuring_influence}), in
+terms of the volume of emails they send, the length of time they ate active
+within the community, and the topics discussed.
+
+%..................................................................................................
+\pb{Email volume:}
 Figure~\ref{fig:emailcount_top_percentile} shows that each participant in
@@
 10\% most influential participants account for 43.75\%, on average, of the
-total emails, a substantially larger proportion than the others.
-
-This dynamic seems stable over time, making differences even more acute.
-At the same time, the overall number of emails sent increases up to 2010
-and remains roughly stable from then onward.  Along with the results from
-\S\ref{subsec:measuring_influence}, this shows a worrying, if slowly
-improving, dependence of the IETF on a small number of influential
-participants.
+total number of emails sent, a substantially larger proportion than the
+other participants.
+
+This ratio seems broadly stable over yet the overall number of emails sent
+increases up to 2010 and remains roughly constant from then onward.  Along
+with the results from \S\ref{subsec:measuring_influence}, this shows a
+worrying, if slowly improving, dependence of the IETF on a small number of
+highly influential participants.
 
@@
 %..................................................................................................
-\pb{Cross-area review} 
+\pb{Cross-area review:} 
 In addition to sending more emails, we test if influential participants
-engage with different parts of the community more. The work of the IETF is
-divided into several areas (e.g., Applications \& Real-time, Security, and
-Routing). This eases administration, but might also act as a barrier to
-broader discussion and review. For example, WebRTC standards developed in
-the Applications and Real-time area might contain elements that could
-benefit from expertise from the Security area, but participants in one area
-might not review work in another.
+engage with different parts of the community more often than do typical
+participants. That is, we measure if influential participants are more
+likely to make contributions to working groups in multiple IETF Areas.
 
 Figure~\ref{fig:areas_participation_top_percentile} shows the mean number
-of areas where IETF participants are active (derived from the mailing lists
-they use).  We see that influential participants engage in more areas of
-the IETF, on average, and that cross-area engagement has improved over
-time, indicating that the community itself has matured. This indicates that
-influential participants benefit the IETF in enabling cross-area discussion
-and review, and can bridge administrative divisions.
+of areas where participants in the IETF are active (based on the set of
+mailing lists they sent do). We see that more influential participants
+engage in more areas of the IETF, on average, than do less influential
+participants: not only do influential participants send more email, they
+send email to working groups in more areas of the IETF. This indicates
+that influential participants benefit the IETF in enabling cross-area
+discussion and review, and can bridge administrative divisions.
+
+It can also be seen that cross-area engagement has improved over time,
+with participants as a whole sending email to groups in more areas over
+tine. This shows that the IETF community as a whole is discussing more 
+broadly than it used to, perhaps further indication of the increasing
+complexity of the standards development and widespread concern about
+feature interactions.
 
@@
 %..................................................................................................
-\pb{Participation duration}
-We next consider for how long participants remain associated with the IETF
-(measured as the delta between first and last emails sent).
+\pb{Participation duration:}
+We next consider how long participants remain associated with the IETF,
+measured as the time between sending their first and last emails to one of
+the mailing lists.
+
 Figure~\ref{fig:age_top_percentile} shows the mean participation duration
@@
 most influential participants are also the most experienced.
+The short participation duration of the less influential participants
+matches the participant churn discussed in Figure \ref{fig:author_new}
+in \S\ref{sec:trends-demographics:affiliation}.
 
@@
 
-We extend this observation by considering what happens to participants once
-they become influential. Figure \ref{fig:heatmap_top10_overlap} shows what
-proportion of the top 10\% influential participants in any given year
-(\emph{y-axis}) was also among the top 10\% influential participants in any
-other year (\emph{x-axis}). This shows that a significant majority of
-participants that become influential, continue to be influential for a
-number of years: the top 10\% are influential for at least 6-7 years on
-average.
+We further consider whether participants who become influential, remain
+influential. Figure \ref{fig:heatmap_top10_overlap} shows what proportion
+of the top 10\% influential participants in any given year (\emph{y-axis})
+was also among the top 10\% influential participants in any other year
+(\emph{x-axis}). This shows that a significant majority of participants
+that become influential, continue to be influential for a number of years:
+the top 10\% are influential for at least 6-7 years on average.
 
@@
 
-Further, Figure \ref{fig:rate_new_influential_participants} shows that
+Finally, Figure \ref{fig:rate_new_influential_participants} shows that
 breaking into the top 10\% of influencers requires an increasing number of
@@
 %..................................................................................................
-\pb{Topics of Discussion} 
+\pb{Topics of Discussion:} 
 The topics discussed on WG mailing lists are a good indicator of the focus

### 2024-09-29 paper.tex
Finish working on §5.2.2

@@
 mail lists, showing a decreasing trend in the number of contributors since
-2007. We note that that the number of participants is broadly correlated
+2007. We note that the number of participants is broadly correlated
 with the rate of RFC publication (Figure \ref{fig:rfcs_by_area}), likely
@@
 \pb{Topics of Discussion:} 
-The topics discussed on WG mailing lists are a good indicator of the focus
-of technical work, and it might be expected that influential participants
-will set the direction of that work.  To explore this, we use the Latent
-Dirichlet Allocation (LDA) \cite{blei2003latent,hoffman2010online} model
-from \texttt{gensim} \cite{rehurek_lrec} to induce 100 topics on the entire
-set of email texts. 
-Each topic is a distribution over words: e.g., a \emph{security} topic might
-have high probability for \emph{cypher}, \emph{rsa}, \emph{auth}, and
-related words. We can also use the model to obtain a vector for any input
-text as a sparse distribution over all topics, e.g., finding that a text is
-comprised of 30\% security and 70\% video streaming topics. With this, we
-generate a vector for each participant in a given time period by
-concatenating all messages sent by that participant in the period and
-feeding it into the LDA model. The result can be loosely interpreted as a
-distribution of topics on which each participant works.
+The topics discussed on working group mailing lists are a good indicator
+of the focus of technical work, and it might be expected that influential
+participants will set the direction of that work. To explore this, we use
+the Latent Dirichlet Allocation (LDA) \cite{blei2003latent,hoffman2010online}
+model from \texttt{gensim} \cite{rehurek_lrec} to induce 100 topics on the
+entire set of email texts.
+
+Each topic is a distribution over words. For example, a \emph{security}
+topic might have high probability for \emph{cypher}, \emph{rsa},
+\emph{auth}, and related words. We can also use the model to obtain a
+vector for any input text as a sparse distribution over all topics, e.g.,
+finding that a text is comprised of 30\% security and 70\% video streaming
+topics. With this, we generate a vector for each participant in a given
+time period by concatenating all messages sent by that participant in the
+period and feeding it into the LDA model. The result can be loosely
+interpreted as a distribution of topics on which each participant works.
 
@@
 increased awareness of privacy, especially in light of the Edward Snowden
-leaks \cite{RFC7258}. We also find that as videoconferencing became
+leaks \cite{RFC7258,RFC9446}. We also find that as videoconferencing became
 increasingly popular, so did the WebRTC protocol that frequently underpins

### 2024-09-29 paper.tex
Update §5.2.3

@@
 
-While these top participants are influential within the \emph{mailing
-lists}, it is unclear how this influence translates into document
-authorship and into gaining  leadership roles within  the organisation.
-
-%..................................................................................................
-\pb{Draft authorship}
+While the top participants are defined by the betweenness centrality metric
+(\S\ref{subsec:measuring_influence}) as being influential within the IETF
+mailing lists, it is unclear how this influence translates into document
+authorship and leadership roles within the IETF.  We explore these aspects
+of influence in the following.
+
+%..................................................................................................
+\pb{Document authorship:}
 The output of the IETF is technical standards and other documents published
-in the RFC series. RFCs are developed by WGs from a sequence of Internet-Drafts,
-with individuals acting as named authors.  Having identified the most
-influential participants in the mailing list community, we can determine
-whether or not these same participants are also the most active authors.
+in the RFC series. RFCs are developed by working groups from a sequence of
+Internet-drafts, with individuals acting as named authors. Having identified
+the most influential participants in the mailing list community, we can
+determine whether or not these same participants are also the most active
+authors.
 
@@
 the proportion of documents authored by mailing list participants, sorted
-by their influence rank. We observe a growing proportion of the community
-involved in draft authoring, skewed towards influential participants who
-tend to write more drafts.  During 2010-2019, each participant in the top
-10 percentile, authored 0.125\% to 0.175\% of the total number of drafts in
-that period. We find that 32\%-42\% of the total drafts, in different years
-during this period, were authored by the top 10\% most influential mailing
-list participants.
+by their influence rank. We observe that:
+(i) influential participants in the email graph, i.e., those ranked top in
+betweenness centrality, tend to write more drafts than do participants with
+lower centrality; and
+(ii) the number of drafts per participant is increasing over time for all
+participants.
+For example, during 2010-2019, each participant in the top 10 percentile,
+authored 0.125\% to 0.175\% of the total number of drafts in that period.
+We find that 32\%-42\% of the total drafts, in different years during this
+period, were authored by the top 10\% most influential mailing list
+participants.
 
@@
 %..................................................................................................
-\pb{Co-authorship and email graph correlation}
+\pb{Document co-authorship and correlation with the email graph:}
 Figure~\ref{fig:drafts_initiation_top_percentile} showed that influential
-mailing list participants tend to author more drafts than others.
+participants on the mailing lists tend to author more drafts than others.
 We next ask if they are also influencers in the draft co-authorship graph.
 
-Thus, we create a draft co-authorship graph where each author is a node,
-and draft co-authorship is an edge. We then measure influence with
+To do this, we create a draft co-authorship graph where each author is a
+node, and draft co-authorship is an edge. We then measure influence with
 betweenness centrality of the authors in each time period.
@@
 Figure \ref{fig:non_overlapping_coauthornetwork_authors_emailnetwork}.
-We observe that not all the prolific authors are that engaged in the email
-discussion: 40\%-50\% of non-overlapping authors are ranked between 20th to
-40th percentile of influence in the email network. These non-overlapping
-authors are typically more junior with respect to their participation
-duration.
+We observe that that are some prolific authors are that \emph{not} engaged
+in the email discussion: 40\%-50\% of non-overlapping authors are ranked
+between 20th to 40th percentile of influence in the email network. These
+non-overlapping authors are typically more junior with respect to their
+participation duration.
 
@@
 %..................................................................................................
-\pb{Leadership roles} 
-WG chairs are selected from the community, and we might expect those
-selected to be influential in the community.
+\pb{Influencers and leadership roles:} 
+Working group chairs are selected from the community, so we might expect
+those selected to be influential in the community.
 Figure~\ref{fig:wg_chair_before_after} shows that this is indeed the case:
-87.5\% of WG chairs are in the top 20\% of mailing list influencers and
-67.4\% are in the top 20\% of document authors in the year before they
-became chairs. This also shows the impact of taking up a leadership role:
-influence in both the mailing list and authorship communities grows in the
-year after participants become a WG chair for the first time.
+87.5\% of working group chairs are in the top 20\% of mailing list
+influencers and 67.4\% are in the top 20\% of document authors in the year
+before they became chairs. This also shows the impact of taking up a
+leadership role: influence in both the mailing list and authorship
+communities grows in the year after participants become a working group
+chair for the first time.
 
@@
 %..................................................................................................
-\pb{Organisations}  While participants in the IETF contribute as
-individuals, they are usually affiliated with an organisation.  To study
-the potential influence of organisations, for each draft we obtain the
-authors' affiliations from the Datatracker. If this is not available, we
-use the domain name of the authors' email addresses, which we map to the
-relevant organisation (e.g., \texttt{@cisco.com} maps to Cisco). For generic
-email addresses, such as \texttt{@gmail.com}, we use the participant name
-if no affiliation is available.
+\pb{Influence of organisations:}
+While participants in the IETF are expected to contribute as individuals,
+they are usually affiliated with an organisation.  To study the potential
+influence of those organisations, for each draft we obtain the authors'
+affiliations from the Datatracker. If this is not available, we use the
+domain name of the authors' email addresses, which we map to the relevant
+organisation (e.g., \texttt{@cisco.com} maps to Cisco). For generic email
+addresses, such as \texttt{@gmail.com}, we use the participant name if no
+affiliation is available.
 
 Figure \ref{fig:coauthor_affiliation_network2019_otheryears} shows the ten
-most frequent affiliated organisations of the top 20\% authors in the
-period 2015-2019, and the number of authors affiliated with these
+most frequent affiliated organisations of the top 20\% of draft authors in
+the period 2015-2019, and the number of authors affiliated with these
 organisations in  each  period.  While the dominance of Cisco is clear,
@@
 collaborations between authors from Huawei and other organisations appears
-to be unaffected by the recent addition of Huawei to the U.S. Entity List
-\cite{bis:2019:entity} as the trends in Figure~\ref{fig:heatmap_2019_coauthordraft_most_frequent_org}
-also hold for 2020-2021.\footnote{\tiny\url{https://mailarchive.ietf.org/arch/msg/ietf-announce/0ywjgSS4LlO0DaWDoLJLRHxJdUk/}}
+to be unaffected by the addition of Huawei to the U.S. Entity List
+\cite{bis:2019:entity} as the trends in Figure 
+\ref{fig:heatmap_2019_coauthordraft_most_frequent_org} also hold for 2020-2021.
 

### 2024-09-29 paper.tex
Update §5.2.4

@@
 Figure \ref{fig:non_overlapping_coauthornetwork_authors_emailnetwork}.
-We observe that that are some prolific authors are that \emph{not} engaged
+We observe that some prolific authors are \emph{not} engaged
 in the email discussion: 40\%-50\% of non-overlapping authors are ranked
@@
 
-We conclude that the IETF is still reasonably centralised, but that this
-has improved over the years. However, removing around 20\%-25\% most
-influential participants still fragments the entire network
-(\S\ref{subsec:measuring_influence}, Figure
-\ref{fig:purge_nodes_largest_connected_component_number_components}). The
-most influential people are the ones with highest degree of engagements
-within the community who also tend to be more involved in draft authoring
-activities (\S\ref{subsec:influencer_behaviour}, Figure
+The IETF is a reasonably centralised organisation, that dominated by a
+relatively small group of influential participants, but this dominance
+is gradually reducing over time.
+However, removing around 20\%-25\% most influential participants still
+fragments the email graph (\S\ref{subsec:measuring_influence}, Figure
+\ref{fig:purge_nodes_largest_connected_component_number_components}).
+
+The most influential participants are those with the highest degree of
+engagement within the community. These also tend to be more involved in
+draft authoring (\S\ref{subsec:influencer_behaviour}, Figure
 \ref{fig:drafts_initiation_top_percentile}). However, we also observe that
-over the years more people (even the ones less influential in the email
-network) are getting involved in draft authoring activities (Figure
+over the years more people, even those less influential in the email
+network, are getting involved in draft authoring activities (Figure
 \ref{fig:drafts_initiation_top_percentile}). This shows that these
@@
 
-A considerable portion of influential participants ($\sim$30\%) are
-affiliated to one of the more prominent organisations (e.g., \emph{Cisco}
-or \emph{Ericsson}) (\S\ref{subsec:impact_influence}, Figure
-\ref{fig:coauthor_affiliation_network2019_otheryears}). Participants from
-different organisations do considerably collaborate (Figure
-\ref{fig:heatmap_2019_coauthordraft_most_frequent_org}). Moreover, the
-level of such collaboration is found to have increased with time.
+Finally, a considerable portion of influential participants ($\sim$30\%)
+are affiliated with one of the more prominent organisations, e.g., Cisco
+or Ericsson, (Figure \ref{fig:coauthor_affiliation_network2019_otheryears}).
+Participants from different organisations do considerably collaborate
+(Figure \ref{fig:heatmap_2019_coauthordraft_most_frequent_org}). Moreover,
+the level of such collaboration is found to have increased with time.
 

### 2024-09-29 paper.tex
Update §5.3 introduction

@@
 %--------------------------------------------------------------------------------------------------
-\subsection{Hierarchy and Communication Dynamics}
+\subsection{Hierarchy and Communication Patterns}
 \label{sec:org-dyn:hierarchy}
@@
 
+As a voluntary and consensus-driven \cite{RFC7282} standards development
+organisation, one might expect the communication patterns in the IETF to
+differ from those in businesses following a more traditional management
+structure. Those in IETF leadership rules, whether working group chairs
+or area directors, must lead by persuasion and example and have little in
+the way of coercive power compared to a typical manager and this can be
+expected to impact how they communicate.
+
 In the following, we quantitatively determine the effect that organisational
@@
 We study how the email communication is impacted by the hierarchical roles
-of individuals. According to the IETF mission statement~\cite{rfc3935}, the
+of individuals. According to the IETF mission statement~\cite{RFC3935}, the
 IETF is made up of volunteers who collaborate to develop consensus-based
@@
 
-In the following we consider the following research questions: 
+We consider the following questions: 
 \begin{enumerate}
   \item
-    Are higher levels of the hierarchy growing or becoming more
-    centralised? Are they associated with an increased domination
-    of the conversation?
+    Are higher levels of the hierarchy growing or becoming more centralised?
+    Are they associated with an increased domination of the conversation?
     (\S\ref{sec:org-dyn:hierarchy:rq1})
@@
     What is the association between the organisation hierarchy and general
-    communication patterns?
-    (\S\ref{sec:org-dyn:hierarchy:rq2})
+    communication patterns?  (\S\ref{sec:org-dyn:hierarchy:rq2})
 
   \item
-    How do people with differing roles communicate with each other?  Does
-    information flow ``up" or ``down" the hierarchy?
+    How do people with differing roles communicate with each other?
+    Does information flow up or down the hierarchy?
     (\S\ref{sec:org-dyn:hierarchy:rq3})
@@
 
-The first hypothesis focuses on the impact of the
-``steepness"~\cite{anderson2010functions} of the hierarchy. This refers to
-the fewer people in higher hierarchy levels the more steep the hierarchy
-is, i.e a centralised control of decision making in higher levels. The
-opposite of this is a ``diffused"~\cite{hussain2019voice} hierarchy where
-the large amount of people in higher levels stifle lower level voice by
-creating ``bystanders" who find it hard to speak up. In general, the
-conclusions about steep hierarchies are confused, where some papers
-conclude a positive and some a negative effect of steep hierarchies. We
-explore what this means for the IETF in RQ1.
-
-The second hypothesis regards the ``power
-distance"~\cite{li2021does,duan2018authoritarian,guo2020inclusive} of
+Question 1 relates to the \emph{steepness} of the hierarchy
+\cite{anderson2010functions}, that may exhibit a high amount of centralised
+decision making, compared to a more \emph{diffused} \cite{hussain2019voice}
+hierarchy where a larger amount of people in higher levels stifle lower
+level voice by creating ``bystanders'' who find it hard to speak up. In
+general, the conclusions about steep hierarchies are confused, where some
+papers conclude a positive and some a negative effect of steep hierarchies.
+We explore what this means for the IETF in \S\ref{sec:org-dyn:hierarchy:rq1}.
+
+Questions 2-4 relate to the \emph{power distance}
+\cite{li2021does,duan2018authoritarian,guo2020inclusive} of
 individuals, which is the number of hierarchy levels between people who are
@@
 large effect of reducing the voice of the lower of the two levels. We
-tackle this for the IETF in three different ways outlined in RQ2-4.
-
-The IETF being a voluntary organisation gives a new perspective on these
-hypotheses, as the existing literature focuses on commercial organisations.
-In contrast to the above conclusions, we demonstrate that despite the
-hierarchy becoming more ``diffused", participants in the higher levels in
-the IETF hierarchy perform a ``facilitator" role, promoting discussion with
-those they interact with and receiving a benefit themselves as a result. We
-also show that higher level participants tend to be a focus of discussion
-with remarks directed ``upward" toward them whereas Regular Participants
-(RPs) engage more in group-discussion with less focus on an individual.
-This suggests that the IETF does not conform to the notion that power
-distance has a negative effect on communication, nor that a more diffused
-hierarchy reduces the voice of lower levels.
+tackle this for the IETF in three different ways outlined in 
+\S\ref{sec:org-dyn:hierarchy:rq2} -- \S\ref{sec:org-dyn:hierarchy:rq4}.
+
+The voluntary nature of the IETF offers a new perspective on these
+questions.  We demonstrate that despite the hierarchy becoming more
+diffused, participants in the higher levels in the IETF hierarchy perform a
+facilitator role, promoting discussion with those they interact with and
+receiving a benefit themselves as a result. We also show that higher level
+participants tend to be a focus of discussion with remarks directed upward
+toward them whereas Regular Participants engage more in group-discussion
+with less focus on an individual.  This suggests that the IETF does not
+conform to the notion that power distance has a negative effect on
+communication, nor that a more diffused hierarchy reduces the voice of
+lower levels.
 

### 2024-09-29 paper.tex
Update §5.3.2

@@
 large effect of reducing the voice of the lower of the two levels. We
-tackle this for the IETF in three different ways outlined in 
+tackle this for the IETF in three different ways in
 \S\ref{sec:org-dyn:hierarchy:rq2} -- \S\ref{sec:org-dyn:hierarchy:rq4}.
@@
 
+\todo{Do we need a separate methodology section?}
+
 % - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
@@
 
-The effect the steepness of an organisation's hierarchy has on the
-communication within is not well understood. This RQ is important because
-once we know whether the IETF hierarchy is becoming steeper, we can look at
-the effects this is having on the communication patterns of participants.
-
-In this RQ, we look at the number of individuals and/or roles that inhabit
-each hierarchy level. We also consider whether individuals in higher levels
-are more active in the conversation and quantify the evolution of the
-mailing list activity within each level of the IETF hierarchy. We find that
-the middle level, Working Group Chairs (WGCs), is gaining in overall
-proportion of activity, whilst the lowest level, RPs, is decreasing. The
-top level, Area Directors (ADs), remains constant in proportion. The
-increase in WGC activity coincides with an increase of WGCs per Working
-Groups (WGs).
+% The effect the steepness of an organisation's hierarchy has on the
+% communication within is not well understood. This RQ is important because
+% once we know whether the IETF hierarchy is becoming steeper, we can look at
+% the effects this is having on the communication patterns of participants.
+
+In this section, we look at the number of individuals that inhabit each
+level of the IETF hierarchy and their communication patterns, and quantify
+the evolution of the mailing list activity within each level of the IETF.
+We find that the middle level, Working Group Chairs, is gaining in overall
+proportion of activity, whilst activity of participants at the lowest level
+is decreasing. The top level of the hierarchy, area directors, remains
+constant in proportion. The increase in working group chair activity
+coincides with an increase in the number of chairs per working group.
 
@@
 
-For \textbf{RQ1}, to determine the steepness or diffusion of the hierarchy,
-the proportion of nodes and activity are determined in each hierarchy
-level. Figure~\ref{fig:activity_proportions}a shows the number of active
-nodes in each level. The bottom level, RPs, are the largest contingent
-consisting of 90--86\% of active participants; this proportion is
-decreasing over time. The next level is WGCs which manage the WGs and are
-6--10\% of active participants; this proportion is rising over time. The
-top level is ADs which manage many WGs within an area and consist of less
-than 1\% of the total active population of the IETF; this stays constant
-over time. Therefore, the WGC level is becoming more diffused over time as
-there are about 15 RPs for every one WGC in the organisation in 2013 versus
-8 RPs per WGC in 2021.
-
-The proportion of activity of nodes in the last year is also plotted and
-split into hierarchy level in Figure~\ref{fig:activity_proportions}b. It is
-clear that individuals at higher levels contribute a disproportionate share
-to communication and this share is increasing over time, which is
-consistent with findings in \cite{khare2022web}. Therefore, less of a
-proportion of communication is flowing from the RPs at the same time as the
-number of WGC roles is increasing. This further suggests that the IETF
-hierarchy is becoming more diffused, in terms of mailing list
-communication, which may make cooperation between hierarchy levels more
-difficult.
+\pb{Role of Working Group Chairs:}
+Figure \ref{fig:activity_proportions}(a) shows the number of participants
+at each level of the hierarchy in IETF: Regular participants are the
+largest contingent, comprising 90--86\% of active participants; this
+proportion is decreasing over time.  The next level is Working Group
+Chairs, comprising  6--10\% of active participants; this proportion is
+rising and becoming more diffused over time, with about 15 regular
+participants for every working group chair in the organisation in 2013
+versus 8 participants per chair in 2021.  The top level is Area Directors
+who consist of less than 1\% of the total active population of the IETF;
+this stays constant over time.
+
+The fraction of emails sent by participants at each level of the hierarchy
+is shown in Figure~\ref{fig:activity_proportions}(b). Individuals at higher
+levels contribute a disproportionate share to communication and this share
+is increasing over time, consistent with \S\ref{sec:org-dyn:influence}. A
+smaller proportion of email are sent by regular participants at the same
+time as the number of working group chairs is increasing, further
+suggesting that the IETF hierarchy is becoming more diffused, in terms of
+mailing list communication, which may make cooperation between hierarchy
+levels more difficult.
 
@@
   \caption{
-    Proportions of RPs, WGCs and ADs by a) number of active individuals and
-    b) activity (number of emails sent to mailing lists). 
+    Proportion of regular participants, working group chairs, and area
+    directors by (a) number of active individuals and (b) mailing list
+    activity.
   }
@@
 
-Figure~\ref{fig:wgc_over_time}a shows the amount of nodes that inhabit the
-WGC hierarchy level over our time period. There is a $35\%$ rise in the
-amount of WGC roles whereas the amount of individuals that fill the WGC
-roles remains mostly constant.
-
-Figure~\ref{fig:wgc_over_time}b contains two estimates of the number of WGs
-over time. The estimated number of WGs is likely the most accurate in the
-middle 50\% of both plotted lines. Their discrepancy is the least (at most
-10 WGs) from late 2014 to early 2019. The start of the COVID-19 pandemic
-may also affect mailing list activity in 2020, and similar trends are
-visible in the data in \cite{mcquistin2021characterising}. A full
-explanation on how the data is estimated can be found in
-section~\ref{sec:WG_activity}.
-
-Figure~\ref{fig:wgc_over_time}c shows how many WGCs exist per WG, using
-both estimates of the WG amount from Figure~\ref{fig:wgc_over_time}b. It is
-clear there is a slow a rise in WGCs per WG over time. The ratio rises from
+%..................................................................................................
+\pb{Number of Working Group Chairs:}
+Figure~\ref{fig:wgc_over_time}(a) shows the number of working group chairs
+our time. There is a $35\%$ rise in the number of chairs while the amount
+of individuals that fill the chair roles remains mostly constant.
+
+Figure~\ref{fig:wgc_over_time}(b) contains two estimates of the number of
+working groups over time, based on tracking group events (i.e., Datatracker
+metadata) and mailing list activity since groups may be recorded as being
+active but have no activity, and vice-versa.  The discrepancy between the
+two estimates is the least (at most 10 working groups) from late 2014 to
+early 2019.
+
+Figure~\ref{fig:wgc_over_time}(c) shows how many chairs exist per working
+group, using both estimates of the number of working groups from
+Figure~\ref{fig:wgc_over_time}(b). It is clear there is a slow a rise in
+the number of chairs per working group over time. The ratio rises from
 between 1.7-1.9 to 2.0-2.2 within the middle 50\%, late 2014 to early
 2019.This perhaps represents a growing understanding in the IETF that it is
-desirable to have two or more chairs per WG in case of
+desirable to have two or more chairs per working group in case of
 illness/unavailability or to avoid conflicts of interest. Therefore, this
 plot again shows that the hierarchy is becoming more diffused, even if the
-raw amount of individuals in the WGC roles has remained constant. The
-multiple roles per individual may result in even more difficulty of RPs to
-use their voice.
-
-An AD version of Figure~\ref{fig:wgc_over_time} is available, but it is
-omitted to save space as all three plots stay mostly constant throughout
-the period. There are about 15 individual ADs split amongst 15 roles in 7-8
-areas with about 2 ADs per Area. The set of IETF Areas is broadly fixed,
-while WGs are created and closed relatively frequently.
+raw amount of individuals in the chair roles has remained constant. The
+multiple roles per individual may result in even more difficulty of regular
+participants to use their voice.
 
@@
   \caption{
-    Overview of WGC numbers and WG numbers over time. a) shows the number
-    of WGC roles (some individuals hold multiple WGC roles) and the number
-    of individuals who are WGCs b) shows two different estimates of the
-    number of WGs (see section~\ref{sec:method}) and c) shows the mean
-    number of WGCs per WG. Notice the non-zero-based y-axis exaggerates the
-    variation in these figures.
+    Changes in the number of working groups and working group chairs over
+    time: (a) shows the number of chairs and the number of individuals who
+    are chairs; (b) shows two different estimates of the number of working
+    groups; and c) shows the mean number of chairs per working group.
+    Notice the non-zero-based y-axis exaggerates the variation in these
+    figures.
   }
@@
 
-An AD version of Figure~\ref{fig:wgc_over_time} is available, but it is
-omitted to save space as all three plots stay mostly constant throughout
-the period. There are about 15 individual ADs split amongst 15 roles in 7-8
-areas with about 2 ADs per Area. The set of IETF Areas is broadly fixed,
-while WGs are created and closed relatively frequently.
-
-Table~\ref{tab:WGC_email} shows the emails sent and received for WGC and
-ADs on mailing lists in the year before they first take that role, and the
-year after. WGCs and ADs are both active in using the mailing lists, ADs
-much more so. However, they don't become more active to any major degree
-after taking the roles. This may suggest that the individuals who take on
-higher hierarchy level roles are already contributing at a higher level.
+%..................................................................................................
+\pb{Number of Area Directors:}
+A version of Figure~\ref{fig:wgc_over_time} considering area directors
+rather than working group chairs is available, but omitted to save space as
+all three plots stay mostly constant throughout the period. There are about
+15 area directors split amongst 15 roles in 7-8 areas with typically 2 area
+directors per Area. The set of IETF Areas changes slowly, while working
+groups are created and closed relatively frequently.
+
+%..................................................................................................
+\pb{Activity prior to appointment:}
+Table~\ref{tab:WGC_email} shows the amount of email sent and received for
+working group chairs and area directors on mailing lists in the year before
+they first take that role, and the year after. Chairs and area directors
+are both active in using the mailing lists, area directors much more so.
+However, they don't become more active to any major degree after taking the
+roles. This may suggest that the individuals who take on higher hierarchy
+level roles are already contributing at a higher level.
 
@@
 
+%..................................................................................................
 \pb{Summary}
-In answer to RQ1 the picture is mixed, the number of individuals with WGC
-roles has remained broadly constant but the number of WGC roles has
-increased, indicating that individuals willing to become WGC have taken on
-more such roles. The number of the higher level AD roles has, by design,
-remained broadly constant over the time studied. The proportion of active
-individuals with WGC roles has grown from 6\% to 10\% and the proportion of
-the conversation taken by those roles has grown from 25\% to 35\% over the
+The answer to the question of whether the higher levels of the hierarchy
+growing or becoming more centralised, and whether they're associated with
+an increased domination of the conversation, is mixed.  The number of
+individuals with working group chair roles has remained broadly constant
+but the number of such roles has increased, indicating that individuals
+willing to become chairs have taken on more such roles.  The number of the
+higher level area director roles has, by design, remained broadly constant
+over the time studied. The proportion of active individuals with working
+group chairs roles has grown from 6\% to 10\% and the proportion of the
+conversation taken by those roles has grown from 25\% to 35\% over the
 period studied. Therefore the hierarchy has become more diffused, which may
-mean RPs experience more difficulty in expressing their voice.
+mean regular participants experience more difficulty in expressing their
+voice.
 

### 2024-10-02 paper.tex
Update §5.3.2

@@
 
-The communication patterns that different levels in the hierarchy
-experience may be  indicative of the cooperation between
-levels~\cite{li2021does,duan2018authoritarian}. For instance, if
-participants at higher levels send lots of email while receiving little,
-this suggests a lack of receptiveness of higher levels to the suggestions
-of lower levels, and low confidence of lower levels to voice their
-opinions. In contrast, we hypothesise that collaboration between layers is
-high, as outlined in the IETF mission statement, which will look more like
-the opposite of this example.
-
-In this vein, we compare the ways that different roles communicate, looking
-at the tendency for communications to be purely inbound, outbound or use
-more complex patterns. We count three edge motifs for different hierarchy
-levels as described in section~\ref{sec:motif}. We find that higher levels
-consistently experience a higher proportion of incoming communication than
-RPs, and all levels send outward communication at similar proportions over
-time. Also, the proportion of people who begin email threads are split by
-hierarchy level. Higher levels show a disproportionate tendency to
-originate email threads.
+In this section, we explore communication patterns within the IETF, to
+understand how participants in different roles (regular participants,
+working group chairs, area directors) communicate and whether those
+patterns of communication are indicative of healthy cooperation in the
+running of the organisation \cite{li2021does,duan2018authoritarian}.
+
+For instance, if those in leadership positions send lots of email while
+receiving little, this suggests a lack of receptiveness of the leadership
+to the suggestions from regular participants, and low confidence of those
+regular participants to voice their opinions. In contrast, we hypothesise
+that if the IETF operates as outlined in its mission statement, in a more
+cooperative manner, then communication between those in different roles
+will be more symmetric.
+
+\pb{Temporal Motifs:}
+We compare the ways that people in different roles communicate, looking
+at the tendency for communications to be purely inbound, purely outbound,
+or to follow more complex patterns. 
+%
+% From our ICWSM 2024 paper, Section 2:
+%
+We describe these in terms of \emph{temporal motifs}: patterns in the
+sequence in which edges are added to the communication graph over time.
+Specifically, we analyse every possible combination of three temporally
+ordered links~\cite{paranjape2017motifs} between at most three nodes.
+The time order in which the nodes occur is important.  The direction of
+edge 1 is arbitrary, but every edge temporally after it has their direction
+oriented with respect to it.
+
+Figure \ref{fig:motif_types} shows all possible types of interaction and
+classifies them into five types: motifs with two nodes; ``Outward Star",
+announcements or dissemination of information; ``Inward Star", questions or
+condensing of information; ``Mixed Star", one-on-one discussion; and
+``Triangle" is group discussion with no individual as the focus. Motifs
+with only two nodes  are rare in this network and we do not present the
+results.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/motifs.pdf}
+  \caption{
+    Classification of three edge motif types with the numbers representing
+    the order of communication.  Light blue squares are  ``Outward Star''
+    motifs; purple are ``Inward Star'' motifs; orange are ``Mixed Star''
+    motifs and grey are ``Triangle'' motifs. The box in the bottom left hand
+    corner represents motifs with only two nodes.
+  }
+  \label{fig:motif_types}
+\end{figure}
+
+Raphtory~\cite{steer2020raphtory} is used to count the prevalence of these
+motifs. We start with the list of 36 different combinations of three
+directed edges shown in figure~\ref{fig:motif_types}. The graph is then
+split into time windows and then the motifs are counted combinatorially for
+each window; each of a node's edges are followed from their origin to three
+edge steps away. Finally, we combine the motif counts using the categories
+shown in figure~\ref{fig:motif_types} by the colour of the squares.
+
+% We find that those in leadership positions (working group chairs, area
+% directors) consistently experience a higher proportion of incoming
+% communication than regular participants, and all levels send outward
+% communication at similar proportions over time. Also, the proportion of
+% people who begin email threads are split by hierarchy level. Higher levels
+% show a disproportionate tendency to originate email threads.
 
@@
 
-To quantify the effect of being in each hierarchy level has on network wide
-communication for \textbf{RQ2}, we calculate the proportion of three node
+\pb{Impact of Role on Communication Patterns:}
+To quantify the effect of being in a leadership position has on network
+wide communication, we calculate the proportion of three node temporal
 motifs, lasting at most a month, in which each node participates over a
-year period, see section~\ref{sec:motif}. These motifs are counted,
-categorised into the three hierarchy levels and then proportions are taken
-of each motif type.
+year period. These temporal motifs are counted, categorised into the three
+hierarchy levels (regular participant, working group chair, area director)
+and then proportions are taken of each temporal motif type.
 
@@
 ``Triangles". The Inward and Mixed Star motifs show about a $4\%$ and
-$10\%$ increase in proportion for higher levels in the hierarchy versus for
-RPs. However, about a $10\%$ increase for RPs is seen for Triangles versus
-WGCs and ADs, and Outward Star proportions remain similar for all levels of
+$10\%$ increase in proportion for higher levels in the hierarchy versus
+regular participants. However, about a $10\%$ increase for regular
+participants is seen for Triangles versus working group chairs and area
+directors, and Outward Star proportions remain similar for all levels of
 the hierarchy.
 
-The larger proportion for WGCs and ADs of Inward and Mixed Star motifs
-whilst the Outward Star motifs remain similar for all levels suggests that
-higher levels receive more direct communication. The higher proportion of
-RP Triangle motifs suggests their discussion is more of a group activity,
-whereas WGCs and ADs have a larger proportion of one-on-one conversations.
-Therefore, we interpret the general communication patterns within the IETF
-as a discussion amongst RPs interspersed with questions sent to and
-announcements from WGCs and ADs. This indicates some collaboration between
-layers and confidence of some RPs' to voice their opinions upwards.
+The larger proportion of those in leadership roles in Inward and Mixed Star
+temporal motifs, whilst Outward Star motifs remain similar for all levels,
+suggests that those in leadership receive more direct communication. The
+higher proportion of Triangle motifs for regular participants suggests
+their discussion is more of a group activity, whereas working group chairs
+and area directors have a larger proportion of one-on-one conversations.  
+We therefore interpret the general communication patterns within the IETF
+as a discussion amongst participants interspersed with questions sent to,
+and announcements from, the working group chairs and area directors.  This
+indicates some collaboration between layers and confidence of some regular
+participants' to voice their opinions upwards.
 
@@
 originate, using a year window, pushed forward by a month each calculation.
-We see that WGCs and ADs send a disproportionate number of originating
-emails to the mailing lists. WGCs send 30--50\% whereas they are 6--10\% of
-active individuals, and ADs 5--10\% versus 0.5\%.
-
-Therefore, the larger proportion of inward motifs for higher levels is in
-part due to WGCs and ADs disproportionately originating email threads.
+We see that working group chairs and area directors send a disproportionate
+number of originating emails to the mailing lists (working group chairs
+send 30--50\% whereas they are 6--10\% of active individuals, and area
+directors 5--10\% versus 0.5\%). 
+Therefore, the larger proportion of inward motifs for those in leadership
+roles is in part due to their disproportionately originating email threads.
 Other IETF participants will then reply to the thread, boosting the
-proportion of inward motifs WGCs and ADs appear in. This may suggest the
-higher levels are good ``condensers of communication" as they receive more
-than they send.
-
-In answer to the RQ, combining the interpretations of the two plots, the
-discussion seems to happen in the following way: The ADs or WGCs are
-commonly the originators of discussion threads. Where conversation includes
-a WGC or AD the conversation tends to be directed around them and usually
-inward towards the WGC or AD. The RPs are more likely to engage in
-discussions amongst themselves (triangular communication patterns) when
-higher level individuals are not present.
+proportion of inward motifs working group chairs and area directors appear
+in. This may suggest the higher levels are good ``condensers of
+communication" as they receive more than they send.
 
@@
 \end{figure}
+
+\pb{Summary:}
+Combining the interpretations of Figures \ref{fig:motif_proportions} and
+\ref{fig:origin_emails}, the discussion seems to happen in the following
+way: area directors or working group chairs are commonly the originators
+of discussion threads. Where conversation includes a working group chair
+or area director the conversation tends to be directed around them and
+usually inward towards the working group chair or area director. The
+regular participants are more likely to engage in discussions amongst
+themselves (triangular communication patterns) when higher level
+individuals are not present.
 

### 2024-10-02 paper.tex
Remove unneeded methodology section
Update 5.3.3

@@
 % - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-\subsubsection{Methodology}
-\label{sec:org-dyn:hierarchy:methodology}
-
-% From our ICWSM 2024 paper, section 4:
-
-\todo{Do we need a separate methodology section?}
-
+% \subsubsection{Methodology}
+% \label{sec:org-dyn:hierarchy:methodology}
+% 
+% % From our ICWSM 2024 paper, section 4:
+% 
+% \todo{Do we need a separate methodology section?}
+% 
 % - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
@@
   \caption{
-    Proportion of ``origin" (non reply) emails sent by RPs, WGCs and ADs,
-    this should be interpreted in conjunction with
-    figure~\ref{fig:activity_proportions} showing the proportion of active
-    individuals in each class.
+    Proportion of ``origin" (non reply) emails sent by regular
+    participants, working group chairs, and area directors, this should be
+    interpreted in conjunction with Figure~\ref{fig:activity_proportions}
+    showing the proportion of active individuals in each class.
   }
@@
 % - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-\subsubsection{Communication up and down the Hierarchy}
+\subsubsection{Communication Up and Down the Hierarchy}
 \label{sec:org-dyn:hierarchy:rq3}
@@
 evident in the effect that power distance~\cite{li2021does} has on
-communication throughout the organisation. The raw proportion of the
-communication between hierarchy levels helps us to understand the
-confidence of lower levels to exercise their voice. This proportion, for a
-cooperative and voluntary organisation should be higher for lower hierarchy
-levels, showing a confidence in their voice.
-
-We categorise edges based on the hierarchy levels they originate from and
-are sent to. Ratios are then calculated of ``upward" versus ``downward"
-communication between RPs, WGCs and ADs. We find that the lower the level,
-the more of a skew to upward communication.  Communication tends to flow
-``up'' the hierarchy more than ``down''.
+communication. The proportion of the communication between hierarchy levels
+helps us to understand the confidence of lower levels to exercise their
+voice: that is, the willingness of regular participants to challenge the
+working group chairs and of the working group chairs to challenge area
+directors.  This proportion, in a cooperative and voluntary organisation,
+should be higher for those in lower positions in the hierarchy, showing a
+confidence in their voice.
+
+% We categorise edges based on the hierarchy levels they originate from and
+% are sent to. Ratios are then calculated of ``upward'' versus ``downward''
+% communication between regular participants, working group chairs, and area
+% directors. We find that the lower the level, the more of a skew to upward
+% communication.  Communication tends to flow ``up'' the hierarchy more than
+% ``down''.
 
@@
 
-For \textbf{RQ3}, we determine the direction communication tends to flow
-through the hierarchy by looking at inbound and outbound communication
-flows. The network's edges are categorised based on their source and target
-node's hierarchy level. For each of the inter-level communication
-combinations the proportion of communication in the last year going in the
-``upward" hierarchy direction is calculated. For instance, if there are six
-emails from RPs to WGCs and twelve from WGCs to RPs then the proportion
-upward is $\frac{6}{6+12}=\frac{1}{3}$.
+We determine the direction communication tends to flow through the
+hierarchy by looking at inbound and outbound communication flows. The
+network's edges are categorised based on their source and target node's
+hierarchy level. For each of the inter-level communication combinations the
+proportion of communication in the last year going in the ``upward''
+hierarchy direction is calculated. For instance, if there are six emails
+from regular participants to working group chairs, and twelve from working
+group chairs to regular participants, then the proportion upward is
+$\frac{6}{6+12}=\frac{1}{3}$.
 
@@
 communication flowing up the hierarchy than down. All three levels are
-mostly upward in direction, with the lowest two levels (RP$\rightarrow$
-WGC) having the largest skew in the proportion. The other two lines
-(RP$\rightarrow$AD and WGC$\rightarrow$AD) show similar periodicity, with
-the WGC$\rightarrow$AD line dropping as far below $0.5$ as above showing a
-large shift in inter-level communication patterns higher up the hierarchy.
+mostly upward in direction, with the lowest two levels (regular participant
+$\rightarrow$ working group chair) having the largest skew in the
+proportion. The other two lines (regular participant $\rightarrow$ area
+director; working group chair $\rightarrow$ area director) show similar
+periodicity, with the working group chairs $\rightarrow$AD line dropping as
+far below $0.5$ as above showing a large shift in inter-level communication
+patterns higher up the hierarchy.
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/up_hierarchy_communication_ratios.png}
+  \caption{
+    Proportion of communications that are ``up'' the hierarchy for different
+    groups. A proportion above 0.5 indicates that the majority of
+    communications are ``upward''.
+  }
+  \label{fig:inter_strata}
+\end{figure}
 
@@
 to bolster our interpretation that the cooperative design of the IETF is
-working. This analysis shows that RPs are communicating much more
-preferentially upwards in the hierarchy, despite the decrease in overall
-share of RP activity. The increase within the WGC level does not have any
-noticeable effect on their level of communication, if anything there is a
-negative relationship. This again suggests both WGCs and ADs perform a kind
-of ``condenser of communication" or ``facilitator" role. In other words,
-this analysis suggests the IETF has a ``bottom-up" communication style,
+working. This analysis shows that regular participants are communicating
+much more preferentially upwards in the hierarchy, despite the decrease in
+overall share of their activity. The increase within the working group
+chair level does not have any noticeable effect on their level of
+communication, if anything there is a negative relationship. This again
+suggests both working group chairs and area directors perform a kind of
+``condenser of communication'' or ``facilitator'' role. In other words,
+this analysis suggests the IETF has a ``bottom-up'' communication style,
 where the higher levels encourage the lower levels' voice.  This is the
@@
 
-Overall for RQ3 then, the pattern observed was communication flowing up the
-hierarchy and this corresponds to the answer found in RQ2. WGCs and ADs
-were more likely to receive communication with individuals than send
-communication to individuals whereas both were more likely than RPs to send
-out messages to the list in general.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/up_hierarchy_communication_ratios.png}
-  \caption{
-    Proportion of communications that are ``up" the hierarchy for different
-    groups. A proportion above 0.5 indicates that the majority of
-    communications are ``upward".
-  }
-  \label{fig:inter_strata}
-\end{figure}
+\pb{Summary:}
+The pattern observed was communication flowing up the hierarchy, broadly
+matching that observed in \S\ref{sec:org-dyn:hierarchy:rq2}.  Working group
+chairs and area directors were more likely to receive communication with
+individuals than send communication to individuals whereas both were more
+likely than regular participants to send out messages to the list in
+general.
 

### 2024-10-02 paper.tex
Update 5.3.5

@@
 
-An important question~\cite{li2021does,guo2020inclusive} for any
-organisation is whether higher level participants encourage those at lower
-levels in the hierarchy to properly engage. The IETF is cooperative and
+An important question for any organisation is whether higher level
+participants encourage those at lower levels in the hierarchy to properly
+engage \cite{li2021does,guo2020inclusive}. The IETF is cooperative and
 voluntary by design, therefore we hypothesise a high level of encouragement
@@
 
-In particular, we perform Mobility Taxonomy analysis to determine the
-effect individual IETF participants have on their future number of
-connections, and the average number of connections of their neighbours.  We
-find that WGCs have an increased tendency, versus RPs, to remain active
-(\textit{Mobility}), an increased tendency for their neighbours to be
-active subsequent to high WGC activity (\textit{Philanthropy}) and for them
-to gain activity after their neighbours are active (\textit{Community}).
-We interpret this as WGCs filling a ``facilitator" role on the mailing
-lists.
+% In particular, we perform Mobility Taxonomy analysis to determine the
+% effect individual IETF participants have on their future number of
+% connections, and the average number of connections of their neighbours.  We
+% find that working group chairs have an increased tendency, versus regular
+% participants, to remain active (\textit{Mobility}), an increased tendency
+% for their neighbours to be active subsequent to high working group chair
+% activity (\textit{Philanthropy}) and for them to gain activity after their
+% neighbours are active (\textit{Community}).  We interpret this as working
+% group chairs filling a ``facilitator'' role on the mailing lists.
+
+%..................................................................................................
+% From our ICWSM 2024 paper, Section 4.3:
+
+\pb{Mobility Taxonomy:}
+We aim to measure the association of the activity of nodes, and their
+neighbours, with their activity in a subsequent time period. To determine
+this association we use the temporal graph analysis technique called the
+Mobility Taxonomy~\cite{barnes2023measuring}, which involves correlating
+the degree and average neighbourhood degree (ND) between two adjoining time
+windows, both half of the chosen time window length for other analysis. The
+temporal graph analysis tool Raphtory~\cite{steer2020raphtory} is used to
+calculate the raw degree and ND numbers before performing these
+correlations.
+
+There are six combinations to correlate node degree and ND in time window
+one and two, the combinations are outlined in detail
+in~\cite{barnes2023measuring}. The four measures used in this paper are
+\textit{Mobility}, \textit{Neighbour Mobility}, \textit{Philanthropy} and
+\textit{Community}.  Each represents the correlation of the degree of a
+node (or set of nodes) at an earlier or later period of time.
+\textit{Mobility} can be thought of as the tendency for a node that is
+active in the first time window to be equally as active in the second.
+\textit{Neighbour Mobility} is similar but for a node's neighbourhood.
+\textit{Philanthropy} can be thought of as the tendency of a node's
+neighbourhood to be active in window two if that node is active in window
+one. Finally, \textit{Community} is the reverse of \textit{Philanthropy} as
+the activity of a node in window two is compared to that node's neighbours
+activity in window one.
+
+Only the nodes which are active in the first window are considered for the
+second window. Nodes inactive in the second time window are said to have a
+degree of 0. The ND in the second time window is taken over the same set of
+neighbours that existed in the first; that is we look at a node's
+neighbours in the first window and measure their degree in the second for
+consistency of comparison.
 
@@
 
-Finally, to determine how individuals and their neighbours affect each
-other's activity levels in subsequent time periods, for \textbf{RQ4}, we
-take correlations of degree between different time windows using the
-Mobility Taxonomy~\cite{barnes2023measuring} aspects called
-\textit{Mobility}, \textit{Neighbour Mobility}, \textit{Philanthropy} and
-\textit{Community} explained in the previous section. In
+\pb{Participant Mobility:}
+To determine how individuals and their neighbours affect each other's
+activity levels in subsequent time periods, we take correlations of degree
+between different time windows using this Mobility Taxonomy. 
 Figure~\ref{fig:mob_tax}, the time window chosen is one year which is split
@@
 one month and the process is repeated. The mid point of the year time
-window is plotted. WGCs and RPs are plotted but the small number of ADs
-mean that the correlations are too noisy to show meaningful effects.
+window is plotted. Working group chairs and regular participants are
+plotted but the small number of area directors mean that the correlations
+are too noisy to show meaningful effects.
 
@@
   \caption{
-    Connection between roles and activity for RP and WGC viewed in terms of
-    a) Mobility b) Neighbour Mobility c) Philanthropy and d) Community over
-    time. See section~\ref{subsec:mob_tax} for definitions of these terms.
+    Connection between roles and activity for regular participants and
+    working group chairs viewed in terms of a) Mobility b) Neighbour
+    Mobility c) Philanthropy and d) Community over time.
   }
@@
 
-The \textit{Mobility} of both RPs and WGCs is plotted in
-Figure~\ref{fig:mob_tax}a. The high amount of correlation between the
-degree of an individual in the first half of the time window with the
-second half shows that all users have a tendency to become more active in
-subsequent time periods if they are highly active in the previous period,
-but this is more pronounced in WGCs. This is found to follow a period of
-about one year for both RPs and WGC, suggesting that a person is likely to
-be as active as there were a year ago. The \textit{Neighbour Mobility}
-plot, Figure~\ref{fig:mob_tax}b, shows a similar tendency for
-neighbourhoods, but with less of a split.
-
-The \textit{Philanthropy} plot, Figure~\ref{fig:mob_tax}c, shows a higher
-correlation for WGCs than RPs between an individual's degree in the first
-snapshot with and their ND in the second. This means that individuals who
-directly interact with WGCs become more active on mailing lists
-subsequently. The \textit{Community} plot, Figure~\ref{fig:mob_tax}d, shows
-the correlation between the ND in the first snapshot with the degree in the
-second.  The higher correlation for WGCs suggests they benefit from
-interactions with highly active neighbours by themselves becoming more
-active in subsequent time periods. These trends for \textit{Philanthropy}
-and \textit{Community} show a reversal in early 2016 that coincides with a
-period where the mean number of WGCs per WG and the number of WGC roles was
-rapidly rising, see figure~\ref{fig:wgc_over_time}. A merging of the
+The \textit{Mobility} of regular participants and working group chairs is
+plotted in Figure~\ref{fig:mob_tax}(a). The high amount of correlation
+between the degree of an individual in the first half of the time window
+with the second half shows that all users have a tendency to become more
+active in subsequent time periods if they are highly active in the previous
+period, but this is more pronounced in working group chairs. This is found
+to follow a period of about one year for both regular participants and
+working group chairs, suggesting that a person is likely to be as active as
+there were a year ago. The \textit{Neighbour Mobility} plot,
+Figure~\ref{fig:mob_tax}(b), shows a similar tendency for neighbourhoods, but
+with less of a split.
+
+The \textit{Philanthropy} plot, Figure~\ref{fig:mob_tax}(c), shows a higher
+correlation for working group chairs than regular participants between an
+individual's degree in the first snapshot with and their ND in the second.
+This means that individuals who directly interact with working group chairs
+become more active on mailing lists subsequently. The \textit{Community}
+plot, Figure~\ref{fig:mob_tax}(d), shows the correlation between the ND in
+the first snapshot with the degree in the second.  The higher correlation
+for working group chairs suggests they benefit from interactions with
+highly active neighbours by themselves becoming more active in subsequent
+time periods. These trends for \textit{Philanthropy} and \textit{Community}
+show a reversal in early 2016 that coincides with a period where the mean
+number of chairs per working group, and the number of working group chairs,
+rapidly rising, see Figure~\ref{fig:wgc_over_time}. A merging of the
 Applications area and Real-time Applications \& Infrastructure area in May
 2015 may be a cause of this reversal of correlation, as the assignment of
-WGs to areas was changed which may cause more non-reciprocal WGC
+working group chairs to areas was changed which may cause more non-reciprocal 
 communication than normal.
 
-Increased \textit{Mobility} for WGCs suggests that they gain advantage from
-their role with their activity leading to similar communication activity in
-the future. Moreover, they experience an increased effect of
-\textit{Philanthropy} and \textit{Community}. The former suggests that if a
-WGC is engaged in a higher amount of discussion, then in the future the
-individuals who they were communicating with will also be highly active.
-The latter suggests that when WGCs are surrounded by people who are active
-in discussion, they are likely to be more active themselves in the future
-too. All of this has less of an effect for RPs who show a lower correlation
-in terms of all three aspects. Our interpretation of this is that active
-WGCs are ``facilitating" discussion in the WG mailing lists; both having
-their activity boosted by their neighbours and, in turn, boosting the
-activities of their neighbours. This aligns well with the hypothesis that
-the cooperative and voluntary design of the IETF lends itself well to
-higher levels encouraging those lower to engage in discussion.
-
-Overall the answer to RQ4 is that the WGCs have a positive effect on their
-immediate contacts and those individuals that directly discuss with WGCs
-are more likely to engage more in discussion in subsequent time periods.
-Conversely though the WGCs who discuss topics with individuals who are
+\pb{Summary:}
+Increased \textit{Mobility} for working group chairs suggests that they
+gain advantage from their role with their activity leading to similar
+communication activity in the future. Moreover, they experience an
+increased effect of \textit{Philanthropy} and \textit{Community}. The
+former suggests that if a working group chair is engaged in a higher amount
+of discussion, then in the future the individuals who they were
+communicating with will also be highly active.  The latter suggests that
+when working group chairs are surrounded by people who are active in
+discussion, they are likely to be more active themselves in the future too.
+All of this has less of an effect for regular participants who show a lower
+correlation in terms of all three aspects. Our interpretation of this is
+that active working group chairs are ``facilitating'' discussion in the WG
+mailing lists; both having their activity boosted by their neighbours and,
+in turn, boosting the activities of their neighbours. This aligns well with
+the hypothesis that the cooperative and voluntary design of the IETF lends
+itself well to higher levels encouraging those lower to engage in
+discussion.
+
+Working group chairs have a positive effect on their immediate contacts.
+Those individuals that directly discuss with working group chairs are more
+likely to engage more in discussion in subsequent time periods.  Conversely
+though the working group chairs who discuss topics with individuals who are
 active in discussion are, themselves, more likely to engage more fully in
-discussion. This points to something like a virtuous cycle of WGCs
+discussion. This points to something like a virtuous cycle of chairs
 encouraging and being encouraged by their direct contacts.

### 2024-10-02 paper.tex
Add summary for §5.3

@@
 
-\todo{summarise hierarchy results}
+These findings presented in \S\ref{sec:org-dyn:hierarchy:rq2} --
+\S\ref{sec:org-dyn:hierarchy:rq4} suggest that the negative association of
+power distance is not as pronounced for the IETF as the literature finds
+for traditional commercial organisations. In fact, the facilitation by
+working group chairs of regular participants suggests the high levels
+actively encourage lower level participation in discussion. 
+Moreover, combining this with the finding that the organisational hierarchy
+has a diffused structure (\S\ref{sec:org-dyn:hierarchy:rq1}), with the
+potential problems this may cause for lower levels, suggests that
+collaboration between hierarchy levels is high within the IETF.
+
+As the IETF is voluntary and collaborative in its mission, this suggests
+that the their process for tackling the problems caused by a diffused
+hierarchy and power distance have worked well. One suggested improvement
+for the IETF is to encourage working group chairs to get involved in more
+group discussion in the mailing lists. This may help to boost their level
+of \textit{Philanthropy}, \textit{Community} and Triangle motifs, which are
+indicators of healthy discussion. However, we do not advocate for our
+analysis techniques to be the only metrics maximised for.  Moreover, these
+conclusions come from analysis of only communication structure not content.
+A future look into content may elucidate differing relationships between
+hierarchy levels.
 

### 2024-10-03 paper.tex
Add content from Tracing Linguistic Markers of Influence... (ACL 2023)

@@
 \documentclass[twocolumn,10pt]{article}
+\pdfoutput=1 % This must be in the first 5 lines to tell arXiv to use pdfLaTeX
 \usepackage[l2tabu,orthodox]{nag}
@@
 \usepackage[cm]{fullpage}
+\usepackage{multirow}
 
@@
 \newcommand{\pb}[1]{\vspace{0.75ex}\noindent{\textbf{#1}}}
+\newcommand{\liwc}[1]{\textsc{#1}}
+\newcommand{\word}[1]{\textit{`#1'}}
 
@@
         Prashant Khare \and
+        Ravi Shekhar \and
         Matthew Russell Barnes \and
@@
 
-% Tracing Linguistic Markers
-% - RQ1: How do linguistic traits differ between more and less influential
-%   participants? 
-% - RQ2: How do linguistic traits vary for participants at different levels
-%   of the organisation hierarchy? 
-% - RQ3: How does linguistic behaviour of participants change as they gain
-%   influence?
+Moving on from discussing the people involved in the IETF (\S\ref{sec:trends-demographics})
+and their interactions (\S\ref{sec:org-dyn}), we now consider communications
+within the IETF from the perspective of language to explore whether there
+are differences in linguistic patterns and language use in more influential
+participants, and to consider how the use of language reflects the
+consensus-driven nature of the standards development process.
+
+% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+% The following is adapted from our ACL 2023 Tracing Linguistic Markers of
+% Influence paper, section 2.
+
+\pb{Linguistic Inquiry and Word Count} (LIWC) \cite{pennebaker2015development}
+is a well-recognised psycholinguistic lexicon; it provides word counts for
+85 different linguistic, psychological, personal concern, and informal
+language marker categories. We generate a \emph{LIWC representation} of
+the emails sent by a participant in the analysis that follows. This is
+calculated by aggregating the word counts within each linguistic category
+for each participant using the LIWC 2015 dictionary (academic license).
+We filter out 104 ambiguous words that are present in LIWC but have
+technology, security, and network context meaning in IETF, using manually
+curated lists, for e.g., attack, argument, secure etc.,
+
+We also normalise by the total number of emails sent by that participant. 
+Such a normalisation is more appropriate here than normalising by total
+number of words written, as many IETF emails include long technical
+sections.  This generates a representation of a participant as their mean
+usage of each LIWC category; while this is a relatively reduced,
+low-dimensional representation of a person's language, it has the advantage
+of being interpretable and psychologically well-motivated.
+
+A limitation is that we used the standard LIWC-based analysis approach,
+which is purely lexical and does not take into account the context in which
+a word appears. Consequently, many words that have very specific senses in
+the context of the IETF get miscounted as occurrences of LIWC categories.
+This could be addressed in future by a more advanced method of mapping to
+LIWC categories that would account for context.
+
+%..................................................................................................
+% The following is from our ACL 2023 Tracing Linguistic Markers of
+% Influence paper RQ1: How do linguistic traits differ between more
+% and less influential participants?
+
+\pb{Language and Influence:} How do linguistic traits differ between more
+and less influential participants? That is, are there differences in the
+way influential participants, based on centrality in the email social
+graph, make use of language compared to those who are less influential?
+
+We study mailing list data from the period 2015-2019, comprising 300,806
+emails from 5,363 unique participants. We used a 5-year subset of the data
+due to the computation cost, while still giving a reasonable period to
+observe the participation consistency in the IETF community, and chose the
+last period prior to the distortion of the pandemic.  For each participant,
+we calculate the centrality-based influence score and LIWC representation,
+We fit a linear regression model using LIWC representations to predict
+influence percentile and observe the magnitude and directions of
+significant coefficients.
+
+The following LIWC categories are highly correlated ($p < 0.05$) with
+centrality-based influence: \liwc{we, informal, risk, adjective, anger,
+they}, and \liwc{bio}.  Categories such as \liwc{netspeak, sexual, health,
+death, body} are correlated with lower influence. This suggests that
+influential people tend to indicate a collaborative and community-oriented
+approach with first-person plural (\liwc{we}) and third-person plural
+category (\liwc{they}) usage. This is consistent with
+\cite{kacewicz2014pronoun} and \cite{guinote2017power}, who show that
+influential people use more first-person plural. They also use more
+organisational language, which is shown by the negative correlation of
+informal slang language categories (\liwc{netspeak, sexual, body}). We see
+some unexpected hidden trends due to word ambiguity (e.g., words like
+`\textit{trust}' and `\textit{live}'), which are discussed below.
+
+%..................................................................................................
+% The following is from our ACL 2023 Tracing Linguistic Markers of
+% Influence paper RQ2: How do linguistic traits vary for participants
+% at different levels of the organisation hierarchy?
+
+\pb{Language and Organisational Role:}
+How do linguistic traits vary for participants at different levels of the
+organisation hierarchy?
+
+Centrality-based influence considers participants based on their email
+network, but role-based influence, based on the position of a person in the
+organisational hierarchy (regular participant, working group chair, area
+director) is equally crucial as they are involved in organisational
+decision making.\footnote{In the top 10\% \textit{mail-based} influential
+participants, less than 30\% are WG chairs with significant
+\textit{role-based} influence.} 
+For the same time period, 2025-2019, we split the data into two categories:
+(a) emails sent by working group chairs, and (b) emails sent by regular
+participants who have never been a working group chair (the number of area
+directors is small, so we exclude them from this analysis).  We calculate
+the LIWC representations for each person, train a logistic regression model
+to predict category, and observe the LIWC category coefficients.
+
+From \ref{table:all_results}, we see that working group chairs are more
+social and collaborative, as is shown by \liwc{we} and \liwc{social}
+categories. This is in line with our findings for influence-based
+centrality above and resuslts on leadership engagement from the literature
+\cite{strzalkowski2012modeling, liu2022pronoun, kacewicz2014pronoun,
+guinote2017power}.
+Chairs also use more tentative statements (\liwc{tentat}) in discussions,
+primarily focused on technical feedback and revisions, or suggesting
+alternatives. For example:
+\textit{``With the risk of disturbing with statements, but avoiding too
+many questions:This \underline{seems} against the goal of reducing
+headers.''} 
+and
+\textit{``Question is do we need to carry around an outer IP-in-IP header
+for that \underline{or} not?''}.
+
+\begin{table*}[]
+  \small
+  \begin{center}
+  \begin{tabular}{l|l|l}
+  \toprule
+    \multirow{2}{*}{Language and influence} 
+      & High influence
+      & \liwc{bio, we, informal, they, negemo, anger, risk, adjective}  \\ 
+    \cmidrule{2-3}
+      & Low influence
+      & \liwc{sexual, death, ingest, netspeak, health, female, body, affiliation, conj} \\ 
+  \hline
+    \multirow{2}{*}{Language and role} 
+      & WG Chair influence
+      & \liwc{tentat, ipron, social, see, feel, we} \\
+    \cmidrule{2-3}
+      & non-WG Chair
+      & \liwc{cogproc, relativ, affiliation, i, reward} \\
+  \hline
+    \multirow{2}{*}{Changes in Language}
+      & Top 10 percentile
+      & \liwc{adverb, prep, anger, auxverb, male, cogproc, achiev, risk, focuspresent} \\
+    \cmidrule{2-3}
+      & Below 50$^{th}$ percentile
+      & \liwc{function, ppron, shehe, ipron, number, certain, sexual, informal} \\
+  \hline
+  \end{tabular}
+  \end{center}
+  \caption{LIWC categories where $p<0.05$.}
+  \label{table:all_results}
+\end{table*}
+
+%..................................................................................................
+% The following is from our ACL 2023 Tracing Linguistic Markers of
+% Influence paper RQ3: How does linguistic behaviour of participants
+% change as they gain influence?
+
+\pb{Changes in Use of Language with Changes in Role:}
+Finally, we look at how linguistic behaviour of participants changes as
+they gain influence in the IETF.
+We consider at participants who went from low to high influence over time:
+individuals who had a \emph{centrality-based} influence below the 50th
+percentile when they joined the IETF, and reached the top 10th percentile
+at some point.
+
+For each participant, we generate two different representations based on
+two periods---the year of joining, and year of reaching the top 10th
+percentile for the first time---and assign these to two different classes.
+We then train a logistic regression model to predict these classes, and
+examine the coefficients of the LIWC categories
+
+From Table \ref{table:all_results}, we observe that when participants
+become influential, according to the centrality-based metric, they are
+likely to be more descriptive and engaged in immediate state of issues
+and situations as seen from the correlation of auxiliary verbs
+(\liwc{auxverb}), adverb, risk, and present focus (\liwc{focuspresent}).
+They are also more involved in cognitive processes (\liwc{cogproc}) as
+compared to their previous self when they were new to IETF and had little
+influence.
+
+%..................................................................................................
+\pb{Discussion:}
+To better understand these LIWC categories and what kind of words play a
+role in the behaviour of individual categories, we calculate the frequency
+of words in each LIWC category as they appear in the emails. 
+Next, we consider the top 30 most frequent words in each LIWC category and
+perform regression analysis on centrality-based influence for participants,
+but using only these 30 words as features to generate the participant
+representation. We conducted this experiment separately for each LIWC
+category that was significant in the first experiment.
+
+From the word based analysis we make multiple observations. Firstly, words
+like \word{we} imply a collective approach and is strongly correlated with
+the higher influence. 
+Similarly, the use of word \word{well} is standard, such as politely
+resuming the conversation (e.g., `\textit{well, I agree}') or providing an
+approval over something (e.g., `\textit{this works as well}'). These words
+are well associated with the influential participants.  Otherwise,
+influential participants are generally not observed to be informal and
+other frequent words (other than \word{well}) within \liwc{informal}
+category do not demonstrate a strong correlation with the growing
+influence. Also, \word{well} is the most frequent word in the
+\liwc{informal} category.
+
+More influential people, according to both centrality- and role-based
+metrics, are also observed to engage more with the community.
+Conversations can often reflect situations where, as a part of review and
+feedback process, more influential people highlight limitations in protocol
+standards, stress on specifics, and compare with existing protocols or
+previous versions.
+Several words across different LIWC categories (\liwc{risk}, \liwc{negemo},
+and \liwc{adj}) highlight such behaviour, e.g., \word{problems}, \word{before},
+\word{particular}, \word{specific}, \word{different}, \word{most}, and \word{than}.
+
+However, there are many words with dual sense, like \word{trust} which has
+a very technology specific usage related to network security instead of
+conversations involving trust issues between individuals or trust in any
+given situation. Similarly, the word \word{live} is related with an
+application or network being live, instead of its conventional meaning. We
+also observed that some of the LIWC categories, such as \liwc{bio}, did not
+have specific terms that could clearly establish its significance in favour
+of influential participants (e.g., word \word{problems} and \word{trust}
+reflecting the significance for the category \liwc{risk}), instead such
+categories had several words with quite weak correlation with influential
+participants. Such words collectively drifted the weight of the category
+towards influential participants.
+
+%..................................................................................................
+\pb{Summary:}
+Using two aspects of influence, based on centrality in the email social
+graph and based on organisation role, we were able to unfold several traits
+that differentiate influential participants from others. Many of our
+findings seem corroborated by studies in organisational theory. 
+We observed that influential people exhibit more collaborative and
+community-oriented traits, and also stronger signs of engagement in
+discussions. We also observed that as people go on to become influential
+participants, they evolve in their communication and are seen to be more
+engaging and descriptive in their linguistic style.
+
+% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 
@@
 
+%--------------------------------------------------------------------------------------------------
 \subsection{What Makes a Successful Author?}
@@
 
+%--------------------------------------------------------------------------------------------------
 \subsection{What Makes a Successful Document?}

### 2024-10-03 paper.tex
Briefly summarise some results from our Frontiers paper

@@
 
-\pb{Changes in Use of Language with Changes in Role:}
+\pb{Changes in Language with Changes in Role:}
 Finally, we look at how linguistic behaviour of participants changes as
@@
 %..................................................................................................
-\pb{Discussion:}
+% The following is from our ACL 2023 Tracing Linguistic Markers of
+% Influence paper, Section 4.2.
+\pb{What Kinds of Language are Used?}
 To better understand these LIWC categories and what kind of words play a
@@
 %..................................................................................................
+% The following is from our Frontiers in Psychology paper:
+
+\pb{Use of Sensitive Language:}
+We briefly consider the use of sensitive language, relating to politics,
+race, and religion, in IETF email communications.
+
+Using email data for 2019, the last year prior to the pandemic, we
+categorise the emails according to the role of their sender, based
+on the relative social power of the sender within a working group:
+\emph{working group chair} for emails sent by the current chair of the
+working group to which the email was sent;
+\emph{other chair} for emails sent by the chair of a different working
+group; and
+\emph{regular participant} for emails sent by people who were not, and had
+not been at the time of sampling, a workgroup chair.
+Due to their small numbers, we exclude area directors.
+
+We use three LIWC categories to infer the presence of potentially sensitive
+language in the emails: 
+\liwc{Politics}---words commonly used in political discussions;
+\liwc{Ethnicity}---words that identify national, regional, linguistic,
+  ethnic, or racial identities; and
+\liwc{Religion}--use of religious words.
+Each email is coded according to the working group and role of the sender.
+The emails are pre-processed to remove subject lines, headers, and any
+embedded quotes from previous emails, yielding 101,857 texts by 2,300
+individuals across 176 group mailing lists.  Each text is assigned a score
+on each of the three language categories using the LIWC software and the
+resulting scores are averaged for each individual sender for statistical
+analysis.
+
+Results are shown in Figure \ref{fig:sensitive}. The consistent pattern
+is that people with organizational responsibilities (working group chair,
+other chair) are less likely to use words potentially connected with
+sensitive topics than regular participants without organizational
+responsibilities, even though they are all part of the same discussions
+on the same email lists.
+Averaging across the two working group chair categories, the means indicate
+that people in positions of social power are 3.2 times less likely to use
+sensitive language overall than those who are not (by category:
+\liwc{Politics} 2.0 times; \liwc{Religion} 2.3 times; \liwc{Ethnicity} 5.5
+times).
+It is worth noting, however, that in absolute terms potentially sensitive
+language use is rare in this dataset, as might be expected for primarily
+work-related communication. The modal LIWC score in both raw data and the
+averages scores is zero
+
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=\figureWidthTwoColumn]{figures-prev/frontiers/sensitive.pdf}
+  \caption{
+    Patterns of language use by organizational role. Estimated marginal
+    means for the LIWC categories: (A) Politics, (B) Ethnicity, and (C)
+    Religion. Error bars show 95% confidence intervals.
+  }
+  \label{fig:sensitive}
+\end{figure*}
+
+%..................................................................................................
+% The following is from our ACL 2023 Tracing Linguistic Markers of
+% Influence paper
 \pb{Summary:}
 Using two aspects of influence, based on centrality in the email social
-graph and based on organisation role, we were able to unfold several traits
-that differentiate influential participants from others. Many of our
-findings seem corroborated by studies in organisational theory. 
+graph and based on organisation role, we were able to unfold several
+traits that differentiate influential participants from others. Many
+of our findings seem corroborated by studies in organisational theory.
 We observed that influential people exhibit more collaborative and
-community-oriented traits, and also stronger signs of engagement in
-discussions. We also observed that as people go on to become influential
-participants, they evolve in their communication and are seen to be more
-engaging and descriptive in their linguistic style.
-
-% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-
-% Frontiers in Psychology paper
-% - Use of Language, discussing how it shifts as people become more senior
-
-% LEDA
-% - a taxonomy of dialogue acts (DAs) and a labeled dataset of emails
+community-oriented traits, show stronger signs of engagement in
+discussions, and are systematically less likely to use potentially
+sensitive language.
+We also observed that as people go on to become influential participants,
+they evolve in their communication and are seen to be more engaging and
+descriptive in their linguistic style.
 

### 2024-10-03 paper.tex
Complete §7.1 (successful author)

@@
 
+\todo{introductory remarks}
+
 %--------------------------------------------------------------------------------------------------
@@
 
+We have identified that influential authors write more drafts
+(\S\ref{subsec:impact_influence}), but are these drafts more likely to
+see success?  That is, does the presence of a well-connected, influential,
+author make it more likely that an Internet-draft is adopted by a working
+group for review and development\footnote{There are occasional document,
+known as Area Director sponsored drafts, that are published as standards
+track RFCs without going through the working group process, but these too
+rare ($\sim$3\% of RFCs) to significantly affect our analysis.} and
+eventually published as a standards track RFC.  Using information from the
+IETF Datatracker we compile a dataset for the \emph{adoption} of a draft by
+a working group, comprising 11632 drafts, and employ logistic regression to
+analyse factors that affect working group adoption of draft.
+
+%..................................................................................................
+\pb{Methodology:}
+We compute a set of features that may impact adoption or publication for
+each draft, most of them motivated by the characteristics of influential
+participants, including the following feature groups: 
+ (1) \emph{text}, we derive a term frequency--inverse document
+     frequency (TF-IDF) \cite{salton1983introduction} weighted bag of words vector
+     representation of all email text of messages that either explicitly
+     mention a draft or are part of a thread where the subject line
+     explicitly mentions a draft; 
+ (2) \emph{communication patterns}, i.e., volumes of incoming/outgoing communication
+     between draft authors and IETF participants of varying experience levels;
+ (3) Betweenness \emph{centrality} for authors;
+ (4) \emph{number of emails} sent by authors;
+ (5) \emph{number of drafts} submitted by authors;
+ (6) \emph{length of participation} duration of authors;
+ (7) \emph{number of authors};
+ (8) \emph{number of IETF areas} in which the author is active;
+ (9) \emph{proportion of top authors} in the top 10 and top 20 percent of
+     influential participants;
+(10) \emph{topic entropy ($\eta$)} scores for authors;
+(11) \emph{author affiliations} for each of the 20 most influential
+     organizations (20 boolean features);
+(12) \emph{number of mailing lists} that each author participates in,
+(13) \emph{number of years active}.
+
+For numerical features that relate to each author individually instead of
+the draft as a whole (e.g., \emph{number of emails sent}, \emph{centrality},
+or \emph{number of years active}, the feature defined for each author
+separately) in three variants: (i) for the least influential author, (ii)
+for the most influential author, and (iii) as the average value across all
+authors of the draft. We compute individual author features using
+information from the 5-year period prior to draft submission.
+
+Regarding the text based features, we found that the models tend to assign
+high weights to words such as surnames of active contributors, names of
+prolific working groups, and common technical terms. While this does make
+the models perform slightly better, such terms should not be relevant for
+solving the task as they do not model the relevant part of the conversation.
+To explore how models would behave in a scenario wherein such words were
+not available, we construct a domain-specific stop word list, consisting
+of: (i) one all last names from the Datatracker, (ii) all working group
+names from the Datatracker, and (iii) technological jargon terms obtained
+from the web\footnote{\url{https://www.computerhope.com/jargon.htm}, we
+used a union of the \emph{Internet terms}, \emph{network terms}, and
+\emph{security terms} categories.}. We remove from this list any terms
+appearing in top 5,000 English terms (e.g., sometimes people's names can have
+identical surface forms as some of the terms relevant to organisational
+activities). We will refer to this variant as \emph{text (S)}.
+
+To gain empirical insights into which features provide better prediction
+results, models are run using a feature group alone variant and a variant
+combining each feature group with the \emph{text} features. This approach
+was motivated by the finding that \emph{text}, even though conceptually
+quite different from the rest of the feature set and not our main focus,
+does provide surprisingly strong results on its own. Thus, we wanted to
+empirically investigate in more detail how well it complements the rest of
+the graph-based features. Moreover, we compare all models to a baseline
+which simply assigns all examples to the majority class.
+
+We split the data into \emph{training} (70\%), \emph{development} (10\%),
+and \emph{test} (20\%) subsets. As a scoring function, we use the F1 score
+macro averaged across the positive and negative classes (but we also report
+area under the curve (AUC), precision, and recall). We train the models on
+the training set, optimise hyper-parameters on the development set, and
+report final scores on the test set. The final scores are those obtained by
+the model variant that fared best on the development set. We use logistic
+regression implemented in \texttt{scikit-learn} \cite{scikit-learn}, given
+that it is widely used and well interpretable. While we do not perform
+explicit feature selection, we do have implicit feature selection through
+L1 regularisation. We consider the regularisation strength a hyper-parameter
+and consider values from $[2^{-7}, 2^{-6}, ..., 2^6]$ on the development
+set. As our data set is quite imbalanced (in a roughly 4:1 ratio, 17\% is
+\emph{adopted}), we use different class weights during optimization to
+counteract this. We employ a non-parametric random shuffling test
+\cite{yeh-2000-accurate} to check statistical significance of score
+differences against the baseline. Table~\ref{tbl:res} summarises prediction
+results, which will be discussed in the next section.
+
+\renewcommand{\tabcolsep}{0.5cm}
+\begin{table*}
+\centering
+\begin{tabular}{lcccccccc}
+\toprule
+                        & \multicolumn{4}{c}{$\neg \mathit{TXT}$}
+                        & \multicolumn{4}{c} {$\mathit{TXT}$} \\
+\midrule
+                       & AUC & F1 & P & R & AUC & F1 & P & R \\
+\midrule
+Baseline               &  .500 & .445 & .401 & .500  &  .500 & .445 & .401 & .500 \\
+\midrule
+Comm. patterns         &  .644 & .583 & .566 & .601 &  .723 & .626 & .609 & .644 \\
+Centrality             &  .655 & .577 & .573 & .581 &  .717 & .639 & .625 & .653 \\
+Email count            &  .632 & .566 & .555 & .577 &  .718 & .650 & .639 & .662 \\
+N. years active        &  .608 & .569 & .555 & .584 &  .702 & .610 & .594 & .626 \\
+Number of authors      &  .578 & .536 & .529 & .544 &  .697 & .631 & .621 & .641 \\
+Proportion in top      &  .663 & .594 & .580 & .609 &  .728 & .634 & .614 & .655 \\
+Draft count            &  .500 & .445 & .401 & .500 &  .696 & .630 & .624 & .636 \\
+N. Areas               &  .634 & .572 & .557 & .588 &  .721 & .639 & .622 & .657 \\
+N. Mailing lists       &  .634 & .570 & .557 & .584 &  .718 & .641 & .630 & .653 \\
+Affiliations           &  .605 & .584 & .576 & .592 &  .708 & .626 & .609 & .643 \\
+Topic entropy ($\eta$) &  .622 & .563 & .550 & .577 &  .715 & .642 & .628 & .657 \\
+Text                   & - & - & -  & -  &  .681 & .624 & .622 & .627 \\
+Text (S)               & - & - & -  & -  &  .667 & .598 & .593 & .614 \\
+\midrule
+All feats              &  .692 & .624 & .600 & .651 &  .744 & .645 & .625 & .666 \\
+All feats (S)          &  .692 & .624 & .600 & .651 &  .726 &  .636 &  .627 &  .664 \\
+\bottomrule
+\end{tabular}
+\caption{
+  Results for predicting adoption. Each row presents the scores of a model
+  using the corresponding feature group either alone ($\neg \mathit{TXT}$)
+  or combined with the Text features ($\mathit{TXT}$). \emph{All feats}
+  denotes all features except the text. Rows labeled with (S)  use the
+  \emph{Text (S)} variant of the text features.
+}
+\label{tbl:res}
+\end{table*}
+
+Next, we wanted to confirm these results at the level of individual
+features. To this end, we inspect the statistically significant
+coefficients learned by the model corresponding to each individual feature.
+For this experiment, we use a slightly different setup. We do not consider
+the \emph{text} features\footnote{There are thousands of text features
+(terms), making it prohibitively complex to include them. Moreover, they
+are not central to our research questions about the IETF community.}. We
+begin by first applying the Variance Inflation Factor (VIF) to exclude all
+features with $\mathit{VIF} > 5$. This, to an extent, mitigates the
+collinearity that we know exists as some features are by construction
+highly correlated. We then standardise each remaining feature and then fit
+a Logistic Regression model from the \texttt{statsmodels} package
+\cite{seabold2010statsmodels} on the entire data set.
+Table~\ref{tbl:resstat} presents the results of this experiment.
+
+\renewcommand{\tabcolsep}{0.3cm}
+\begin{table}
+  \centering
+  \begin{tabular}{lccc}
+    \toprule
+      Feature       & Weight  & P-value \\
+    \midrule
+      Centrality of most influential  & 0.1691 &   0.000 \\
+      Centrality of least influential & -0.0698 &  0.048 \\
+      Proportion of authors in top 10  & 0.1520 &  0.000 \\
+    \midrule
+      Has a Cisco author  & 0.1301 &  0.000 \\
+      Has an AT\&T author  & 0.1003 &  0.000 \\
+      Has a Juniper author  & 0.0738 &  0.000 \\
+      Has a China Mobile author & -0.0697  & 0.000 \\
+      Has an Alcatel-Lucent author  & 0.0474 &  0.022 \\
+    \midrule
+      Topic entropy max & 0.1728 & 0.000 \\
+      Topic entropy min & -0.1135 &  0.002 \\
+    \bottomrule
+  \end{tabular}
+  \caption{
+    Weights of the model that are statistically significant at $p \leq 0.05$
+    grouped by similarity of features -- author influence measures vs.\
+    affiliation features vs.\ topic entropy ($\eta$).
+  }
+  \label{tbl:resstat}
+\end{table}
+
+%..................................................................................................
+\pb{Predicting Adoption:}
+To confirm empirical implications from Section \ref{subsec:impact_influence},
+and provide a more in-depth look at the features, we perform statistical
+analysis summarised in Table~\ref{tbl:resstat}. The most interesting
+insight is that the coefficients corresponding to the \emph{Centrality of
+the most influential author} and \emph{Proportion of authors in the top 10
+percentile in influence}, are, indeed, positive and among the largest in
+absolute value. This highlights that among the people who are influential
+in the email networks, there is a substantial number of individuals who are
+proficient in contributing to successful drafts.
+
+Another interesting observation is that author affiliations (which are
+also indirectly connected to influential participants, as demonstrated in 
+\S\ref{subsec:impact_influence}, Figure \ref{fig:coauthor_affiliation_network2019_otheryears})
+are an important feature group, particularly affiliations such as \textit{Cisco}
+and \textit{AT\&T}. Moreover, topic entropy is also an important feature
+group, which was also shown to be related to influential participants in
+our analysis from \S\ref{subsec:impact_influence}, Figure \ref{fig:topic_diversity}.
+
+%..................................................................................................
 % ICWSM 2022 paper section 4.3
+\pb{Summary:}
 The statistical analysis shows that influence of draft authors in the email
-networks does impact the possibility of a draft getting adopted
-(\S\ref{subsec:results}, Table \ref{tbl:res} \& \ref{tbl:resstat}). This
-might hint at the ability of participants who hold a domain expertise to be
-able to engage better with the community. Several WG chairs are already in
-the top percentile influential category in both the email and co-authorship
-networks before taking up these leadership roles, which further elevates
-after taking up such leadership roles (Figure
-\ref{fig:wg_chair_before_after}).
-
-Most importantly the statistical analysis shows that being affiliated with
-a prominent organisation positively impacts the chances of a draft in
-getting adopted by a WG, thereby directly driving the innovation process
-(Table \ref{tbl:res} \& \ref{tbl:resstat}).
+networks does impact the possibility of a draft getting adopted (Tables
+\ref{tbl:res} \& \ref{tbl:resstat}). This might hint at the ability of
+participants who hold a domain expertise to be able to engage better with
+the community. Several WG chairs are already in the top percentile
+influential category in both the email and co-authorship networks before
+taking up these leadership roles, which further elevates after taking up
+such leadership roles (Figure \ref{fig:wg_chair_before_after}).  Our
+analysis also shows that being affiliated with a prominent organisation
+positively impacts the chances of a draft in getting adopted by a working
+group, thereby directly driving the innovation process (Table \ref{tbl:res}
+\& \ref{tbl:resstat}).
 

### 2024-10-03 paper.tex
Complete §7.2 (successful document)

@@
 
+The trends that we observed in Section~\ref{sec:trends-documents}
+highlight that, over time, the complexity of the standardisation process is
+increasing: RFCs take longer to produce, and relate to a larger number of
+other documents; authors come from a changing and diversifying set of
+affiliations and countries; and the volume of email interaction for each
+RFC is increasing, with differences in how senior and junior authors and
+contributors interact.  We next seek to understand how these factors might
+impact the successful deployment of an RFC in-the-wild, and to test if such
+success can be predicted.
+
+% IMC 2021 paper Section 4.1
+
+We build on the work of Nikkhah et al.~\cite{nikkhah2017statistical}.
+However, while the focus of that paper was on a set of features that were
+derived from the standards documents themselves, we augment this with a set
+of features derived from the author and email interaction datasets
+described in the previous section.
+
+To explore the factors that influence protocol deployment, we follow a
+multi-step process:
+\begin{enumerate}
+  \item We reproduce the logistic regression model in~\cite{nikkhah2017statistical},
+    to obtain similar results (as shown in Table~\ref{tbl:mlres}).
+    That paper published an expert annotated dataset, labelling RFCs as
+    ``successfully deployed'' or not. This pertains to whether the RFC was
+    implemented in-the-wild and saw widespread (largely commercial) uptake.
+    This dataset covers 251 RFCs published between 1983 and 2011, and
+    includes 20 features. We create a baseline logistic regression model
+    with $F_1 = 0.762$, $\mathit{AUC} = 0.650$. The results are similar to
+    those reported in~\cite{nikkhah2017statistical} ($\mathit{AUC} = 0.670$).
+
+  \item We curate a set of additional \emph{classification features} driven
+    by our findings in Section~\ref{sec:trends-documents}. These are derived
+    from the RFCs, their authors, and the mailing list interactions. Given
+    that these features are only available for RFCs where Datatracker
+    metadata is available, we model these features across a subset of the
+    original labelled RFCs from~\cite{nikkhah2017statistical}, looking only
+    at the 155 RFCs where all features can be calculated. We will show that
+    modelling these RFCs with the expanded feature set substantially
+    expands the predictive power of the (Step 1) baseline logistic
+    regression model, with $F_1 = 0.820$, $\mathit{AUC} = 0.822$.
+
+  \item We then train classification models to predict if an RFC will be
+    deployed, using our expanded feature set.  We test the model with the
+    manually labelled RFCs and show that we can successfully classify
+    RFCs as deployed or not, with a decision tree-based model having
+    $F_1 = 0.822$, $\mathit{AUC} = 0.838$. 
+\end{enumerate}
+
+\begin{table}
+  \centering
+  \begin{tabular}{lrrr}
+    \toprule
+      Model                              & $F_1$ & $\mathit{AUC}$ & $F1_{\mathrm{macro}}$ \\
+    \midrule
+      Most frequent class                & .757  &      .500      &     .379     \\
+      Baseline                           & .758  &      .616      &     .597     \\
+      Baseline + FS                      & .762  &      .650      &     .610     \\
+    \midrule
+      Most frequent class                & .724  &      .500      &     .379     \\
+      Baseline                           & .670  &      .559      &     .547     \\
+      Baseline + FS                      & .690  &      .620      &     .563     \\
+      Logistic regression all feats      & .728  &      .724      &     .666     \\
+      Logistic reg. all feats + FS & .820  &      .822      &     .789     \\
+      Decision tree all feats + FS       & .822  &      .838      &     .788     \\
+    \bottomrule
+  \end{tabular}
+\caption{
+  Classifier scores on the entire dataset (251 RFCs, above) and those with
+  our features available (155 RFCs), with or without feature selection (FS).
+}
+\label{tbl:mlres}
+\end{table}
+
+%..................................................................................................
+% IMC 2021 paper Section 4.2
+\pb{Classification Features:}
+We begin with the set of features that were originally derived in
+\cite{nikkhah2017statistical}:
+IETF Area; Scope (Local, End-to-End (E2E), Bounded (BN), or Unbounded
+(UB)); Type (New (N), New with Incumbent (NI), Backward compatible
+extension (EB), or Extension (E)); Change to others (CO); Scalability
+(SCAL); Security (SCRT); Performance (PERF); Adds value (AV); and Network
+effect (NE). We refer the reader to~\cite{nikkhah2017statistical} for a
+full explanation of each of these feature.
+
+We further define an additional set of document-based features, derived
+from our characterisations in \S\ref{sec:trends-documents}: days from first
+draft to RFC publication; number of drafts before RFC publication; number
+of outbound citations to RFCs/Internet-Drafts; page count; number of
+inbound citations from articles listed in Microsoft Academic, one and two
+years after publication; number of inbound citations from other RFCs
+published within one and two years after publication; RFC updates or
+obsoletes a previous RFC; keywords per page; and topics (where we use
+Latent Dirichlet Allocation (LDA) \cite{blei2003latent} to induce 50 topics
+on the texts of all existing RFCs, and use the 50-dimensional probability
+distribution over topics for a given RFC as the feature vector).
+
+Next, we derive a set of author-based features, from the work described in
+\S\ref{sec:trends-demographics}: number of authors on the RFC; if at least
+one author of previously published  RFC; if at least one author in North
+America, Europe, or Asia; if at least one author from Cisco, Huawei, or
+Ericsson; authors with diverse affiliations; authors in more than one
+continent;  at least one academic author; and at least one consultant
+author.
+
+Finally, we derive a set of features based on email interactions, from
+\S\ref{sec:org-dyn:participants}: number of emails mentioning the
+Internet-Drafts that precede publication of the RFC; mean number of emails
+sent to all RFC authors, for each sender contribution-duration category
+(young, mid-age,  senior); mean number of contributors sending emails to
+any RFC author, for each contribution-duration category; number of emails
+sent to most junior and senior RFC authors, for each sender
+contribution-duration category; and number of contributors sending emails
+to most junior and senior RFC authors, for each contribution-duration
+category.
+
+%..................................................................................................
+% IMC 2021 paper Section 4.3
+\pb{Modelling Methodology:}
+With our expanded feature set, we next build a classifier to predict the
+success (i.e., deployment) of RFCs.
+
+Our expanded feature set, including all variants of each feature, is very
+large, totalling 177 features. Given that our additional features can only
+be calculated for a limited number of data points (155 RFCs),  it would be
+infeasible to conduct statistical analysis, and to build a predictive
+model, as there are too few data points. To address this, we take a number
+of steps to reduce the feature space, while maintaining interpretability:
+
+\begin{enumerate}
+  \item Since the largest feature groups are the topics (50) and
+    interaction features (54) we reduce both by applying the $\chi^2$ test
+    to leave only the top 5 features in each group.
+
+  \item We remove collinearity by using the Variance Inflation Criterion
+    (VIF), removing all features with a VIF value above 5.
+
+  \item We apply forward Feature Selection (FS) to identify features of
+    high predictive value, following~\cite{nikkhah2017statistical}.
+    Starting from an empty feature set, in each iteration of the forward
+    procedure, we expand the feature set with the feature that provides the
+    largest increase in the $AUC$ score. The procedure ends when there are
+    no more unused features that yield score improvements over the current
+    feature set. 
+\end{enumerate}
+The final set of features with and without FS is given in Tables
+\ref{tbl:statsnofs} and \ref{tbl:statsfs} respectively.
+
+Using the above feature set, we then train two classification models relying
+on logistic regression and a decision tree.  For assessing predictive
+performance of the models we use leave-one-out cross-validation, while for
+the final statistical analysis we fit a logistic regression model on the
+entire dataset and report the statistically significant coefficients (at
+significance level $p \leq 0.1$).
+
+\input{tables-prev/imc-2021/lr_fsnone-trim.tex}
+\input{tables-prev/imc-2021/lr_fsauc-trim.tex}
+
+\begin{table}
+  \centering
+  \begin{tabular}{lrrr}
+    \toprule
+      Model & $F_1$ & $\mathit{AUC}$ & $F1_{macro}$ \\
+    \midrule
+      Most frequent class                & .757   & .500   & .379 \\
+      Baseline                           & .758   & .616   & .597 \\
+      Baseline + FS                      & .762   & .650   & .610 \\
+    \midrule
+      Most frequent class                & .724   & .500   & .379 \\
+      Baseline                           & .670   & .559   & .547 \\
+      Baseline + FS                      & .690   & .620   & .563 \\
+      Logistic regression all feats      & .728   & .724   & .666 \\
+      Logistic regression all feats + FS & .820   & .822   & .789 \\
+      Decision tree all feats + FS       & .822   & .838   & .788 \\
+    \bottomrule
+  \end{tabular}
+  \caption{
+    Classifier scores on the entire dataset (251 RFCs, above) and those
+    with our features available (155 RFCs), with or without feature
+    selection (FS).
+  }
+  \label{tbl:mlres}
+\end{table}
+
+%..................................................................................................
+% IMC 2021 paper Section 4.4
+\pb{Evaluative Results:}
+Using the trained models, we next evaluate their accuracy and explore which
+features are most predictive of RFC deployment.
+
+As evaluative metrics, we use the $F_1$ score and the area under the ROC
+curve ($AUC$), as in \cite{nikkhah2017statistical}. We find the standard
+$F_1$ score gives overly optimistic performance estimates due the data
+being skewed towards the positive class. Consequently, we also report an
+$F1_{macro}$ score that takes into account both the positive and negative
+class, reflecting performance more realistically. As a baseline comparison,
+we present results of our re-implementation of \cite{nikkhah2017statistical}
+using their feature set.
+
+The prediction results of the models and feature sets are summarized in
+Table \ref{tbl:mlres}. Our baseline model performs at a similar level to
+that from \cite{nikkhah2017statistical}.  We further obtain considerable
+performance improvements with our additional features. Furthermore, we find
+that additional performance can be obtained by using feature selection.
+The best performing model, with an F1 score of 0.822, is the decision tree
+trained on the entire feature set.
+Note, we also tested several non-linear models (neural networks, support
+vector machines with non-linear kernels). These attained similar or worse
+results as our decision tree model, and therefore we omit them due to space
+constraints.
+
+We next investigate the most important features in predicting success.  We
+posit that these can offer useful insight for working group chairs and RFC
+editors.  The feature importance results are presented in
+Tables~\ref{tbl:statsnofs} and \ref{tbl:statsfs}.  Some meaningful patterns
+emerge here. We observe features such as adding value, being scalable,
+obsoleting other RFCs, page count, increased keyword usage, no incumbent
+RFC to compete against and limited scope all being positively correlated
+with the likelihood of an RFC becoming deployed. In contrast, we find that
+having a broad (unbounded) scope negatively impacts deployment.
+
+We also observe some curious trends that speak to the limitations of our
+dataset. The model finds that having an author in Asia is negatively
+correlated with deployment, though with only borderline statistical
+significance. Further analysis shows that only 10\% of labelled RFCs have
+an author in Asia.  As shown in \S\ref{sec:trends-demographics}, we see
+that demographics of the IETF are changing, with a recent notable increase
+in author representation from Asia. This might suggest that new authors
+need time to learn what makes a deployable protocol; equally our deployment
+data may be biased towards the Internet in North America and Europe. This
+finding requires much more exploration.
+
+We further see that some of the topics extracted by LDA are also useful.
+For example, Topic 13 is characterised by a cluster of terms associated
+with MPLS, a widely deployed routing protocol. We find a negative
+correlation with this topic, but this is likely because there is a
+significant number of RFCs that propose modifications or additional
+features for the protocol, some of which do not see deployment. Overall,
+most of the results are in line with expectations, but more annotated data
+is required for further insights.
+
+%..................................................................................................
+% IMC 2021 paper Section 4.5
+\pb{Discussion:}
+% Building on existing work.
+Features associated with building on existing RFCs positively correlate
+with deployment. RFCs that obsolete earlier versions of the same protocol
+are likely to be deployed, indicating that the IETF community is good at
+identifying and maintaining protocols that are seeing use. This is also
+highlighted by the significance of the inbound RFC citation and ``adds
+value'' features. These show that RFCs that are later cited by other
+documents, or that add value to other protocols in the stack, are more
+likely to be deployed. Protocols that are extensible, and can be extended
+and adapted for new uses, are more likely to see deployment. The discussion
+in \cite{clark:2002:tussle} on careful choice of extension points
+resonates; as do the implications for starting new work.
+
+% Limited scope.
+We find that having end-to-end scope (i.e., where only the endpoints of a
+connection needs to implement the RFC) is positively correlated with
+deployment, whereas having an unbounded scope (i.e., where the entire
+Internet may need to be updated) is negatively correlated. This indicates
+that well-scoped RFCs, that are cognisant of their deployment challenges,
+are more likely to see deployment. A recent IAB workshop \cite{rfc8980}
+noted that deployment often occurs in unforeseen ways: limiting the changes
+needed to deploy a protocol facilities this.
+
+% Normative requirements.
+The number of keywords (e.g., SHOULD, MUST) used per page, which shows the
+number of normative requirements an RFC imposes on implementations, also
+correlates with deployment. This suggests that well-specified requirements
+lead to robust, interoperable, implementations that see wide use.
+Unsurprisingly, features such as citation rates are also predictive of
+deployment.
+
+% Diversity.
+Notwithstanding the limitations of the dataset, we find that the majority
+of author demographic features are not significant. While \S\ref{sec:trends-demographics}
+identified recent shifts in the demographics of authors, these do not
+appear to have a major influence in the deployment of the RFCs that they
+write.
+
+% Meeting a need.
+Our results broadly support the conclusions of \cite{RFC5218}, where the
+IAB noted that, in addition to being technically sound, a successful
+protocol must meet a need, be incrementally deployable, and be open to
+extension and maintenance.
+
+%..................................................................................................
+\pb{Summary:}
+Our results show that RFCs that build on existing work, have well-defined
+requirements, limited scope, and meet a need have improved chances of being
+widely deployed. These document-based features have implications for how
+the IETF should design and specify protocols.
+
+We found that the majority of our author-based features, including
+geographic and affiliation diversity, did not significantly impact the
+deployment prospects of RFCs. Further work is needed to expand the dataset
+to ensure that this result holds across a larger set of RFCs.
+
+Finally, the focus of our modelling in this paper has been on the impact of
+document, author, and interaction-based features on protocol deployment.
+However, this is only one part of the lifecycle of an RFC. It remains to
+consider the impact of these, and other, features on the key stages of an
+Internet-Draft's development towards becoming an RFC, such as working group
+adoption.
+
 %==================================================================================================

### 2024-10-03 paper.tex
Tweak intro + abstract

@@
   interested parties and standardising the protocols that power the
-  Internet, such as IPv4/6 or HTTP/s. Our work covers more 20 years of IETF
+  Internet, such as IP or HTTP. Our work covers more 20 years of IETF
   data, and analyses the participants of the IETF, the emails they
@@
 
-% A good paper introduction is fairly formulaic. If you follow a simple set
-% of rules, you can write a very good introduction. The following outline can
-% be varied. For example, you can use two paragraphs instead of one, or you
-% can place more emphasis on one aspect of the intro than another. But in all
-% cases, all of the points below need to be covered in an introduction, and
-% in most papers, you don't need to cover anything more in an introduction.
-%
 % Paragraph 1: Motivation. At a high level, what is the problem area you
@@
 
+Network protocol standards are crucial for successful operation of the
+Internet.  A successful standard provides a basis for interoperability
+between systems developed by competing vendors, and supports the growth of
+an open ecosystem of products and services. Further, the process by which
+network protocol standards are developed, comprising multiple rounds of
+open feedback and review, has proven remarkably effective in designing
+high-quality and robust protocols, many of which see widespread deployment
+and use. Understanding the Internet standards development process, and how
+it produces successful protocols is, therefore, important if we are to
+understand the Internet and how it has evolved.
+
+One of the main organisations that develops protocol standards is the
+Internet Engineering Task Force (IETF). The IETF was founded in 1986,
+following on from the US Government-funded effort that developed the early
+Internet. It has since grown to become a global community of network
+protocol designers, vendors, network operators, and researchers that
+develop and publish open network protocol standards and operational
+guidelines.  The IETF publishes its standards, and other documents, in
+the RFC series.\footnote{\url{https://www.rfc-editor.org}} This series
+comprises around 9,000 documents, authored over more than 50 years,
+and provides a rich history of the development of the Internet and
+its protocols \cite{RFC8700}.
+
 % Paragraph 2: What is the specific problem considered in this paper? This
@@
 % establish specific context and background.
+
+Standardisation is an inherently social and political process
+\cite{RFC2026,RFC7282}, requiring cooperation and consensus among a growing
+number of stakeholders.
+
+It is important to gain a coherent understanding of the activities that
+take place within the IETF, as well as the key factors that may predict
+the success of a protocol standard.
 

### 2024-10-03 paper.tex
Authors and affiliations

@@
 \usepackage{multirow}
+\usepackage{authblk}
 
@@
        with Computational Methods}
-\author{Mladen Karan \and
-        Prashant Khare \and
-        Ravi Shekhar \and
-        Matthew Russell Barnes \and
-        Waleed Iqbal \and
-        Junaid Qadir \and
-        Stephen McQuistin \and
-        Richard G. Clegg \and
-        Patrick Healey \and
-        Matthew Purver \and
-        Ignacio Castro \and
-        Gareth Tyson \and
-        Colin Perkins}
+\author[1]{Mladen Karan}
+\author[1]{Prashant Khare}
+\author[2]{Ravi Shekhar}
+\author[1]{Matthew Russell Barnes}
+\author[1]{Waleed Iqbal}
+\author[3]{Junaid Qadir}
+\author[4]{Stephen McQuistin}
+\author[1]{Richard G. Clegg}
+\author[1]{Patrick Healey}
+\author[5]{Matthew Purver}
+\author[1]{Ignacio Castro}
+\author[6]{Gareth Tyson}
+\author[7]{Colin Perkins}
+
+\affil[1]{Queen Mary University of London}
+\affil[2]{University of Essex}
+\affil[3]{Information Technology University}
+\affil[4]{University of St Andrews}
+\affil[5]{Jozef Stefan Institute}
+\affil[6]{Hong Kong University of Science \& Technology}
+\affil[7]{University of Glasgow}
 

### 2024-10-04 scripts/fetch.sh
Fetch using curl directly

@@ -1,15 +1,0 @@
-#!/bin/sh
-
-if [ $# != 2 ]; then
-  echo "Usage: $0 <url> <outputfile>"
-  exit 1
-fi
-
-if [ -f $2 ]; then
-  echo "fetch $1 -> $2 (if changed)";
-  curl -L --progress-bar -o $2 -z $2 $1;
-else
-  echo "fetch $1 -> $2";
-  curl -L --progress-bar -o $2       $1;
-fi
-

### 2024-10-05 scripts/fetch-ietf-history-for-drafts.py
Update

@@ -1,0 +1,96 @@
+#!/usr/bin/env python3
+#
+# Copyright (c) 2024 University of Glasgow
+#
+# Redistribution and use in source and binary forms, with or without
+# modification, are permitted provided that the following conditions are met:
+#
+# 1. Redistributions of source code must retain the above copyright notice, this
+#    list of conditions and the following disclaimer.
+#
+#2. Redistributions in binary form must reproduce the above copyright notice,
+#   this list of conditions and the following disclaimer in the documentation
+#   and/or other materials provided with the distribution.
+#
+# THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS"
+# AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
+# IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE
+# DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE
+# FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL
+# DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR
+# SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER
+# CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY,
+# OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
+# OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+import sys
+
+from datetime import timedelta
+
+from ietfdata.datatracker     import *
+from ietfdata.datatracker_ext import *
+from ietfdata.rfcindex        import *
+
+def stream_acronym(dt, stream):
+    if stream is not None:
+        return dt.stream(stream).slug
+    else:
+        return None
+
+def group_acronym(dt, group):
+    if group is not None:
+        acronym = dt.group(group).acronym
+        if acronym == "none":
+            return None
+        else:
[…]

### 2024-10-05 scripts/drafts-by-date.py
Scripts to fetch drafts

@@ -1,0 +1,102 @@
+#!/usr/bin/env python3
+#
+# Copyright (c) 2024 University of Glasgow
+#
+# Redistribution and use in source and binary forms, with or without
+# modification, are permitted provided that the following conditions are met:
+#
+# 1. Redistributions of source code must retain the above copyright notice, this
+#    list of conditions and the following disclaimer.
+#
+#2. Redistributions in binary form must reproduce the above copyright notice,
+#   this list of conditions and the following disclaimer in the documentation
+#   and/or other materials provided with the distribution.
+#
+# THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS"
+# AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
+# IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE
+# DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE
+# FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL
+# DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR
+# SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER
+# CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY,
+# OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
+# OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+import json
+import sys
+
+from pathlib  import Path
+from datetime import date, timedelta
+
+if len(sys.argv) != 3:
+    print(f"Usage: {sys.argv[0]} history-for-drafts.json <output_file>")
+    sys.exit(1)
+
+with open(sys.argv[1], "r") as inf:
+    drafts = json.load(inf)
+
+results = {}
+
+oldest = None
+newest = None
+
+for draft in drafts:
+    prev_start = None
+    for h in draft["history"]:
+        name  = h['draft'] + "-" + h['revision']
[…]

### 2024-10-05 scripts/fetch-ietf-drafts.py

@@ -1,0 +1,93 @@
+#!/usr/bin/env python3
+#
+# Copyright (c) 2024 University of Glasgow
+#
+# Redistribution and use in source and binary forms, with or without
+# modification, are permitted provided that the following conditions are met:
+#
+# 1. Redistributions of source code must retain the above copyright notice, this
+#    list of conditions and the following disclaimer.
+#
+#2. Redistributions in binary form must reproduce the above copyright notice,
+#   this list of conditions and the following disclaimer in the documentation
+#   and/or other materials provided with the distribution.
+#
+# THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS"
+# AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
+# IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE
+# DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE
+# FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL
+# DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR
+# SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER
+# CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY,
+# OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
+# OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+import json
+import sys
+
+from datetime             import timedelta
+from ietfdata.datatracker import *
+
+def stream_acronym(dt, stream):
+    if stream is not None:
+        return dt.stream(stream).slug
+    else:
+        return None
+
+def group_acronym(dt, group):
+    if group is not None:
+        acronym = dt.group(group).acronym
+        if acronym == "none":
+            return None
+        else:
+            return acronym
+    else:
+        return None
+
+
+def path_for_draft(draft):
+    elem = draft.split("-")
+    if len(elem) < 3:
[…]

### 2024-10-05 scripts/fetch-ietf-dt.py
Fetch JSON files from the IETF Datatracker

@@ -1,0 +1,75 @@
+#!/usr/bin/env python3
+#
+# Copyright (c) 2024 University of Glasgow
+#
+# Redistribution and use in source and binary forms, with or without
+# modification, are permitted provided that the following conditions are met:
+#
+# 1. Redistributions of source code must retain the above copyright notice, this
+#    list of conditions and the following disclaimer.
+#
+#2. Redistributions in binary form must reproduce the above copyright notice,
+#   this list of conditions and the following disclaimer in the documentation
+#   and/or other materials provided with the distribution.
+#
+# THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS"
+# AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
+# IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE
+# DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE
+# FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL
+# DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR
+# SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER
+# CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY,
+# OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
+# OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+import json
+import os
+import requests
+import sys
+
+from typing import Any, Dict, Iterator
+
+
+def fetch_multi(session, dt_url, uri) -> Iterator[Dict[Any, Any]]:
+    while uri is not None:
+        r = session.get(f"{dt_url}{uri}")
+        if r.status_code == 200:
+            meta = r.json()['meta']
+            objs = r.json()['objects']
+            for obj in objs:
+                yield obj
+            uri = meta["next"]
+        else:
[…]

### 2024-10-06 scripts/drafts-by-date.py
Plot internet-drafts by date

@@ -59,44 +59,44 @@ for draft in drafts:
 
 output = {}
 
-curr = oldest
-prev_month = curr.isoformat()[0:7]
-month_total = 0
-month_count = 0
+curr = date.fromisoformat("1969-01-01")
+prev_year = curr.isoformat()[0:4]
+year_total = 0
+year_count = 0
 while curr < date.today():
-    month = curr.isoformat()[0:7]
-    if month not in output:
-        output[month] = {"month": month, "min": 99999, "max": 0, "avg": -1}
+    year = curr.isoformat()[0:4]
+    if year not in output:
+        output[year] = {"year": year, "min": 99999, "max": 0, "avg": -1}
 
-    if month != prev_month:
-        output[prev_month]["avg"] = int(month_total / month_count)
-        month_total = 0
-        month_count = 0
-    prev_month = month
+    if year != prev_year:
+        output[prev_year]["avg"] = int(year_total / year_count)
+        year_total = 0
+        year_count = 0
+    prev_year = year
 
     count = 0
     for index in results:
         if curr >= results[index]["start"] and curr <= results[index]["until"]:
             count += 1
 
-    month_total += count
-    month_count += 1
+    year_total += count
+    year_count += 1
 
-    if count < output[month]["min"]:
-        output[month]["min"] = count
-    if count > output[month]["max"]:
-        output[month]["max"] = count
+    if count < output[year]["min"]:
+        output[year]["min"] = count
+    if count > output[year]["max"]:
+        output[year]["max"] = count
 
     curr += timedelta(days = 1)
 
-output[prev_month]["avg"] = int(month_total / month_count)
+output[prev_year]["avg"] = int(year_total / year_count)
 
 
 with open(sys.argv[2], "w") as outf:
     for index in output.keys():
-        month_date = output[index]["month"]
-        month_min  = output[index]["min"]
-        month_avg  = output[index]["avg"]
-        month_max  = output[index]["max"]
-        print(f"{month_date},{month_min},{month_avg},{month_max}", file=outf)
+        year_date = output[index]["year"]
+        year_min  = output[index]["min"]
+        year_avg  = output[index]["avg"]
+        year_max  = output[index]["max"]
+        print(f"{year_date},{year_min},{year_avg},{year_max}", file=outf)
 

### 2024-10-06 scripts/plot-drafts-by-date.py

@@ -1,0 +1,78 @@
+#!/usr/bin/env python3
+#
+# Copyright (c) 2024 University of Glasgow
+#
+# Redistribution and use in source and binary forms, with or without
+# modification, are permitted provided that the following conditions are met:
+#
+# 1. Redistributions of source code must retain the above copyright notice, this
+#    list of conditions and the following disclaimer.
+#
+#2. Redistributions in binary form must reproduce the above copyright notice,
+#   this list of conditions and the following disclaimer in the documentation
+#   and/or other materials provided with the distribution.
+#
+# THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS"
+# AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
+# IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE
+# DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE
+# FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL
+# DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR
+# SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER
+# CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY,
+# OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
+# OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+import sys
+import pandas as pd
+import matplotlib.pyplot as plt
+import matplotlib.ticker as ticker
+
+from datetime import datetime
+ 
+if len(sys.argv) != 3:
+    print(f"Usage: python3 {sys.argv[0]} <datafile.csv> <output_file.pdf>")
+    sys.exit(1)
+
+csv_path = sys.argv[1]
+pdf_path = sys.argv[2]
+
+tics = []
+data = {}
+with open(csv_path, "r") as csv_file:
+    for line in csv_file.readlines():
[…]

### 2024-10-06 scripts/drafts-by-date.py
Tweak plot

@@ -63,7 +63,7 @@ curr = date.fromisoformat("1969-01-01")
 prev_year = curr.isoformat()[0:4]
 year_total = 0
 year_count = 0
-while curr < date.today():
+while curr < date.fromisoformat("2020-12-31"):
     year = curr.isoformat()[0:4]
     if year not in output:
         output[year] = {"year": year, "min": 99999, "max": 0, "avg": -1}

### 2024-10-06 scripts/plot-drafts-by-date.py

@@ -52,7 +52,7 @@ mins = [data[date][0] for date in dates]
 avgs = [data[date][1] for date in dates]
 maxs = [data[date][2] for date in dates]
 
-plt.rc('font',**{'family':'serif','serif':['Times'], 'size': 8})
+plt.rc('font',**{'family':'serif','serif':['Times'], 'size': 10})
 plt.rc('text', usetex=True)
 plt.rc('axes', axisbelow=True)
 plt.rcParams['pdf.fonttype'] = 42
@@ -63,7 +63,7 @@ plt.plot(dates, avgs, color="#2678b2", label="Mean")
 plt.fill_between(dates, mins, maxs, color="#2678b2", alpha=0.3, lw=0, label="Range")
 plt.gca().set_ylim(bottom=0)
 
-plt.xticks(tics, rotation='vertical')
+plt.xticks(tics)
 plt.xlabel("Year")
 plt.ylabel("Number of Active Internet-Drafts")
 
@@ -71,7 +71,7 @@ plt.legend(loc='upper left')
 
 # Annotate with date IETF created
 plt.plot([1986, 1986], [0, 1000], ':', color='black')
-plt.text(1982, 1100, 'IETF Created', color='black')
+plt.text(1981, 1100, 'IETF Created', color='black')
 
 plt.plot()
 plt.savefig(pdf_path, format="pdf", bbox_inches="tight")

### 2024-10-06 paper.tex
Tweak authors/affiliations
Fix affiliation
Discuss Internet-drafts

@@
 
-\title{An Empirical Analysis of the Internet Engineering Task Force 
-       with Computational Methods}
+\title{An Empirical Analysis of the Internet Engineering Task Force with Computational Methods}
+
 \author[1]{Mladen Karan}
@@
 \affil[2]{University of Essex}
-\affil[3]{Information Technology University}
+\affil[3]{Qatar University}
 \affil[4]{University of St Andrews}
-\affil[5]{Jozef Stefan Institute}
+\affil[5]{Jo\v{z}ef Stefan Institute}
 \affil[6]{Hong Kong University of Science \& Technology}
 \affil[7]{University of Glasgow}
+
+\date{7 October 2024}
 
@@
 
-\todo{Plot RFC publication count as a cumulative distribution}
-
-In addition to the published RFCs, we track work-in-progress documents,
-known as Internet-drafts, that are intended for eventual publication as
-RFCs.
-\todo{Add internet-draft counts over time as mentioned in Section \ref{sec:org-dyn:participants}}
-
 %..................................................................................................
@@
   \label{fig:pub_wgs_yearly}
+\end{figure}
+
+%..................................................................................................
+\pb{Internet-Drafts:}
+Figure \ref{fig:drafts-by-date} shows the number of working documents,
+known as Internet-drafts, that were active in the IETF over time. An
+Internet-draft is a limited lifetime working document that expires after
+six months, when replaced by a new version, or then published as an RFC.
+
+We observe that the number of active drafts correlates more strongly with
+the number of active working groups (Figure \ref{fig:pub_wgs_yearly}) than
+with RFC publication (Figure \ref{fig:rfcs_by_area}).  This is largely to
+be expected: an Internet-Draft is a work-in-progress document that is
+proposed for adoption by, and discussion in, an IETF working group prior to
+eventual publication as an RFC.  Competing proposals are often submitted as
+drafts, to allow a working group to review them and make a decision how to
+proceed, and hence not all drafts become RFCs. As we discuss in
+\S\ref{sec:success-factors:documents}, draft adoption is a significant
+milestone.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures/drafts-by-date.pdf}
+  \caption{
+    Number of active drafts over time. Internet-Drafts, working documents
+    of the IETF for work that was intended for eventual publication as an
+    RFC, were introduction at the 13th IETF meeting in 1989.
+  }
+  \label{fig:drafts-by-date}
 \end{figure}
@@
 \subsection{What Makes a Successful Author?}
+\label{sec:success-factors:authors}
+
 % ICWSM 2022 paper Section 4
@@
 \subsection{What Makes a Successful Document?}
+\label{sec:success-factors:documents}
 

### 2024-10-06 paper.tex
Expand discussion of drafts in §5.1

@@
 observe the entire volume of mentions.  We observe a strong increase in the
-number of mentions over time.  This is largely driven by the growing number
-of drafts being published. In fact, we find a Pearson correlation of 0.89
-between the number of drafts published and the number of mentions. This
-speaks to the influence of emails that mention drafts in driving draft
-production.
-\todo{Add and cross-reference a discussion of number of drafts published
-over time in Section \ref{sec:trends-documents}. Does this tie into the
-discussion of increasing complexity of draft publication?}
+number of mentions over time.
+
+% This is largely driven by the growing number of drafts being published. In
+% fact, we find a Pearson correlation of 0.89 between the number of drafts
+% published and the number of mentions. This speaks to the influence of
+% emails that mention drafts in driving draft production.
+
+When compared with the number of active drafts (Figure \ref{fig:drafts-by-date}
+in \S\ref{sec:trends-documents:production}) we see that the number of mentions
+of work-in-progress drafts and RFCs is increasing faster than the number of
+such documents. This further supports the hypothesis of increasing standards
+complexity---discussion is increasingly referencing prior standards and
+other work in progress, as would be expected if the task of ensuring
+compatibility with that work is growing in complexity.
 

### 2024-10-06 paper.tex
Introduction to §7

@@
 Section \ref{sec:success-factors} then considers factors that lead to
-success, exploring what makes a document likely to be completed and
-used, and what makes a successful participant.
+success, exploring what makes a document likely to be published as an
+RFC and what makes a successful participant.
 Finally, we review related work in Section \ref{sec:related} and conclude
@@
 
-\todo{introductory remarks}
+Finally, we consider factors that lead to success, exploring what makes a
+successful participant (\S\ref{sec:success-factors:authors}) and what makes
+a document likely to be successfully published as an RFC (\S\ref{sec:success-factors:documents}).
+
+Results show that influential authors (i.e., those with high centrality in
+the email social graph) and authors affiliated with organisations that
+participate strongly in IETF are more likely to have their documents
+adopted and published as RFCs. However, documents that build on existing
+work, have well-defined requirements, limited scope, and meet a need have
+improved chances of success with authorships playing less of a role. Good,
+well-connected, people are important, but so is focussed technical work.
 
@@
 
-Finally, the focus of our modelling in this paper has been on the impact of
+Finally, the focus of our modelling in this section has been on the impact of
 document, author, and interaction-based features on protocol deployment.

### 2024-10-06 paper.tex
Fix tables

@@
   \end{tabular}
-        \caption{Mean number of emails received for one year before/after becoming a WGC (AD) for the first time. The error is one standard deviation.}
+  \caption{
+    Mean number of emails received for one year before/after becoming a WGC
+    (AD) for the first time. The error is one standard deviation.
+  }
   \label{tab:WGC_email}
@@
 
-\begin{table*}[]
+\begin{table*}
+  \centering
   \small
-  \begin{center}
   \begin{tabular}{l|l|l}
@@
   \end{tabular}
-  \end{center}
   \caption{LIWC categories where $p<0.05$.}
@@
 
-\begin{table}
-  \centering
-  \begin{tabular}{lrrr}
-    \toprule
-      Model & $F_1$ & $\mathit{AUC}$ & $F1_{macro}$ \\
-    \midrule
-      Most frequent class                & .757   & .500   & .379 \\
-      Baseline                           & .758   & .616   & .597 \\
-      Baseline + FS                      & .762   & .650   & .610 \\
-    \midrule
-      Most frequent class                & .724   & .500   & .379 \\
-      Baseline                           & .670   & .559   & .547 \\
-      Baseline + FS                      & .690   & .620   & .563 \\
-      Logistic regression all feats      & .728   & .724   & .666 \\
-      Logistic regression all feats + FS & .820   & .822   & .789 \\
-      Decision tree all feats + FS       & .822   & .838   & .788 \\
-    \bottomrule
-  \end{tabular}
-  \caption{
-    Classifier scores on the entire dataset (251 RFCs, above) and those
-    with our features available (155 RFCs), with or without feature
-    selection (FS).
-  }
-  \label{tbl:mlres}
-\end{table}
-
 %..................................................................................................

### 2024-10-06 paper.tex
Abstract

@@
 \usepackage{authblk}
+\usepackage{hyperref}
 
@@
 
-  The Internet requires interoperability between networks, systems, and
-  applications, as well as cooperation among a growing number of
-  stakeholders. The Internet Engineering Task Force (IETF) is critical in
-  supporting this cooperation and interoperability by bringing together
-  interested parties and standardising the protocols that power the
-  Internet, such as IP or HTTP. Our work covers more 20 years of IETF
-  data, and analyses the participants of the IETF, the emails they
-  exchange, and the documents they produce. We show how these data can be
-  used to understand who are these Internet actors, and their interests as
-  well as how Internet's growth and maturity has given rise to a longer and
-  complex standardisation process where participants gain and exercise
-  influence and how this is reflected in the language they use and the
-  interactions they have.
+  Technical operation of the Internet requires interoperability between
+  networks, systems, and applications from different vendors as well as
+  cooperation among a growing number of stakeholders. The Internet
+  Engineering Task Force (IETF) is an open standards developing
+  organisation that plays a critical role in enabling this cooperation and
+  interoperability by bringing together interested parties to develop the
+  protocols that power the Internet, such as IP or HTTP. In this paper, we
+  review more 20 years of data about the IETF, studying who participates in
+  the IETF, the emails they exchange, and the documents they produce. We
+  show how these data can be used to understand who develops technical
+  standards for the Internet, and their interests, and how Internet's
+  growth and maturity has given rise to a longer and complex standards
+  development process. We explore where participants gain and exercise
+  influence, and how this is reflected in the language they use and in
+  their interactions.
+
+  This paper is a summary and collation of our prior published work,
+  that can be found in full in:
+  \cite{mcquistin:2021:characterising,khare:2022:web-we-weave,
+  healey:2023:power,mcquistin:2023:errare,khare:2023:tracing,
+  karan:2023:leda,healey:2023:power-frontiers,barnes:2024:temporal}.
 

### 2024-10-06 paper.tex
Introduction

@@
 
-% Paragraph 1: Motivation. At a high level, what is the problem area you
-% are working in and why is it important? It is important to set the larger
-% context here. Why is the problem of interest and importance to the larger
-% community?
-
 Network protocol standards are crucial for successful operation of the
@@
 its protocols \cite{RFC8700}.
-
-% Paragraph 2: What is the specific problem considered in this paper? This
-% paragraph narrows down the topic area of the paper. In the first
-% paragraph you have established general context and importance. Here you
-% establish specific context and background.
-
-Standardisation is an inherently social and political process
-\cite{RFC2026,RFC7282}, requiring cooperation and consensus among a growing
-number of stakeholders.
-
-It is important to gain a coherent understanding of the activities that
-take place within the IETF, as well as the key factors that may predict
-the success of a protocol standard.
-
-% Paragraph 3: "In this paper, we show that...". This is the key paragraph
-% in the introduction - you summarize, in one paragraph, what are the main
-% contributions of your paper, given the context established in paragraphs
-% 1 and 2. What's the general approach taken? Why are the specific results
-% significant? The story is not what you did, but rather:
-%  - what you show, new ideas, new insights
-%  - why interesting, important?
-% State your contributions: these drive the entire paper.  Contributions
-% should be refutable claims, not vague generic statements.
-
-In this paper, we ...
-
-% Paragraph 4: What are the differences between your work, and what others
-% have done? Keep this at a high level, as you can refer to future sections
-% where specific details and differences will be given, but it is important
-% for the reader to know what is new about this work compared to other work
-% in the area.
-
-This paper summarises and expands on our prior work
-\cite{mcquistin:2021:characterising,khare:2022:web-we-weave,
-healey:2023:power,mcquistin:2023:errare,khare:2023:tracing,
-karan:2023:leda,healey:2023:power-frontiers,barnes:2024:temporal}.
-
-% Paragraph 5: "We structure the remainder of this paper as follows." Give
-% the reader a road-map for the rest of the paper. Try to avoid redundant
-% phrasing, "In Section 2, In section 3, ..., In Section 4, ... ", etc.
-
-We structure the remainder of this paper as follows.
+This history is further augmented by the rich archive of supporting
+documents, work-in-progress drafts, email discussion, and other
+metadata about the standards development process that the IETF makes
+available.
+
+This gives a unique opportunity to explore the development of the Internet
+through its standards. Standards development is an inherently social and
+political process \cite{RFC2026,RFC7282}, requiring cooperation and
+consensus among a growing number of stakeholders. Due to the importance
+of the Internet, it's critical to gain a coherent understanding of the
+activities that take place within the IETF, to understand who develops
+Internet standards and what factors may predict the success of an idea
+or proposal for a new standard.
+
+In this paper, we explore the operation of the IETF to understand who
+develops Internet standards, how the participants interact, and what
+factors may lead to success.
 Section \ref{sec:background} introduces the IETF and the Internet Standards
@@
 complexity, and correctness.
-Section \ref{sec:trends-demographics} moves on to consider the people who
-are involved in Internet standards development and how the demographics of
-the community have shifted over time.
+item Section \ref{sec:trends-demographics} moves on to consider the people
+who are involved in Internet standards development and how the demographics
+of the community have shifted over time.
 Section \ref{sec:org-dyn} studies the interaction between participants,
-explores who has influence and impact in the IETF and how that influence
-is evident in their activities, and the dynamics of communications within
-the IETF.
+explores who has influence and impact in the IETF and how that influence is
+evident in their activities, and the dynamics of communications within the
+IETF.
 Section \ref{sec:language} considers communications within the IETF from
 the perspective of language to explore whether there are differences in
-linguistic patterns and language use in more influential participants,
-and to consider how the use of language reflects the consensus-driven
-nature of the standards development process.
+linguistic patterns and language use in more influential participants, and
+to consider how the use of language reflects the consensus-driven nature of
+the standards development process.
 Section \ref{sec:success-factors} then considers factors that lead to
-success, exploring what makes a document likely to be published as an
-RFC and what makes a successful participant.
+success, exploring what makes a document likely to be published as an RFC
+and what makes a successful participant.
 Finally, we review related work in Section \ref{sec:related} and conclude
 in Section \ref{sec:conclusions}.
+
+This paper is a summary and collation of our prior published work,
+that can be found in full in:
+\cite{mcquistin:2021:characterising,khare:2022:web-we-weave,
+healey:2023:power,mcquistin:2023:errare,khare:2023:tracing,
+karan:2023:leda,healey:2023:power-frontiers,barnes:2024:temporal}.
 

### 2024-10-06 paper.tex
Figure and Table placement and formatting; tweaks to text

@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures/drafts-by-date.pdf}
+  \caption{
+    Number of active drafts over time. Internet-Drafts, working documents
+    of the IETF for work that was intended for eventual publication as an
+    RFC, were introduction at the 13th IETF meeting in 1989.
+  }
+  \label{fig:drafts-by-date}
+\end{figure}
+
 %..................................................................................................
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures/drafts-by-date.pdf}
-  \caption{
-    Number of active drafts over time. Internet-Drafts, working documents
-    of the IETF for work that was intended for eventual publication as an
-    RFC, were introduction at the 13th IETF meeting in 1989.
-  }
-  \label{fig:drafts-by-date}
-\end{figure}
-
 %..................................................................................................
@@
 
-\begin{figure}
+\begin{figure}[t]
   \centering
@@
 \input{tables-prev/tma-2023/top-rfcs-errata.tex}
-\input{tables-prev/tma-2023/errata-stats-area.tex}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/errata-submission-dates-area.pdf}
+  \caption{
+    Cumulative distribution of the number of days from RFC publication to
+    errata filing by IETF area.
+    The areas are as described in Figure~\ref{fig:rfcs_by_area}.
+  }
+  \label{fig:errata_submission_days}
+\end{figure}
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/errata-submission-dates-area.pdf}
-  \caption{
-    Cumulative distribution of the number of days from RFC publication to
-    errata filing by IETF area.
-    The areas are as described in Figure~\ref{fig:rfcs_by_area}.
-  }
-  \label{fig:errata_submission_days}
-\end{figure}
-
 %..................................................................................................
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/errata-by-rfc-by-area.pdf}
+  \caption{
+    Cumulative distribution of errata filed per RFC by grouped by IETF
+    area.  The areas are as described in Figure~\ref{fig:rfcs_by_area}.
+  }
+  \label{fig:errata_per_rfc}
+\end{figure}
+
 %..................................................................................................
@@
 three notable outliers:
+
+\input{tables-prev/tma-2023/errata-stats-area.tex}
+
 \begin{enumerate}
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/errata-by-rfc-by-area.pdf}
-  \caption{
-    Cumulative distribution of errata filed per RFC by grouped by IETF
-    area.  The areas are as described in Figure~\ref{fig:rfcs_by_area}.
-  }
-  \label{fig:errata_per_rfc}
-\end{figure}
-
 Tables~\ref{tab:errata-stats-stream} and \ref{tab:errata-stats-status}
@@
 
+% \todo{Should we include ``Impact of Citations'' from TMA 2023?}
+
 \input{tables-prev/tma-2023/errata-stats-stream.tex}
@@
 
-% \todo{Should we include ``Impact of Citations'' from TMA 2023?}
+\begin{figure}
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/location-percent.pdf}
+  \caption{
+    Errata counts by percentile location in document (0 is the beginning;
+    100 is the end).
+  }
+  \label{fig:errata_location_percent_count}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/section-title-counts.pdf}
+  \caption{
+    Errata counts by section title for the more frequent section titles.
+  }
+  \label{fig:errata_section_wise_counts}
+\end{figure}
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/location-percent.pdf}
-  \caption{
-    Errata counts by percentile location in document (0 is the beginning;
-    100 is the end).
-  }
-  \label{fig:errata_location_percent_count}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/tma-2023/section-title-counts.pdf}
-  \caption{
-    Errata counts by section title for the more frequent section titles.
-  }
-  \label{fig:errata_section_wise_counts}
-\end{figure}
-
 %--------------------------------------------------------------------------------------------------
@@
 Huawei and ongoing shifts in the fortunes of many prominent US and European
-technology companies (\S\ref{sec:trends-demographics:affiliation}). The
-ongoing influence of Cisco and strong input from academic institutions is
-also noted.
+technology companies (\S\ref{sec:trends-demographics:affiliation}). 
+% The ongoing influence of Cisco and strong input from academic institutions
+% is also noted.
 
@@
 
-\begin{figure}
+\begin{figure}[t!]
   \centering
@@
   \label{fig:author_affiliations_normalised}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/authors/top5_affiliations_normalised_acad.pdf}
+  \caption{
+    Academic affiliations (normalised)
+  }
+  \label{fig:author_affiliations_normalised_acad}
 \end{figure}
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/authors/top5_affiliations_normalised_acad.pdf}
-  \caption{
-    Academic affiliations (normalised)
-  }
-  \label{fig:author_affiliations_normalised_acad}
-\end{figure}
-
 %..................................................................................................
@@
 working on is published.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/emails/pid_count_emailing_yearly.pdf}
-  \caption{
-    Number of email messages per year and number of unique participants
-    exchanging emails per year.
-  }
-  \label{fig:pid_count_emailing_yearly}
-\end{figure}
 
@@
 \end{enumerate}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/emails/pid_count_emailing_yearly.pdf}
+  \caption{
+    Number of email messages per year and number of unique participants
+    exchanging emails per year.
+  }
+  \label{fig:pid_count_emailing_yearly}
+\end{figure}
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/emails/rfc_authors_degree.pdf}
-  \caption{
-    CDF showing drift in annual degree (interaction with their network)
-    of RFC authors over the period 2000-2020.
-  }
-  \label{fig:degree_dist_rfc_authors}
-\end{figure}
-
 %..................................................................................................
@@
 the complexity, and these discussions are likely to lead to more drafts.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/emails/rfc_authors_degree.pdf}
+  \caption{
+    CDF showing drift in annual degree (interaction with their network)
+    of RFC authors over the period 2000-2020.
+  }
+  \label{fig:degree_dist_rfc_authors}
+\end{figure}
 
@@
 Section \ref{sec:org-dyn:participants} and, for each year, build a social
-graph based on the \emph{active community} of participants (nodes) who have
-email interactions (edges) with any other participant in the previous five
+graph based on the \emph{active community} of participants who have
+email interactions with any other participant in the previous five
 years.
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/lineplot_proportion_emailscount_top_percentile.pdf}
-  \caption{
-    Proportion of emails sent (\%) by participants in each period (with
-    95\% confidence interval) according to their betweenness-centrality
-    percentile (x-axis).  Y2-axis (purple) shows yearly count of emails.
-  }
-  \label{fig:emailcount_top_percentile}
-\end{figure}
-
 %..................................................................................................
@@
 
-\begin{figure}
+\begin{figure}[t!]
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/lineplot_proportion_emailscount_top_percentile.pdf}
+  \caption{
+    Proportion of emails sent (\%) by participants in each period (with
+    95\% confidence interval) according to their betweenness-centrality
+    percentile (x-axis).  Y2-axis (purple) shows yearly count of emails.
+  }
+  \label{fig:emailcount_top_percentile}
+\end{figure}
+
+\begin{figure}[t!]
   \centering
@@
   \label{fig:areas_participation_top_percentile}
+\end{figure}
+
+\begin{figure}[t!]
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/lineplot_age_top_percentile.pdf}
+  \caption{
+    Mean participation duration (with 95\% confidence interval) according
+    to their betweenness-centrality percentile in the x-axis, ranked from
+    top (0-10) to bottom percentile (91-100).
+  }
+  \label{fig:age_top_percentile}
 \end{figure}
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/lineplot_age_top_percentile.pdf}
-  \caption{
-    Mean participation duration (with 95\% confidence interval) according
-    to their betweenness-centrality percentile in the x-axis, ranked from
-    top (0-10) to bottom percentile (91-100).
-  }
-  \label{fig:age_top_percentile}
-\end{figure}
-
 We further consider whether participants who become influential, remain
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/heatmap_top10_percent_overlap.pdf}
-  \caption{
-    Yearly overlap (\%) of top 10\% influencers.
-  }
-  \label{fig:heatmap_top10_overlap}
-\end{figure}
-
 Finally, Figure \ref{fig:rate_new_influential_participants} shows that
@@
 \begin{figure}
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/heatmap_top10_percent_overlap.pdf}
+  \caption{
+    Yearly overlap (\%) of top 10\% influencers.
+  }
+  \label{fig:heatmap_top10_overlap}
+\end{figure}
+
+\begin{figure}[t!]
   \centering
@@
 time period by concatenating all messages sent by that participant in the
-period and feeding it into the LDA model. The result can be loosely
-interpreted as a distribution of topics on which each participant works.
+period and feeding it into the LDA model. The result can be 
+interpreted as a distribution of topics for each participant.
 
@@
 
-We next measure the topical diversity of a participant by observing the
-entropy of their topic vector, their \emph{topic entropy}, defined as
-follows:
+We next measure diversity of a participant via the entropy of their topic
+vector, their \emph{topic entropy}, defined as follows:
 \begin{equation}
@@
 topic and $\sum_i p_i = 1$.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/topic-div.png}
+  \caption{
+    Mean topic entropy (with 95\% confidence interval) according to their
+    betweenness-centrality percentile in the x-axis, ranked from top (0-10)
+    to bottom percentile (91-100).
+  }
+  \label{fig:topic_diversity}
+\end{figure}
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/topic-div.png}
-  \caption{
-    Mean topic entropy (with 95\% confidence interval) according to their
-    betweenness-centrality percentile in the x-axis, ranked from top (0-10)
-    to bottom percentile (91-100).
-  }
-  \label{fig:topic_diversity}
-\end{figure}
-
 % - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/coauthor_affiliation_network/lineplot_non_overlapping_authors_emailnetwork_percentile.pdf}
-  \caption{
-    Percentage of Top 20\% authors that are not 20\% influencers according
-    to their betweenness-centrality (email graph) percentile in the x-axis,
-    ranked from  top (0-10) to bottom percentile (91-100).
-  }
-  \label{fig:non_overlapping_coauthornetwork_authors_emailnetwork}
-\end{figure}
-
 %..................................................................................................
@@
 chair for the first time.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/WG_chair_percentile_CDF_before_after.pdf}
-  \caption{
-    CDF of the percentile of betweenness-centrality of WG chairs in the
-    email and co-authorship graph, one year before and after becoming
-    chairs.
-  }
-  \label{fig:wg_chair_before_after}
-\end{figure}
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/coauthor_affiliation_network/2019_affiliations_in_otheryears_coauthor_network.pdf}
-  \caption{
-    Authors in the most common affiliations of top 20\% authors
-    (co-authorship graph) in 2015-2019.
-  }
-  \label{fig:coauthor_affiliation_network2019_otheryears}
+\begin{figure}[t!]
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/coauthor_affiliation_network/lineplot_non_overlapping_authors_emailnetwork_percentile.pdf}
+  \caption{
+    Percentage of Top 20\% authors that are not 20\% influencers according
+    to their betweenness-centrality (email graph) percentile in the x-axis,
+    ranked from  top (0-10) to bottom percentile (91-100).
+  }
+  \label{fig:non_overlapping_coauthornetwork_authors_emailnetwork}
+\end{figure}
+
+\begin{figure}[t!]
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/WG_chair_percentile_CDF_before_after.pdf}
+  \caption{
+    CDF of the percentile of betweenness-centrality of WG chairs in the
+    email and co-authorship graph, one year before and after becoming
+    chairs.
+  }
+  \label{fig:wg_chair_before_after}
 \end{figure}
@@
 
-\begin{figure}
+\begin{figure}[t!]
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/coauthor_affiliation_network/2019_affiliations_in_otheryears_coauthor_network.pdf}
+  \caption{
+    Authors in the most common affiliations of top 20\% authors
+    (co-authorship graph) in 2015-2019.
+  }
+  \label{fig:coauthor_affiliation_network2019_otheryears}
+\end{figure}
+
+\begin{figure}[t!]
   \centering
@@
 to the suggestions from regular participants, and low confidence of those
-regular participants to voice their opinions. In contrast, we hypothesise
-that if the IETF operates as outlined in its mission statement, in a more
+regular participants to voice their opinions. In contrast, 
+if the IETF operates as outlined in its mission statement, in a more
 cooperative manner, then communication between those in different roles
@@
 
+\begin{figure*}[t!]
+  \centering
+  \includegraphics[width=\figureWidthTwoColumn]{figures-prev/icwsm-2024/motif_strata_comparison.png}
+  \caption{
+    Communication patterns analysed by three edge temporal motifs for RPs,
+  }
+  \label{fig:motif_proportions}
+\end{figure*}
+
 \pb{Impact of Role on Communication Patterns:}
-To quantify the effect of being in a leadership position has on network
+To quantify the effect being in a leadership position has on network
 wide communication, we calculate the proportion of three node temporal
 motifs, lasting at most a month, in which each node participates over a
-year period. These temporal motifs are counted, categorised into the three
+year period. These temporal motifs are counted, categorised into
 hierarchy levels (regular participant, working group chair, area director)
-and then proportions are taken of each temporal motif type.
+and then proportions are taken of each temporal motif.
 
@@
 and announcements from, the working group chairs and area directors.  This
-indicates some collaboration between layers and confidence of some regular
+indicates collaboration between layers and confidence of regular
 participants' to voice their opinions upwards.
-
-\begin{figure*}[t]
-  \centering
-  \includegraphics[width=\figureWidthTwoColumn]{figures-prev/icwsm-2024/motif_strata_comparison.png}
-  \caption{
-    Communication patterns analysed by three edge temporal motifs for RPs,
-  }
-  \label{fig:motif_proportions}
-\end{figure*}
 
@@
 patterns higher up the hierarchy.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/up_hierarchy_communication_ratios.png}
-  \caption{
-    Proportion of communications that are ``up'' the hierarchy for different
-    groups. A proportion above 0.5 indicates that the majority of
-    communications are ``upward''.
-  }
-  \label{fig:inter_strata}
-\end{figure}
 
@@
 
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/up_hierarchy_communication_ratios.png}
+  \caption{
+    Proportion of communications that are ``up'' the hierarchy for different
+    groups. A proportion above 0.5 indicates that the majority of
+    communications are ``upward''.
+  }
+  \label{fig:inter_strata}
+\end{figure}
+
 % - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
@@
 
+\begin{figure*}[t!]
+  \centering
+  \includegraphics[width=\figureWidthTwoColumn]{figures-prev/icwsm-2024/hierarchy_status_comparison.png}
+  \caption{
+    Connection between roles and activity for regular participants and
+    working group chairs viewed in terms of a) Mobility b) Neighbour
+    Mobility c) Philanthropy and d) Community over time.
+  }
+  \label{fig:mob_tax}
+\end{figure*}
+
 %..................................................................................................
@@
 are too noisy to show meaningful effects.
-
-\begin{figure*}[t]
-  \centering
-  \includegraphics[width=\figureWidthTwoColumn]{figures-prev/icwsm-2024/hierarchy_status_comparison.png}
-  \caption{
-    Connection between roles and activity for regular participants and
-    working group chairs viewed in terms of a) Mobility b) Neighbour
-    Mobility c) Philanthropy and d) Community over time.
-  }
-  \label{fig:mob_tax}
-\end{figure*}
 
@@
 
+\begin{figure*}[t!]
+  \centering
+  \includegraphics[width=\figureWidthTwoColumn]{figures-prev/frontiers/sensitive.pdf}
+  \caption{
+    Patterns of language use by organizational role. Estimated marginal
+    means for the LIWC categories: (A) Politics, (B) Ethnicity, and (C)
+    Religion. Error bars show 95% confidence intervals.
+  }
+  \label{fig:sensitive}
+\end{figure*}
+
 Results are shown in Figure \ref{fig:sensitive}. The consistent pattern
@@
 averages scores is zero
-
-\begin{figure*}[t]
-  \centering
-  \includegraphics[width=\figureWidthTwoColumn]{figures-prev/frontiers/sensitive.pdf}
-  \caption{
-    Patterns of language use by organizational role. Estimated marginal
-    means for the LIWC categories: (A) Politics, (B) Ethnicity, and (C)
-    Religion. Error bars show 95% confidence intervals.
-  }
-  \label{fig:sensitive}
-\end{figure*}
 
@@
 
-Regarding the text based features, we found that the models tend to assign
-high weights to words such as surnames of active contributors, names of
-prolific working groups, and common technical terms. While this does make
-the models perform slightly better, such terms should not be relevant for
-solving the task as they do not model the relevant part of the conversation.
-To explore how models would behave in a scenario wherein such words were
-not available, we construct a domain-specific stop word list, consisting
-of: (i) one all last names from the Datatracker, (ii) all working group
-names from the Datatracker, and (iii) technological jargon terms obtained
-from the web\footnote{\url{https://www.computerhope.com/jargon.htm}, we
-used a union of the \emph{Internet terms}, \emph{network terms}, and
-\emph{security terms} categories.}. We remove from this list any terms
-appearing in top 5,000 English terms (e.g., sometimes people's names can have
-identical surface forms as some of the terms relevant to organisational
-activities). We will refer to this variant as \emph{text (S)}.
-
-To gain empirical insights into which features provide better prediction
-results, models are run using a feature group alone variant and a variant
-combining each feature group with the \emph{text} features. This approach
-was motivated by the finding that \emph{text}, even though conceptually
-quite different from the rest of the feature set and not our main focus,
-does provide surprisingly strong results on its own. Thus, we wanted to
-empirically investigate in more detail how well it complements the rest of
-the graph-based features. Moreover, we compare all models to a baseline
-which simply assigns all examples to the majority class.
-
-We split the data into \emph{training} (70\%), \emph{development} (10\%),
-and \emph{test} (20\%) subsets. As a scoring function, we use the F1 score
-macro averaged across the positive and negative classes (but we also report
-area under the curve (AUC), precision, and recall). We train the models on
-the training set, optimise hyper-parameters on the development set, and
-report final scores on the test set. The final scores are those obtained by
-the model variant that fared best on the development set. We use logistic
-regression implemented in \texttt{scikit-learn} \cite{scikit-learn}, given
-that it is widely used and well interpretable. While we do not perform
-explicit feature selection, we do have implicit feature selection through
-L1 regularisation. We consider the regularisation strength a hyper-parameter
-and consider values from $[2^{-7}, 2^{-6}, ..., 2^6]$ on the development
-set. As our data set is quite imbalanced (in a roughly 4:1 ratio, 17\% is
-\emph{adopted}), we use different class weights during optimization to
-counteract this. We employ a non-parametric random shuffling test
-\cite{yeh-2000-accurate} to check statistical significance of score
-differences against the baseline. Table~\ref{tbl:res} summarises prediction
-results, which will be discussed in the next section.
-
 \renewcommand{\tabcolsep}{0.5cm}
-\begin{table*}
+\begin{table*}[t!]
 \centering
@@
 
+Regarding the text based features, we found that the models tend to assign
+high weights to words such as surnames of active contributors, names of
+prolific working groups, and common technical terms. While this does make
+the models perform slightly better, such terms should not be relevant for
+solving the task as they do not model the relevant part of the conversation.
+To explore how models would behave in a scenario wherein such words were
+not available, we construct a domain-specific stop word list, consisting
+of: (i) one all last names from the Datatracker, (ii) all working group
+names from the Datatracker, and (iii) technological jargon terms obtained
+from the web\footnote{\url{https://www.computerhope.com/jargon.htm}, we
+used a union of the \emph{Internet terms}, \emph{network terms}, and
+\emph{security terms} categories.}. We remove from this list any terms
+appearing in top 5,000 English terms (e.g., sometimes people's names can have
+identical surface forms as some of the terms relevant to organisational
+activities). We will refer to this variant as \emph{text (S)}.
+
+To gain empirical insights into which features provide better prediction
+results, models are run using a feature group alone variant and a variant
+combining each feature group with the \emph{text} features. This approach
+was motivated by the finding that \emph{text}, even though conceptually
+quite different from the rest of the feature set and not our main focus,
+does provide surprisingly strong results on its own. Thus, we wanted to
+empirically investigate in more detail how well it complements the rest of
+the graph-based features. Moreover, we compare all models to a baseline
+which simply assigns all examples to the majority class.
+
+We split the data into \emph{training} (70\%), \emph{development} (10\%),
+and \emph{test} (20\%) subsets. As a scoring function, we use the F1 score
+macro averaged across the positive and negative classes (but we also report
+area under the curve (AUC), precision, and recall). We train the models on
+the training set, optimise hyper-parameters on the development set, and
+report final scores on the test set. The final scores are those obtained by
+the model variant that fared best on the development set. We use logistic
+regression implemented in \texttt{scikit-learn} \cite{scikit-learn}, given
+that it is widely used and well interpretable. While we do not perform
+explicit feature selection, we do have implicit feature selection through
+L1 regularisation. We consider the regularisation strength a hyper-parameter
+and consider values from $[2^{-7}, 2^{-6}, ..., 2^6]$ on the development
+set. As our data set is quite imbalanced (in a roughly 4:1 ratio, 17\% is
+\emph{adopted}), we use different class weights during optimization to
+counteract this. We employ a non-parametric random shuffling test
+\cite{yeh-2000-accurate} to check statistical significance of score
+differences against the baseline. Table~\ref{tbl:res} summarises prediction
+results, which will be discussed in the next section.
+
 Next, we wanted to confirm these results at the level of individual
@@
 
+\renewcommand{\tabcolsep}{0.2cm}
 \input{tables-prev/imc-2021/lr_fsnone-trim.tex}

### 2024-10-06 paper.tex
Balance columns
Related work

@@
 \usepackage{hyperref}
+\usepackage{balance}
 
@@
 
+%..................................................................................................
+% IMC 2021 paper:
+
+To the best of our knowledge, were the first to perform a comprehensive
+statistical analysis of IETF activities considering trends in document
+publication (\S\ref{sec:trends-documents}), author demographics
+(\S\ref{sec:trends-demographics}), and communication patterns.
+\S\ref{sec:org-dyn:participants}. That said, there have been several
+prior efforts that have focused on specific aspects of the IETF:
+
+\begin{itemize}
+  \item Jari Arkko, a former Chair of the IETF, maintains a website
+    (\url{https://www.arkko.com/tools/stats.html}) that provides various
+    statistics about the IETF, including about its documents, authors and
+    their affiliations. The Datatracker also provides some statistics on
+    documents and authors (\url{https://datatracker.ietf.org/stats/}).
+    Attendance statistics are reported in the IETF plenaries (available
+    in the proceedings at \url{https://www.ietf.org/meeting/past/}).
+
+  \item BigBang~\cite{benthall2015testing} is a Python toolkit for
+    analysing online collaborative communities through mailing list data.
+    Niedermayer et al.~\cite{niedermayer2017information} discuss the
+    challenges of working with large mailing list datasets, including
+    conducting  entity resolution where contributors have changeable
+    pseudonyms. Our work addresses a number of these challenges by
+    augmenting the mailing list data with metadata from the IETF
+    Datatracker.
+
+  \item Huitema~\cite{RFC8963} carried out an evaluation of a small subset
+    of RFCs, to understand the various sources of delay in their
+    publication. The author observed that the main source of delay was the
+    working group process, highlighting the need to better understand the
+    dynamics of that process. In addition, the author found that citation
+    counts did not correlate well with the adoption or deployment of
+    specific RFCs.
+\end{itemize}
+
+%..................................................................................................
+% ICWSM 2022 paper:
+
+Our work to characterise participant influence and impact (\S\ref{sec:org-dyn:influence})
+builds on a rich literature regarding influence in online social networks
+and other communities:
+\begin{itemize}
+  \item 
+    Studies of Twitter~\cite{cha2010measuring,weitzel2012measuring,anger2011measuring},
+    Instagram~\cite{zarei2020characterising} and decentralised social networks
+    \cite{harissigmetrics, hassanpleroma}. For example, \cite{ye2010measuring}
+    measured the impact of follower influence on message propagation. Others
+    have focused on devising metrics to capture influence: similarly
+    to~\cite{kourtellis2013identifying,weitzel2012measuring,sole2014centrality},
+    we use betweenness centrality as a metric of influence.
+
+  \item Online collaborative communities such as Wikimedia and Open Stack
+    have also been studied for understanding communities, collaboration,
+    strategy making, and organisational structures, through mailing lists
+    and platform interactions \cite{dobusch2018open, bosu2014social,
+    zhang2020companies}. Bosou and Carver \cite{bosu2014impact} highlight
+    that contributors with much higher reputation are successful in seeking
+    reviews from the community in a much shorter time span and are more
+    likely to get their suggestions accepted. This is well aligned with
+    most of our findings related to the impact of influential people on the
+    Internet-Draft adoption process. Zhang et al.\ observe
+    \cite{zhang2020companies} that in the open source software (OSS)
+    ecosystem, several participating organisations (firms or companies) may
+    engage in intentional or passive collaborations, or they may contribute
+    in an isolated way. They find that an organisation's influence in the
+    collaboration network is positively correlated with its scale of
+    contribution within the ecosystem. In some ways, this is similar to our
+    observation related to prominent affiliated organisations in the
+    co-authorship network and their ability to collaborate and produce
+    drafts.
+
+  % This bullet adapted from our ICWSM 2024 paper
+  \item Building social interaction networks out of email communication
+    data is a well trodden field~\cite{belanger1999communication,
+    mcpherson2001birds}, and the introduction of time evolution into the
+    analysis~\cite{panzarasa2009patterns,viswanath2009evolution} is recent
+    but has quickly gained interest. However, the analysis of email based
+    communication is less prevalent with the notable exception of the
+    well-known, and oft analysed, Enron emails~\cite{klimt2004enron}. The
+    communication dataset we examine is interesting for several reasons: it
+    is a long duration email dataset, it is annotated with meta data
+    representing participant roles, and it represents discussion in a
+    decentralised consensus driven organisation~\cite{RFC7282,RFC3935}
+    rather than a business with a traditional management hierarchy.
+    While the Enron data set is often used for classification of emails
+    into categories \cite{madjarov2012extensive}, there are also several
+    studies of organisation interaction patterns on it.  Namely,
+    \cite{tang2008community,kossinets2008structure} are concerned with
+    developing methodologies for studying how communication evolves over
+    time.  In contrast, \cite{diesner2005communication} focus on providing
+    insights more than on developing a methodology, revealing that
+    organisational structure is reflected in  email communication patterns.
+    Other studies focusing on mailing list data include OSS mailing lists
+    \cite{bird2006mining,rigby2007can}.
+
+\end{itemize}
+
+%..................................................................................................
+% ICWSM 2024 paper
+
+Analysing the temporal patterns in communication, as we do for the IETF
+organisational hierarchy in \S\ref{sec:org-dyn:hierarchy}, is an emerging
+field of research~\cite{masuda2016guide}:
+
+\begin{itemize}
+  \item Some have analysed the time evolution of important nodes in a
+    network~\cite{fire2020rise}, whether their neighbours are similarly
+    important~\cite{pedreschi2022temporal} and some others the correlations
+    of the trajectory of snapshots of networks over
+    time~\cite{lacasa2022correlations}. In this paper we bring new analysis
+    techniques, that expand on the idea of tracking the evolution of node
+    importance, to an email list corpus.
+
+  \item A form of temporal network analysis that is key to understanding
+    communication in social networks is computing temporal
+    motifs~\cite{paranjape2017motifs,kovanen2011temporal} between nodes.
+    This technique involves look at the sequence in which edges attach to
+    nodes over time which lets us see how conversation happens with fine
+    resolution.  Applications include understanding the diffusion of
+    information through a social network using undirected
+    motifs~\cite{sarkar2019using}, and analysing the collaboration and
+    scientific mobility of co-authorship
+    networks~\cite{boekhout2021investigating}. To the best of our
+    knowledge, temporal motifs have not yet been used for analysis of
+    communication in social networks which means knowledge of high
+    time-resolution direct communication between individuals is ripe for
+    the picking.
+
+  \item The effects of hierarchy on the ``voice" of individuals within an
+    organisation is a key aspect~\cite{pfrombeck2023hierarchy} of
+    Organisational Behaviour (OB) research. A steep~\cite{hussain2019voice}
+    hierarchy can have a negative effect on voice by reducing the variety
+    of perspectives~\cite{anderson2010functions} in higher levels, or
+    positive effect where decisions can be made quickly. However, a
+    diffused hierarchy may also have negative
+    effects~\cite{hussain2019voice} on voice by contributing to lower level
+    ``bystanders" who let the higher levels do the talking, or a positive
+    effect with a variety of approaches to management. Similarly, a large
+    ``power distance" can cause problems to lower level voice if the higher
+    levels are ``authoritarian"~\cite{duan2018authoritarian}, however if
+    they are ``benevolent"~\cite{li2021does,guo2020inclusive} then power
+    distance is less of a problem. We aim to test these ideas in regard to
+    the voluntary and collaborative nature of the IETF organisation,
+    whereas the literature mostly focuses on traditional organisations.
+
+\end{itemize}
+
+%..................................................................................................
+
+% The following is adapted from our ACL 2023 Tracing Linguistic Markers of
+% Influence paper:
+
+Considering use of language (\S\ref{sec:language}), most studies of
+influence either focus on community structure rather than language, or use
+language indirectly.  Urena et al.\ \cite{urena2019review} give a survey of
+the former approach. In an example of the latter, Prabhakaran
+\cite{prabhakaran2014staying} compares users with different influence in
+terms of their linguistic similarity or co-adaptation, the increasing
+similarity of interlocutors to each other in how they use language (see
+also \cite{danescu2012echoes,ver2013information,noble2015centre,
+kawabata2016convex,buske2019linguistic}).  
+Some studies focus on modelling influence from text of Enron emails by
+identifying keywords/phrases that indicate influence \cite{bramsen2011extracting,
+gilbert2012phrases}.  Rosenthal \cite{rosenthal2014detecting} and Tchokni
+et al.\ \cite{tchokni2014emoticons} extend this approach to other domains,
+including Twitter, Wikipedia talk pages, and debates, and include a wider
+range of linguistic markers.
+
+%..................................................................................................
+The work in \S\ref{sec:success-factors} builds, in part, on Nikkhah et
+al.~\cite{nikkhah2017statistical} by statistically exploring RFC adoption.
+We build on this work with improved modelling, and by incorporating
+additional features that characterise not only the RFCs themselves but also
+the standardisation process that produced them.
+
 %==================================================================================================
@@
 %==================================================================================================
+\balance
 \bibliographystyle{abbrv}

### 2024-10-06 paper.tex
Related Work

@@
 
-% This should come near the end, and focussing on discussing how your work
-% relates to that of others. Any relevant related work should have been
-% cited already, so this is not a list of related work, it's a discussion
-% of how that work relates.
-%
-% Why not put related work after the introduction? 1) because describing
-% alternative approaches gets between the reader and your idea; and 2)
-% because the reader knows nothing about the problem yet, so your
-% (carefully trimmed) description of various technical trade-offs is
-% absolutely incomprehensible.
-%
-% When writing the related work:
-%  - Give credit to others where it's due; this doesn't diminish the
-%    credit you get from your paper.
-%  - Acknowledge weaknesses in your approach.
-%  - Ensure related work is accurate and up-to-date
-
 %..................................................................................................
@@
 
+\pb{Documents, Demographics, Communication Patterns:}
 To the best of our knowledge, were the first to perform a comprehensive
@@
 prior efforts that have focused on specific aspects of the IETF:
-
-\begin{itemize}
-  \item Jari Arkko, a former Chair of the IETF, maintains a website
+Jari Arkko, a former Chair of the IETF, maintains a website
 (\url{https://www.arkko.com/tools/stats.html}) that provides various
-    statistics about the IETF, including about its documents, authors and
-    their affiliations. The Datatracker also provides some statistics on
-    documents and authors (\url{https://datatracker.ietf.org/stats/}).
-    Attendance statistics are reported in the IETF plenaries (available
-    in the proceedings at \url{https://www.ietf.org/meeting/past/}).
-
-  \item BigBang~\cite{benthall2015testing} is a Python toolkit for
-    analysing online collaborative communities through mailing list data.
-    Niedermayer et al.~\cite{niedermayer2017information} discuss the
-    challenges of working with large mailing list datasets, including
-    conducting  entity resolution where contributors have changeable
-    pseudonyms. Our work addresses a number of these challenges by
-    augmenting the mailing list data with metadata from the IETF
-    Datatracker.
-
-  \item Huitema~\cite{RFC8963} carried out an evaluation of a small subset
-    of RFCs, to understand the various sources of delay in their
-    publication. The author observed that the main source of delay was the
-    working group process, highlighting the need to better understand the
-    dynamics of that process. In addition, the author found that citation
-    counts did not correlate well with the adoption or deployment of
-    specific RFCs.
-\end{itemize}
+statistics about the IETF, including about its documents, authors and their
+affiliations. The Datatracker also provides some statistics on documents
+and authors (\url{https://datatracker.ietf.org/stats/}).  Attendance
+statistics are reported in the IETF plenaries (available in the proceedings
+at \url{https://www.ietf.org/meeting/past/}).
+
+BigBang~\cite{benthall2015testing} is a Python toolkit for analysing online
+collaborative communities through mailing list data.  Niedermayer et
+al.~\cite{niedermayer2017information} discuss the challenges of working
+with large mailing list datasets, including conducting  entity resolution
+where contributors have changeable pseudonyms. Our work addresses a number
+of these challenges by augmenting the mailing list data with metadata from
+the IETF Datatracker.
+
+Huitema~\cite{RFC8963} carried out an evaluation of a small subset of RFCs,
+to understand the various sources of delay in their publication. The author
+observed that the main source of delay was the working group process,
+highlighting the need to better understand the dynamics of that process. In
+addition, the author found that citation counts did not correlate well with
+the adoption or deployment of specific RFCs.
 
@@
 
+\pb{Influence and Impact:}
 Our work to characterise participant influence and impact (\S\ref{sec:org-dyn:influence})
 builds on a rich literature regarding influence in online social networks
-and other communities:
-\begin{itemize}
-  \item 
+and other communities.
 Studies of Twitter~\cite{cha2010measuring,weitzel2012measuring,anger2011measuring},
@@
 
-  \item Online collaborative communities such as Wikimedia and Open Stack
-    have also been studied for understanding communities, collaboration,
-    strategy making, and organisational structures, through mailing lists
-    and platform interactions \cite{dobusch2018open, bosu2014social,
-    zhang2020companies}. Bosou and Carver \cite{bosu2014impact} highlight
-    that contributors with much higher reputation are successful in seeking
-    reviews from the community in a much shorter time span and are more
-    likely to get their suggestions accepted. This is well aligned with
-    most of our findings related to the impact of influential people on the
-    Internet-Draft adoption process. Zhang et al.\ observe
-    \cite{zhang2020companies} that in the open source software (OSS)
-    ecosystem, several participating organisations (firms or companies) may
-    engage in intentional or passive collaborations, or they may contribute
+Online collaborative communities such as Wikimedia and Open Stack have also
+been studied for understanding communities, collaboration, strategy making,
+and organisational structures, through mailing lists and platform
+interactions \cite{dobusch2018open, bosu2014social, zhang2020companies}.
+Bosou and Carver \cite{bosu2014impact} highlight that contributors with
+much higher reputation are successful in seeking reviews from the community
+in a much shorter time span and are more likely to get their suggestions
+accepted. This is well aligned with most of our findings related to the
+impact of influential people on the Internet-Draft adoption process. Zhang
+et al.\ observe \cite{zhang2020companies} that in the open source software
+(OSS) ecosystem, several participating organisations (firms or companies)
+may engage in intentional or passive collaborations, or they may contribute
 in an isolated way. They find that an organisation's influence in the
@@
 observation related to prominent affiliated organisations in the
-    co-authorship network and their ability to collaborate and produce
-    drafts.
+co-authorship network and their ability to collaborate and produce drafts.
 
 % This bullet adapted from our ICWSM 2024 paper
-  \item Building social interaction networks out of email communication
-    data is a well trodden field~\cite{belanger1999communication,
-    mcpherson2001birds}, and the introduction of time evolution into the
-    analysis~\cite{panzarasa2009patterns,viswanath2009evolution} is recent
-    but has quickly gained interest. However, the analysis of email based
+
+Building social interaction networks out of email communication data is a
+well trodden field~\cite{belanger1999communication, mcpherson2001birds},
+and the introduction of time evolution into the
+analysis~\cite{panzarasa2009patterns,viswanath2009evolution} is recent but
+has quickly gained interest. However, the analysis of email based
 communication is less prevalent with the notable exception of the
 well-known, and oft analysed, Enron emails~\cite{klimt2004enron}. The
-    communication dataset we examine is interesting for several reasons: it
-    is a long duration email dataset, it is annotated with meta data
-    representing participant roles, and it represents discussion in a
-    decentralised consensus driven organisation~\cite{RFC7282,RFC3935}
-    rather than a business with a traditional management hierarchy.
-    While the Enron data set is often used for classification of emails
-    into categories \cite{madjarov2012extensive}, there are also several
-    studies of organisation interaction patterns on it.  Namely,
+communication dataset we examine is interesting for several reasons: it is
+a long duration email dataset, it is annotated with meta data representing
+participant roles, and it represents discussion in a decentralised
+consensus driven organisation~\cite{RFC7282,RFC3935} rather than a business
+with a traditional management hierarchy.  While the Enron data set is often
+used for classification of emails into categories
+\cite{madjarov2012extensive}, there are also several studies of
+organisation interaction patterns on it.  Namely,
 \cite{tang2008community,kossinets2008structure} are concerned with
-    developing methodologies for studying how communication evolves over
-    time.  In contrast, \cite{diesner2005communication} focus on providing
-    insights more than on developing a methodology, revealing that
-    organisational structure is reflected in  email communication patterns.
-    Other studies focusing on mailing list data include OSS mailing lists
+developing methodologies for studying how communication evolves over time.
+In contrast, \cite{diesner2005communication} focus on providing insights
+more than on developing a methodology, revealing that organisational
+structure is reflected in  email communication patterns.  Other studies
+focusing on mailing list data include OSS mailing lists
 \cite{bird2006mining,rigby2007can}.
 
-\end{itemize}
-
 %..................................................................................................
@@
 
+\pb{Temporal Patterns in Communication:}
 Analysing the temporal patterns in communication, as we do for the IETF
@@
 
-\begin{itemize}
-  \item Some have analysed the time evolution of important nodes in a
+Some have analysed the time evolution of important nodes in a
 network~\cite{fire2020rise}, whether their neighbours are similarly
-    important~\cite{pedreschi2022temporal} and some others the correlations
-    of the trajectory of snapshots of networks over
+important~\cite{pedreschi2022temporal} and some others the correlations of
+the trajectory of snapshots of networks over
 time~\cite{lacasa2022correlations}. In this paper we bring new analysis
@@
 
-  \item A form of temporal network analysis that is key to understanding
+A form of temporal network analysis that is key to understanding
 communication in social networks is computing temporal
-    motifs~\cite{paranjape2017motifs,kovanen2011temporal} between nodes.
-    This technique involves look at the sequence in which edges attach to
-    nodes over time which lets us see how conversation happens with fine
-    resolution.  Applications include understanding the diffusion of
-    information through a social network using undirected
-    motifs~\cite{sarkar2019using}, and analysing the collaboration and
-    scientific mobility of co-authorship
-    networks~\cite{boekhout2021investigating}. To the best of our
-    knowledge, temporal motifs have not yet been used for analysis of
-    communication in social networks which means knowledge of high
-    time-resolution direct communication between individuals is ripe for
-    the picking.
-
-  \item The effects of hierarchy on the ``voice" of individuals within an
+motifs~\cite{paranjape2017motifs,kovanen2011temporal} between nodes.  This
+technique involves look at the sequence in which edges attach to nodes over
+time which lets us see how conversation happens with fine resolution.
+Applications include understanding the diffusion of information through a
+social network using undirected motifs~\cite{sarkar2019using}, and
+analysing the collaboration and scientific mobility of co-authorship
+networks~\cite{boekhout2021investigating}. To the best of our knowledge,
+temporal motifs have not yet been used for analysis of communication in
+social networks which means knowledge of high time-resolution direct
+communication between individuals is ripe for the picking.
+
+The effects of hierarchy on the ``voice" of individuals within an
 organisation is a key aspect~\cite{pfrombeck2023hierarchy} of
 Organisational Behaviour (OB) research. A steep~\cite{hussain2019voice}
-    hierarchy can have a negative effect on voice by reducing the variety
-    of perspectives~\cite{anderson2010functions} in higher levels, or
-    positive effect where decisions can be made quickly. However, a
-    diffused hierarchy may also have negative
-    effects~\cite{hussain2019voice} on voice by contributing to lower level
-    ``bystanders" who let the higher levels do the talking, or a positive
-    effect with a variety of approaches to management. Similarly, a large
-    ``power distance" can cause problems to lower level voice if the higher
-    levels are ``authoritarian"~\cite{duan2018authoritarian}, however if
-    they are ``benevolent"~\cite{li2021does,guo2020inclusive} then power
-    distance is less of a problem. We aim to test these ideas in regard to
-    the voluntary and collaborative nature of the IETF organisation,
-    whereas the literature mostly focuses on traditional organisations.
-
-\end{itemize}
-
-%..................................................................................................
-
+hierarchy can have a negative effect on voice by reducing the variety of
+perspectives~\cite{anderson2010functions} in higher levels, or positive
+effect where decisions can be made quickly. However, a diffused hierarchy
+may also have negative effects~\cite{hussain2019voice} on voice by
+contributing to lower level ``bystanders" who let the higher levels do the
+talking, or a positive effect with a variety of approaches to management.
+Similarly, a large ``power distance" can cause problems to lower level
+voice if the higher levels are
+``authoritarian"~\cite{duan2018authoritarian}, however if they are
+``benevolent"~\cite{li2021does,guo2020inclusive} then power distance is
+less of a problem. We aim to test these ideas in regard to the voluntary
+and collaborative nature of the IETF organisation, whereas the literature
+mostly focuses on traditional organisations.
+
+%..................................................................................................
 % The following is adapted from our ACL 2023 Tracing Linguistic Markers of
@@
 
+\pb{Use of Language:}
 Considering use of language (\S\ref{sec:language}), most studies of
@@
 %..................................................................................................
+\pb{Success Factors:}
 The work in \S\ref{sec:success-factors} builds, in part, on Nikkhah et

### 2024-10-06 paper.tex
Conclusions

@@
 complexity, and correctness.
-item Section \ref{sec:trends-demographics} moves on to consider the people
+Section \ref{sec:trends-demographics} moves on to consider the people
 who are involved in Internet standards development and how the demographics
@@
 
+We explored the key standardisation activities within the IETF, considering
+trends in document production, complexity, and correctness
+(\S\ref{sec:trends-documents}); the people involved in Internet standards
+development and how the demographics of the community have shifted over
+time (\S\ref{sec:trends-demographics}); interaction between participants,
+who has influence and impact in the IETF and how that influence is evident
+in their activities, and the dynamics of communications within the IETF
+(\S\ref{sec:org-dyn}); use of language, to explore whether there are
+differences in linguistic patterns and language use in more influential
+participants, and to consider how the use of language reflects the
+consensus-driven nature of the standards development process
+(\S\ref{sec:language}); and factors that lead to success, exploring what
+makes a document likely to be published as an RFC and what makes a
+successful participant (\S\ref{sec:success-factors}).
+
+In the process, we have revealed a vibrant ecosystem, albeit one with a
+number of key challenges:
+\begin{enumerate}
+  \item The rate of standards development has slowed over time and the IETF
+    is showing signs of struggling with managing the complexity of
+    standards development.
+
+  \item While authorship is becoming more diverse as the IETF grows, there
+    are still significant issues with representation from certain regions,
+    including Africa and South America. If the IETF is to become more
+    representative, then further effort is required to build more diverse
+    engagement, both in terms of geography, organisational affiliation, and
+    gender diversity \cite{kaeo:2023:women}.
+
+  \item Despite significant growth over time, a relatively small subset of
+    well-connected and highly influential participants remain an important
+    driving force in the IETF. The community is growing better connected,
+    and has become less fragmented, over time, and is growing in resilience
+    to departure of influential participants, but there is more to be done.
+
+\end{enumerate}
+
+On the positive side, we observed that influential people in IETF exhibit
+collaborative and community-oriented traits, and also stronger signs of
+engagement in discussions. We also observed that as people go on to become
+influential participants, they evolve in their communication and are seen
+to be more engaging and descriptive in their linguistic style and patterns
+of communication.  The IETF has a reputation for being a difficult
+community to engage with (``loud men, talking loudly'') that is not
+altogether unwarranted \cite{cath:2023:loud,cath:2021:the-technology}, but
+there is also evidence that those in positions of influence, and leadership
+roles, are broadly cooperative and supportive of newcomers and those with
+less experience, within their own cultural framework.
+
+Further research is needed to fully understand the IETF and its role in the
+Internet governance ecosystem, but our analysis to date shows a complex and
+evolving organisation---with some expected growing pains---that is meeting
+the ongoing challenges of developing technical standards for the modern,
+global, Internet.
+
 %==================================================================================================

### 2024-10-06 paper.tex
Metadata

@@
 \usepackage{authblk}
+\usepackage{balance}
 \usepackage{hyperref}
-\usepackage{balance}
+\hypersetup{pdfinfo={
+  Title={An Empirical Analysis of the Internet Engineering Task Force with
+         Computational Methods},
+  Author={Mladen Karan, Prashant Khare, Ravi Shekhar, Matthew Russell Barnes,
+          Waleed Iqbal, Junaid Qadir, Stephen McQuistin, Richard G. Clegg,
+          Patrick Healey, Matthew Purver, Ignacio Castro, Gareth Tyson, and
+          Colin Perkins}
+}}
 
@@
 
-\date{7 October 2024}
+\date{\emph{Working Draft} -- 7 October 2024}
 
@@
 \bibliography{paper}
-
-%==================================================================================================
-% The following information gets written into the PDF file information:
-\ifpdf
-  \pdfinfo{
-    /Title        (An Empirical Analysis of the Internet Engineering Task Force with Computational Methods)
-    /Author       (...)
-    /Subject      (Internet Standards)
-    /Keywords     (IETF, Internet, Standardisation)
-    /CreationDate (D:20240922164700Z)
-    /ModDate      (D:20240922164700Z)
-    /Creator      (LaTeX)
-    /Producer     (pdfTeX)
-  }
-  % Suppress unnecessary metadata, to ensure the PDF generated by pdflatex is
-  % identical each time it is built. This needs pdfTeX 3.14159265-2.6-1.40.17
-  % or later.
-  \ifdefined\pdftrailerid
-    \pdftrailerid{}
-    \pdfsuppressptexinfo=15
-  \fi
-\fi
 %==================================================================================================

### 2024-10-07 paper.tex
Update Introduction

@@
 
-This gives a unique opportunity to explore the development of the Internet
-through its standards. Standards development is an inherently social and
-political process \cite{RFC2026,RFC7282}, requiring cooperation and
+The availability of this data gives a unique opportunity to explore the
+development of the Internet. Standards development is an inherently social
+and political process \cite{RFC2026,RFC7282}, requiring cooperation and
 consensus among a growing number of stakeholders. Due to the importance
@@
 develops Internet standards, how the participants interact, and what
-factors may lead to success.
-Section \ref{sec:background} introduces the IETF and the Internet Standards
-development process, and reviews the data with which we work.
-Section \ref{sec:trends-documents} considers trends in document production,
-complexity, and correctness.
-Section \ref{sec:trends-demographics} moves on to consider the people
-who are involved in Internet standards development and how the demographics
-of the community have shifted over time.
-Section \ref{sec:org-dyn} studies the interaction between participants,
-explores who has influence and impact in the IETF and how that influence is
-evident in their activities, and the dynamics of communications within the
-IETF.
-Section \ref{sec:language} considers communications within the IETF from
-the perspective of language to explore whether there are differences in
-linguistic patterns and language use in more influential participants, and
-to consider how the use of language reflects the consensus-driven nature of
-the standards development process.
-Section \ref{sec:success-factors} then considers factors that lead to
-success, exploring what makes a document likely to be published as an RFC
-and what makes a successful participant.
-Finally, we review related work in Section \ref{sec:related} and conclude
-in Section \ref{sec:conclusions}.
-
-This paper is a summary and collation of our prior published work,
-that can be found in full in:
-\cite{mcquistin:2021:characterising,khare:2022:web-we-weave,
-healey:2023:power,mcquistin:2023:errare,khare:2023:tracing,
-karan:2023:leda,healey:2023:power-frontiers,barnes:2024:temporal}.
+factors may lead to success:
+\begin{itemize}
+  \item Section \ref{sec:background} introduces the IETF and the Internet
+    Standards development process, and reviews the available data.
+
+  \item Section \ref{sec:trends-documents} considers trends in the
+    production, complexity, and correctness of IETF standards documents
+    over time.
+
+  \item Section \ref{sec:trends-demographics} moves on to consider the
+    people who are involved in Internet standards development and how the
+    demographics of the community have shifted over time.
+
+  \item Section \ref{sec:org-dyn} studies the interaction between
+    participants, explores who has influence and impact in the IETF and how
+    that influence is evident in their activities, and the dynamics of
+    communications within the IETF.
+
+  \item Section \ref{sec:language} considers communications within the IETF
+    from the perspective of language to explore whether there are
+    differences in linguistic patterns and language use in more influential
+    participants, and to consider how the use of language reflects the
+    consensus-driven nature of the standards development process.
+
+  \item Section \ref{sec:success-factors} discusses what factors lead to
+    success, exploring what makes a document likely to be published as an
+    RFC and what makes a successful author.
+
+\end{itemize}
+
+Finally, we review related work in Section \ref{sec:related} and draw some
+conclusions in Section \ref{sec:conclusions}. This paper is a summary and
+collation of our prior published work \cite{mcquistin:2021:characterising,
+khare:2022:web-we-weave, healey:2023:power, mcquistin:2023:errare,
+khare:2023:tracing, karan:2023:leda, healey:2023:power-frontiers,
+barnes:2024:temporal} that is submitted to stimulate discussion.
 

### 2024-10-07 paper.tex
Tweaks

@@
 has not been previously seen, a new person identifier is generated. This
-accounts for a small portion (~10\%) of messages. This is reasonable, given
-that Datatracker profiles have become necessary for many of the IETF's
-day-to-day activities.
+accounts for a small portion (10\%) of messages. This is reasonable, since
+Datatracker profiles are necessary for many of the IETF's day-to-day
+activities.
 
@@
 performs necessary post-processing. This library is available as open
-source and Ongoing development is coordinated on GitHub.%
+source and ongoing development is coordinated on GitHub.%
 \footnote{\url{https://github.com/glasgow-ipl/ietfdata}}
@@
 discussed our work with IETF leadership (IETF and IAB Chairs, the IETF
-Executive Director, and the IRTF Chair, who is a co-author on this paper)
+Executive Director)
 and the RFC Editor to ensure that our access falls within their acceptable
@@
   \caption{
-    Number of active drafts over time. Internet-Drafts, working documents
-    of the IETF for work that was intended for eventual publication as an
-    RFC, were introduction at the 13th IETF meeting in 1989.
+    Number of active drafts over time. Internet-Drafts, the working draft
+    series of the IETF, were introduced at the 13th IETF meeting in 1989.
   }
@@
 proceed, and hence not all drafts become RFCs. As we discuss in
-\S\ref{sec:success-factors:documents}, draft adoption is a significant
+\S\ref{sec:success-factors:documents}, adoption is a significant
 milestone.
@@
 Figure~\ref{fig:keyword_usage_rates} further confirms the growing complexity
-of RFCs, showing how the use of normative language keywords has evolved
-over time.  Keywords are used in RFCs to indicate the normative requirements
-an RFC imposes on implementations. Figure~\ref{fig:keyword_usage_rates}
-shows the total number of occurrences of each of the RFC 2119~\cite{RFC2119}
+of RFCs, showing how the use of normative language has evolved.
+Keywords are used in RFCs to indicate the normative requirements
+an RFC imposes on implementations \cite{RFC2119}. Figure~\ref{fig:keyword_usage_rates}
+shows the total number of occurrences of each of the 
 keywords (i.e., MUST, MUST NOT, REQUIRED, SHALL, SHALL NOT, SHOULD, SHOULD
@@
 shown, the median number of keywords per page grew from 2001 through to
-2010, indicating a growing number of requirements being expressed in RFCs,
+2010, indicating a growing number of requirements being,
 before plateauing in recent years.
@@
 in the IETF. Care is needed to ensure that this relevance is maintained,
-however: we observe that the author pool has grown less diverse in terms of
+however: the author pool has grown less diverse in terms of
 companies that are represented. 35.4\% of authors came from the top 10
@@
 standardisation, but also its resilience to the loss of major participants.
-In this section we consider the following research questions: 
+In this section we consider the following:
 \begin{enumerate}
@@
 IETF Datatracker we compile a dataset for the \emph{adoption} of a draft by
-a working group, comprising 11632 drafts, and employ logistic regression to
+a working group, comprising 11,632 drafts, and employ logistic regression to
 analyse factors that affect working group adoption of draft.
@@
   \item We remove collinearity by using the Variance Inflation Criterion
-    (VIF), removing all features with a VIF value above 5.
+    (VIF), removing all features with VIF $>5$.
 
@@
 most of the results are in line with expectations, but more annotated data
-is required for further insights.
+is needed for further insights.
 
@@
 \S\ref{sec:org-dyn:participants}. That said, there have been several
-prior efforts that have focused on specific aspects of the IETF:
-Jari Arkko, a former Chair of the IETF, maintains a website
+prior efforts that have focused on specific aspects of the IETF.
+Arkko, a former Chair of the IETF, maintains a website
 (\url{https://www.arkko.com/tools/stats.html}) that provides various
@@
 builds on a rich literature regarding influence in online social networks
-and other communities.
-Studies of Twitter~\cite{cha2010measuring,weitzel2012measuring,anger2011measuring},
+and other communities including
+studies of Twitter~\cite{cha2010measuring,weitzel2012measuring,anger2011measuring},
 Instagram~\cite{zarei2020characterising} and decentralised social networks
@@
 organisational hierarchy in \S\ref{sec:org-dyn:hierarchy}, is an emerging
-field of research~\cite{masuda2016guide}:
-
-Some have analysed the time evolution of important nodes in a
+field of research~\cite{masuda2016guide}.
+Prior work has analysed the time evolution of important nodes in a
 network~\cite{fire2020rise}, whether their neighbours are similarly
@@
 the trajectory of snapshots of networks over
-time~\cite{lacasa2022correlations}. In this paper we bring new analysis
-techniques, that expand on the idea of tracking the evolution of node
-importance, to an email list corpus.
+time~\cite{lacasa2022correlations}. Compared to this, we brought new
+analysis techniques, that expand on the idea of tracking the evolution of
+node importance, to an email list corpus.
 
@@
 ``benevolent"~\cite{li2021does,guo2020inclusive} then power distance is
-less of a problem. We aim to test these ideas in regard to the voluntary
+less of a problem. We tested these ideas in regard to the voluntary
 and collaborative nature of the IETF organisation, whereas the literature
@@
 al.~\cite{nikkhah2017statistical} by statistically exploring RFC adoption.
-We build on this work with improved modelling, and by incorporating
+We extended this work with improved modelling, and by incorporating
 additional features that characterise not only the RFCs themselves but also
@@
 
-We explored the key standardisation activities within the IETF, considering
-trends in document production, complexity, and correctness
+We explored the key standardisation activities within the IETF, one of the
+primary technical standards developing organisations for the Internet,
+considering trends in document production, complexity, and correctness
 (\S\ref{sec:trends-documents}); the people involved in Internet standards
@@
     including Africa and South America. If the IETF is to become more
-    representative, then further effort is required to build more diverse
+    representative, then further effort is clearly required to build more diverse
     engagement, both in terms of geography, organisational affiliation, and

### 2024-10-08 scripts/fetch-ietf-dt.py
Fetch schema

@@ -23,6 +23,7 @@
 # OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
 # OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
 
+import datetime
 import json
 import os
 import requests
@@ -31,7 +32,16 @@ import sys
 from typing import Any, Dict, Iterator
 
 
-def fetch_multi(session, dt_url, uri) -> Iterator[Dict[Any, Any]]:
+def fetch_schema(session, dt_url, uri) -> Dict[str, Any]:
+    r = session.get(f"{dt_url}{uri}/schema/")
+    if r.status_code == 200:
+        return r.json()
+    else:
+        print(f"Cannot fetch schema: {r.status_code} {uri}")
+        sys.exit(1)
+
+
+def fetch_multi(session, dt_url, uri) -> Iterator[Dict[str, Any]]:
     while uri is not None:
         r = session.get(f"{dt_url}{uri}")
         if r.status_code == 200:
@@ -41,7 +51,7 @@ def fetch_multi(session, dt_url, uri) -> Iterator[Dict[Any, Any]]:
                 yield obj
             uri = meta["next"]
         else:
-            print(f"Cannot fetch: {r.status_code}")
+            print(f"Cannot fetch multi: {r.status_code} {uri}")
             sys.exit(1)
 
 
@@ -61,12 +71,16 @@ else:
 
 dt_url    = os.environ.get("IETFDATA_DT_URL", "https://datatracker.ietf.org/")
 session   = requests.Session()
+results   = {"fetched": datetime.datetime.now().isoformat(),
+             "schema": None, 
+             "data": []
+            }
 
-results   = []
+results["schema"] = fetch_schema(session, dt_url, prefix)
 
 for item in fetch_multi(session, dt_url, query_uri):
     print(f"   {item['resource_uri']}")
-    results.append(item)
+    results["data"].append(item)
 
 print(f"   Write {out_file}")
 with open(out_file, "w") as outf:

### 2024-10-09 scripts/fetch-ietf-dt.py
Add prefix to download

@@ -71,7 +71,8 @@ else:
 
 dt_url    = os.environ.get("IETFDATA_DT_URL", "https://datatracker.ietf.org/")
 session   = requests.Session()
-results   = {"fetched": datetime.datetime.now().isoformat(),
+results   = {"prefix": prefix,
+             "fetched": datetime.datetime.now().isoformat(),
              "schema": None, 
              "data": []
             }
